---
title: "Validation against the original Bentley influ implementation"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Validation against the original Bentley influ implementation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

## Purpose

The original `influ` code developed by Nokome Bentley used the `proto` object
system and GLMs [@Bentley2012]. That implementation is retained in `influ2` as
a frozen validation artefact, not as part of the package API. This vignette
runs the old code explicitly and compares it with the model-neutral S3 engine.

The compact fixture is intentionally separate from the realistic lobster
examples in the [main vignette](influ2.html). Its role is regression testing:
known inputs should reproduce known values and visual behaviour.

## Fit the common model

```{r common-model}
library(influ2)

fixture_file <- system.file(
  "extdata", "bentley-poisson-data.csv", package = "influ2"
)
bentley_data <- read.csv(
  fixture_file,
  colClasses = c(year = "factor", area = "factor", vessel = "factor")
)

bentley_glm <- glm(
  catch ~ year + area + vessel,
  family = poisson(link = "log"),
  data = bentley_data
)
new_diagnostic <- influ(bentley_glm, focus = "year")
```

The legacy source remains outside the namespace under `inst/legacy`. It is
loaded into an isolated environment only for this comparison.

```{r legacy-model}
legacy_available <- requireNamespace("proto", quietly = TRUE)
if (legacy_available) {
  legacy_environment <- new.env(parent = globalenv())
  legacy_environment$proto <- proto::proto
  sys.source(
    system.file("legacy", "influ-proto.R", package = "influ2"),
    envir = legacy_environment
  )
  legacy_diagnostic <- legacy_environment$Influence$new(
    model = bentley_glm,
    data = bentley_data,
    response = "catch",
    focus = "year"
  )
  legacy_diagnostic$calc()
  # The original code predates R 4.0 and assumes that data.frame() converts
  # text to factors. Restore that historical plotting assumption here without
  # modifying the frozen implementation.
  legacy_diagnostic$influences$level <- factor(
    legacy_diagnostic$influences$level,
    levels = legacy_diagnostic$influences$level
  )
}
legacy_available
```

## Influence plots

The left-hand figure is produced by the original `proto` method. The
right-hand figure is produced from the new `influ_diag`. Both show area and
vessel influence as ratios centred on one.

```{r bentley-influence-plots, eval=legacy_available, fig.show='hold', out.width='49%', fig.width=6.5, fig.height=5.5, fig.cap='Influence plots from the original Bentley proto implementation (left) and the new influ2 engine (right).', fig.alt=c('Legacy Bentley influence plot for area and vessel by year.', 'New influ2 influence plot for area and vessel by year.')}
legacy_diagnostic$influPlot()

plot(
  new_diagnostic,
  type = "influence",
  term = c("area", "vessel"),
  scale = "ratio"
) +
  ggplot2::facet_grid(term ~ scale, scales = "free_y") +
  ggplot2::labs(title = "New influ2")
```

## CDI plots

The complete coefficient-distribution-influence display combines the fitted
effect, the distribution of observations, and the resulting influence. Both
figures below use area as the model term and year as the focus.

```{r bentley-cdi-plots, eval=legacy_available, fig.show='hold', out.width='49%', fig.width=8, fig.height=5.5, fig.cap='CDI plots from the original Bentley proto implementation (left) and the new influ2 engine (right).', fig.alt=c('Legacy Bentley coefficient-distribution-influence plot for area.', 'New influ2 coefficient-distribution-influence plot for area.')}
legacy_diagnostic$cdiPlot(term = "area")
plot(new_diagnostic, type = "cdi", term = "area")
```

## Numerical agreement

Routine `testthat` tests compare the new result against values frozen from the
legacy implementation. The reference is not regenerated during tests, so a
change in the new implementation cannot silently move both sides of the
comparison.

```{r bentley-parity}
reference <- read.csv(system.file(
  "extdata", "bentley-poisson-reference.csv", package = "influ2"
))
current <- subset(
  influ_effects(new_diagnostic),
  term %in% c("area", "vessel") & scale == "link",
  select = c(focus, level, term, estimate)
)
names(current)[names(current) == "estimate"] <- "link_influence"

comparison <- merge(
  reference,
  current,
  by = c("focus", "level", "term"),
  suffixes = c("_legacy", "_new")
)
comparison$absolute_difference <- abs(
  comparison$link_influence_legacy - comparison$link_influence_new
)
comparison
stopifnot(max(comparison$absolute_difference) < 1e-8)
```

The final overlay makes exact agreement visually apparent: legacy results are
points, and new results are lines.

```{r bentley-parity-plot, fig.width=7, fig.height=5, fig.cap='Direct overlay of legacy Bentley influence values and new influ2 results.', fig.alt='Legacy influence points overlaid by new influ2 lines for area and vessel.'}
ggplot2::ggplot(comparison, ggplot2::aes(x = level, group = term)) +
  ggplot2::geom_hline(yintercept = 1, linetype = 3, colour = "grey55") +
  ggplot2::geom_point(
    ggplot2::aes(y = natural_influence, shape = "Legacy proto"),
    size = 3
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = exp(link_influence_new), colour = "New influ2"),
    linewidth = 0.9
  ) +
  ggplot2::facet_wrap(~term, ncol = 1, scales = "free_y") +
  ggplot2::scale_colour_manual(values = c("New influ2" = "#0072B2")) +
  ggplot2::scale_shape_manual(values = c("Legacy proto" = 21)) +
  ggplot2::labs(
    x = "Year",
    y = "Influence ratio",
    colour = NULL,
    shape = NULL
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

The frozen values can be deliberately regenerated with
`tools/legacy/regenerate-bentley-fixture.R`. That script is never run as part
of the ordinary test suite.

## References
