---
title: "Bentley validation"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Bentley validation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(ggplot2))
theme_set(theme_bw())
```

## Purpose

The original `influ` code developed by Nokome Bentley used the `proto` object
system and GLMs [@Bentley2012]. That implementation is retained in `influ2` as
a frozen validation artefact, not as part of the package API. This vignette
runs the old code explicitly and compares it with the model-neutral S3 engine.

The comparison uses the same realistic `lobsters_per_pot` example as the
[main vignette](influ2.html). A smaller frozen fixture remains in the test
suite for fast regression testing.

## Fit the common model

```{r common-model}
library(influ2)
data(lobsters_per_pot)

bentley_glm <- glm(
  lobsters ~ year + month + poly(depth, 3) + poly(soak, 3),
  family = poisson(link = "log"),
  data = lobsters_per_pot
)
new_diagnostic <- influ(bentley_glm, focus = "year")
```

The legacy source remains outside the namespace under `inst/legacy`. It is
loaded into an isolated environment only for this comparison.

```{r legacy-model}
legacy_available <- requireNamespace("proto", quietly = TRUE)
if (legacy_available) {
  legacy_environment <- new.env(parent = globalenv())
  legacy_environment$proto <- proto::proto
  sys.source(
    system.file("legacy", "influ-proto.R", package = "influ2"),
    envir = legacy_environment
  )
  legacy_diagnostic <- legacy_environment$Influence$new(
    model = bentley_glm,
    data = lobsters_per_pot,
    response = "lobsters",
    focus = "year"
  )
  legacy_diagnostic$calc()
  # The original code predates R 4.0 and assumes that data.frame() converts
  # text to factors. Restore that historical plotting assumption here without
  # modifying the frozen implementation.
  legacy_diagnostic$influences$level <- factor(
    legacy_diagnostic$influences$level,
    levels = legacy_diagnostic$influences$level
  )
}
legacy_available
```

## Influence plots

The left-hand figure is produced by the original `proto` method. The
right-hand figure is produced from the new `influ_diag`. Both show monthly,
depth, and soak-time influence as ratios centred on one across 18 years.

```{r bentley-influence-plots, eval=legacy_available, fig.show='hold', out.width='49%', fig.width=7, fig.height=7, fig.cap='Influence plots for the lobster example from the original Bentley proto implementation (left) and the new influ2 engine (right).', fig.alt=c('Legacy Bentley influence plot for month, depth, and soak time by year.', 'New influ2 influence plot for month, depth, and soak time by year.')}
legacy_diagnostic$influPlot()

plot(
  new_diagnostic,
  type = "influence",
  term = c("month", "poly(depth, 3)", "poly(soak, 3)"),
  scale = "ratio"
) +
  facet_grid(term ~ scale) +
  labs(title = "New influ2")
```

## CDI plots

The complete coefficient-distribution-influence display combines the fitted
effect, the distribution of observations, and the resulting influence. Both
figures below use month as the model term and year as the focus.

Both top-left panels now show the same relative monthly effects: log-scale
term contributions are centred on the observed monthly distribution, then
exponentiated. The reference is one on a logarithmic axis. This removes the
arbitrary choice of month 1 as the zero coefficient; a relative effect of 1.2
means a 20% higher monthly contribution to expected catch than the reference.
The legacy label "Coefficient" and the new label "Relative month effect"
describe the same fitted quantity in this example.

The interval widths deliberately differ. The original plot uses plus or minus
one standard error on the centred log scale, followed by exponentiation. The
new plot uses approximate 95% confidence intervals, including uncertainty in
the estimated centre and its covariance with each month's effect. Its annual
influence panel also includes uncertainty intervals.

```{r bentley-cdi-plots, eval=legacy_available, fig.show='hold', out.width='49%', fig.width=8, fig.height=5.5, fig.cap='CDI plots for month from Bentley proto (left) and influ2 (right). The centred relative-effect points agree; legacy top-panel bars use plus or minus one standard error, while influ2 uses 95% confidence intervals.', fig.alt=c('Legacy Bentley CDI plot with centred monthly multipliers and one-standard-error bars.', 'New influ2 CDI plot with matching centred monthly multipliers and 95% confidence intervals.')}
legacy_diagnostic$cdiPlot(term = "month")
plot(new_diagnostic, type = "cdi", term = "month")
```

To inspect the earlier reference-coded display instead, use
`coefficient_reference = "model"`. In this GLM that puts month 1 at zero and
shows the other months relative to it, on the log scale. Alternatively,
`coefficient_scale = "link"` keeps the new centring but displays additive log
effects. Neither option changes the annual influence calculations.

## Numerical agreement

First, compare the top-panel centred log effects and their standard errors.
Exponentiating the estimates gives the matching relative-effect points shown
above. Checking the standard errors separately verifies that the difference
between the plotted intervals comes from their stated coverage.

```{r bentley-coefficient-parity, eval=legacy_available}
legacy_coefficients <- aggregate(
  legacy_diagnostic$preds[c("fit.month", "se.fit.month")],
  list(level = legacy_diagnostic$preds$month),
  mean
)
names(legacy_coefficients) <- c(
  "level", "centred_estimate_legacy", "centred_std_error_legacy"
)
current_coefficients <- subset(
  new_diagnostic$coefficients,
  term == "month",
  select = c(level, centred_estimate, centred_std_error)
)
coefficient_comparison <- merge(
  legacy_coefficients, current_coefficients, by = "level"
)
coefficient_comparison$relative_effect <- exp(
  coefficient_comparison$centred_estimate
)
coefficient_comparison
stopifnot(
  max(abs(coefficient_comparison$centred_estimate_legacy -
    coefficient_comparison$centred_estimate)) < 1e-8,
  max(abs(coefficient_comparison$centred_std_error_legacy -
    coefficient_comparison$centred_std_error)) < 1e-8
)
```

The old and new engines are also compared directly on all 54 year-by-term
contrasts in the lobster example. Routine `testthat` tests retain both this
realistic comparison and a smaller frozen numerical fixture.

```{r bentley-parity, eval=legacy_available}
legacy <- reshape(
  as.data.frame(legacy_diagnostic$influences),
  direction = "long",
  varying = names(legacy_diagnostic$influences)[-1],
  v.names = "link_influence_legacy",
  timevar = "term",
  times = names(legacy_diagnostic$influences)[-1]
)
rownames(legacy) <- NULL

current <- subset(
  influ_effects(new_diagnostic),
  term != "year" & scale == "link",
  select = c(level, term, estimate)
)
names(current)[names(current) == "estimate"] <- "link_influence_new"

comparison <- merge(
  legacy[c("level", "term", "link_influence_legacy")],
  current,
  by = c("level", "term")
)
comparison$natural_influence_legacy <- exp(comparison$link_influence_legacy)
comparison$absolute_difference <- abs(
  comparison$link_influence_legacy - comparison$link_influence_new
)
head(comparison)
max(comparison$absolute_difference)
stopifnot(max(comparison$absolute_difference) < 1e-8)
```

The same validation covers the two scalar metrics from the original
implementation. `overall` summarises mean absolute influence, while `trend`
summarises the ordered change through the focus levels.

```{r bentley-metrics, eval=legacy_available}
legacy_metrics <- subset(
  legacy_diagnostic$summary,
  term != "intercept" & term != "year",
  select = c(term, overall, trend)
)
new_metrics <- reshape(
  influ_metrics(new_diagnostic)[c("term", "metric", "estimate")],
  direction = "wide",
  idvar = "term",
  timevar = "metric"
)
names(new_metrics) <- sub("^estimate\\.", "", names(new_metrics))
metric_comparison <- merge(
  legacy_metrics,
  new_metrics,
  by = "term",
  suffixes = c("_legacy", "_new")
)
metric_comparison
stopifnot(
  max(abs(metric_comparison$overall_legacy - metric_comparison$overall_new)) < 1e-8,
  max(abs(metric_comparison$trend_legacy - metric_comparison$trend_new)) < 1e-8
)
```

The final overlay makes exact agreement visually apparent: legacy results are
points, and new results are lines.

```{r bentley-parity-plot, eval=legacy_available, fig.width=7, fig.height=7, fig.cap='Direct overlay of legacy Bentley influence values and new influ2 results for the lobster example.', fig.alt='Legacy influence points overlaid by new influ2 lines for month, depth, and soak time.'}
ggplot(comparison, aes(x = level, group = term)) +
  geom_hline(yintercept = 1, linetype = 3, colour = "grey55") +
  geom_point(
    aes(y = natural_influence_legacy, shape = "Legacy proto"),
    size = 3
  ) +
  geom_line(
    aes(y = exp(link_influence_new), colour = "New influ2"),
    linewidth = 0.9
  ) +
  facet_wrap(~term, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("New influ2" = "#0072B2")) +
  scale_shape_manual(values = c("Legacy proto" = 21)) +
  labs(
    x = "Year",
    y = "Influence ratio",
    colour = NULL,
    shape = NULL
  ) +
  theme(legend.position = "bottom")
```

## References
