---
title: "Bentley validation"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Bentley validation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(ggplot2))
```

## Purpose

The original `influ` code developed by Nokome Bentley used the `proto` object
system and GLMs [@Bentley2012]. That implementation is retained in `influ2` as
a frozen validation artefact, not as part of the package API. This vignette
runs the old code explicitly and compares it with the model-neutral S3 engine.

The comparison uses the same realistic `lobsters_per_pot` example as the
[main vignette](influ2.html). A smaller frozen fixture remains in the test
suite for fast regression testing.

## Fit the common model

```{r common-model}
library(influ2)
data(lobsters_per_pot)

bentley_glm <- glm(
  lobsters ~ year + month + poly(depth, 3) + poly(soak, 3),
  family = poisson(link = "log"),
  data = lobsters_per_pot
)
new_diagnostic <- influ(bentley_glm, focus = "year")
```

The legacy source remains outside the namespace under `inst/legacy`. It is
loaded into an isolated environment only for this comparison.

```{r legacy-model}
legacy_available <- requireNamespace("proto", quietly = TRUE)
if (legacy_available) {
  legacy_environment <- new.env(parent = globalenv())
  legacy_environment$proto <- proto::proto
  sys.source(
    system.file("legacy", "influ-proto.R", package = "influ2"),
    envir = legacy_environment
  )
  legacy_diagnostic <- legacy_environment$Influence$new(
    model = bentley_glm,
    data = lobsters_per_pot,
    response = "lobsters",
    focus = "year"
  )
  legacy_diagnostic$calc()
  # The original code predates R 4.0 and assumes that data.frame() converts
  # text to factors. Restore that historical plotting assumption here without
  # modifying the frozen implementation.
  legacy_diagnostic$influences$level <- factor(
    legacy_diagnostic$influences$level,
    levels = legacy_diagnostic$influences$level
  )
}
legacy_available
```

## Influence plots

The left-hand figure is produced by the original `proto` method. The
right-hand figure is produced from the new `influ_diag`. Both show monthly,
depth, and soak-time influence as ratios centred on one across 18 years.

```{r bentley-influence-plots, eval=legacy_available, fig.show='hold', out.width='49%', fig.width=7, fig.height=7, fig.cap='Influence plots for the lobster example from the original Bentley proto implementation (left) and the new influ2 engine (right).', fig.alt=c('Legacy Bentley influence plot for month, depth, and soak time by year.', 'New influ2 influence plot for month, depth, and soak time by year.')}
legacy_diagnostic$influPlot()

plot(
  new_diagnostic,
  type = "influence",
  term = c("month", "poly(depth, 3)", "poly(soak, 3)"),
  scale = "ratio"
) +
  facet_grid(term ~ scale, scales = "free_y") +
  labs(title = "New influ2")
```

## CDI plots

The complete coefficient-distribution-influence display combines the fitted
effect, the distribution of observations, and the resulting influence. Both
figures below use month as the model term and year as the focus.

```{r bentley-cdi-plots, eval=legacy_available, fig.show='hold', out.width='49%', fig.width=8, fig.height=5.5, fig.cap='CDI plots for month in the lobster example from the original Bentley proto implementation (left) and the new influ2 engine (right).', fig.alt=c('Legacy Bentley coefficient-distribution-influence plot for month.', 'New influ2 coefficient-distribution-influence plot for month.')}
legacy_diagnostic$cdiPlot(term = "month")
plot(new_diagnostic, type = "cdi", term = "month")
```

## Numerical agreement

The old and new engines are also compared directly on all 54 year-by-term
contrasts in the lobster example. Routine `testthat` tests retain both this
realistic comparison and a smaller frozen numerical fixture.

```{r bentley-parity}
legacy <- reshape(
  as.data.frame(legacy_diagnostic$influences),
  direction = "long",
  varying = names(legacy_diagnostic$influences)[-1],
  v.names = "link_influence_legacy",
  timevar = "term",
  times = names(legacy_diagnostic$influences)[-1]
)
rownames(legacy) <- NULL

current <- subset(
  influ_effects(new_diagnostic),
  term != "year" & scale == "link",
  select = c(level, term, estimate)
)
names(current)[names(current) == "estimate"] <- "link_influence_new"

comparison <- merge(
  legacy[c("level", "term", "link_influence_legacy")],
  current,
  by = c("level", "term")
)
comparison$natural_influence_legacy <- exp(comparison$link_influence_legacy)
comparison$absolute_difference <- abs(
  comparison$link_influence_legacy - comparison$link_influence_new
)
head(comparison)
max(comparison$absolute_difference)
stopifnot(max(comparison$absolute_difference) < 1e-8)
```

The final overlay makes exact agreement visually apparent: legacy results are
points, and new results are lines.

```{r bentley-parity-plot, fig.width=7, fig.height=7, fig.cap='Direct overlay of legacy Bentley influence values and new influ2 results for the lobster example.', fig.alt='Legacy influence points overlaid by new influ2 lines for month, depth, and soak time.'}
ggplot(comparison, aes(x = level, group = term)) +
  geom_hline(yintercept = 1, linetype = 3, colour = "grey55") +
  geom_point(
    aes(y = natural_influence_legacy, shape = "Legacy proto"),
    size = 3
  ) +
  geom_line(
    aes(y = exp(link_influence_new), colour = "New influ2"),
    linewidth = 0.9
  ) +
  facet_wrap(~term, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("New influ2" = "#0072B2")) +
  scale_shape_manual(values = c("Legacy proto" = 21)) +
  labs(
    x = "Year",
    y = "Influence ratio",
    colour = NULL,
    shape = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## References
