---
title: "CPUE indices"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{CPUE indices}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
  fig.height = 5, message = FALSE, warning = FALSE)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
library(influ2)
library(ggplot2)
theme_set(theme_bw())
```

## An assessment-ready table

`cpue_index()` calculates the expected response for a fixed reference
population in every year. It produces the familiar assessment columns:
`Year`, `Mean`, `Median`, `SD`, `CV`, `Qlower`, and `Qupper`, alongside method,
distribution, and link metadata. The result is an `influ_index` object;
`as.data.frame()` gives the table for reporting or CSV export.

This is a different quantity from the centred **year-effect contrasts** in
the influence and step plots. Both are useful, but neither is automatically
an area-integrated abundance or biomass index.

We use the same simulated lobster catches as the [main article](influ2.html).
The model below includes a monthly random intercept and a negative-binomial
observation distribution. No simulation of new catches is needed for its index.

```{r index-model, eval=has_glmmTMB}
data(lobsters_per_pot)
lobster_model <- glmmTMB::glmmTMB(
  lobsters ~ year + poly(depth, 3) + poly(soak, 3) + (1 | month),
  family = glmmTMB::nbinom2(), data = lobsters_per_pot
)
```

We explicitly choose median observed depth and a 24-hour soak. The year
column is omitted: `cpue_index()` repeats this same profile in every observed
year. The monthly random effect is set to zero. That represents a zero-effect
month on the link scale, **not** the average response across the population
of months. These choices belong to the scientific definition of the index.

```{r index-calculate, eval=has_glmmTMB}
reference <- data.frame(depth = median(lobsters_per_pot$depth), soak = 24)
lobster_index <- cpue_index(
  lobster_model, year = "year", method = "standardised",
  reference_data = reference, units = "lobsters per pot"
)
knitr::kable(as.data.frame(lobster_index), digits = 3)
```

`method = "standardized"` works identically. There is no need to refit a model
when printing, exporting, or plotting this result.

```{r index-plot, eval=has_glmmTMB, fig.cap="Standardised expected lobsters per pot at the stated depth and soak-time reference profile, with the monthly random effect set to zero. The ribbon is a pointwise 95% delta-method confidence interval for the index, not the spread of individual catches."}
plot_index(lobster_index)
```

## What do the uncertainty columns mean?

For GLMs, GAMs, and glmmTMB, `Mean` is the fitted expected-response estimate;
`SD` is its propagated standard error. `CV` is `SD / Mean` when the estimate
is positive. `Median` is deliberately missing: an MLE is not a posterior
median. Positive estimates use log-scale delta-method intervals.

For a complete brms fit, `Mean`, `Median`, `SD`, and the interval columns
summarise posterior draws of the **expected-response index**. They do not
describe the mean and median of individual catches. The native
[`posterior_epred()`](https://paulbuerkner.com/brms/reference/posterior_epred.brmsfit.html)
calculation combines the distributional parameters, including zero and
positive components where appropriate, within each draw.

```{r index-brms, eval=FALSE}
# Use an existing complete fit, not the compact influence-vignette fixture.
bayesian_index <- cpue_index(
  complete_brms_fit, year = "year", reference_data = reference,
  ndraws = 1000, batch_size = 250, draw_batch_size = 100,
  retain = "summary"
)
as.data.frame(bayesian_index)
```

Prediction batches preserve the same draw identities across years and
reference rows. Only the annual summaries are retained by default. Setting
`retain = "draws"` retains a draw-by-year matrix, not the much larger
reference-row-by-draw prediction matrix. The fitted model and reference data
are not embedded in the output object. Native prediction routines can still
allocate their own working memory.

`uncertainty = "none"` omits uncertainty summaries. For frequentist models
this also avoids covariance propagation. For brms it still averages existing
posterior expected-response draws, so it is not a shortcut that substitutes
posterior-mean coefficients and changes the quantity being estimated.

## A common reference population and relative scaling

Supply multiple reference rows and explicit weights to standardise over a
population. Predictions are averaged on the response scale, after applying
the inverse link, using the same weights in each year. The function never
chooses missing factor values or an exposure variable on the user's behalf.

```{r index-weighted-reference, eval=has_glmmTMB}
population <- data.frame(depth = c(25, 45, 65), soak = 24)
relative_index <- cpue_index(
  lobster_model, year = "year", reference_data = population,
  reference_weights = c(0.2, 0.5, 0.3), rescale = 1
)
geo_mean(relative_index$table$Mean)
```

The weights here are illustrative population proportions, **not cell areas**.
For models of catch with an exposure offset, put the offset in the formula
and explicitly supply the desired exposure in the reference data. For a
per-pot index from a catch model, this would usually be one pot. This example
already models catches from individual pots.

`rescale = 1` normalises across all returned years. Frequentist uncertainty
includes covariance with that denominator. For brms, each complete annual
draw is normalised before summarising; the geometric mean of the resulting
posterior-mean column need not be exactly one. Do not separately normalise
the mean, median, and interval columns: that changes their relative scales.

`geo_mean()` is also available independently. It calculates on the log scale
to avoid numerical overflow, includes zeros, and supports `na.rm = TRUE`.
Negative values and infinities are rejected.

## Comparing models

`plot_compare()` accepts calculated `influ_index` objects. Use the same
response definition, reference population, weights, exposure, and scale.
Automatic metadata checks cannot establish scientific comparability.

```{r index-model-comparison, eval=has_glmmTMB, fig.cap="Expected-response indices from two negative-binomial glmmTMB candidates at the same reference profile. Both use the same monthly random-effect convention and response units. Intervals describe each index separately, not uncertainty in the difference between the models."}
linear_model <- glmmTMB::glmmTMB(
  lobsters ~ year + depth + soak + (1 | month),
  family = glmmTMB::nbinom2(), data = lobsters_per_pot
)
linear_index <- cpue_index(linear_model, reference_data = reference,
  units = "lobsters per pot")
plot_compare(list(Curved = lobster_index, Linear = linear_index))
```

When supplied fitted models or `influ_diag` objects instead, `plot_compare()`
keeps its existing year-effect-contrast behaviour. It does not silently switch
to response predictions. `plot_step()` also remains a year-effect comparison.

## Current boundaries

Response standardisation currently supports GLM, GAM, glmmTMB, and complete,
univariate brms fits. glmmTMB indices require a converged ML fit and do not
yet support random effects in the zero component. brms autocorrelation and
Gaussian-process predictions need a separate joint-prediction adapter. GAM
smooths are evaluated as fitted, including random-effect smooths; their target
can differ from the zero-group-effect convention used for glmmTMB and brms.

For binomial GLMs and glmmTMB, native response predictions are probabilities.
For brms binomial models they are expected counts at the reference number of
trials: supply one trial for a per-trial comparison. Quasi families and
transformed responses are not supported here. Use an explicit lognormal
family instead of asking the package to guess a back-transformation.

For sdmTMB and tinyVAST, `cpue_index(method = "year_effect")` can already
produce a table from the existing diagnostic contrasts. Full spatial
response standardisation, integration over latent-effect distributions,
and `integrate_index()` remain separate development work. They must preserve
spatial and spatiotemporal dependence and make areas, units, and catchability
assumptions explicit; they are not silently substituted by this interface.
