---
title: "CPUE indices"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{CPUE indices}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
  fig.height = 5, message = FALSE, warning = FALSE)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
has_mgcv <- requireNamespace("mgcv", quietly = TRUE)
library(influ2)
library(ggplot2)
theme_set(theme_bw())
```

## An assessment-ready table

`cpue_index()` calculates the expected response for a fixed reference
population in every year. It produces the familiar assessment columns:
`Year`, `Mean`, `Median`, `SD`, `CV`, `Qlower`, and `Qupper`, alongside method,
distribution, and link metadata. The result is an `influ_index` object;
`index_table()` gives a reporting table, omitting an unavailable median by
default. `as.data.frame()` retains the full, stable schema for existing code.

This is a different quantity from the centred **year-effect contrasts** in
the influence and step plots. Both are useful, but neither is automatically
an area-integrated abundance or biomass index.

Both `cpue_index()` and the separate `integrate_index()` work with GLM, GAM,
glmmTMB, complete brms, sdmTMB, and univariate tinyVAST fits. Spatial effects
are not required for either calculation. The difference is the question:
standardisation produces a **weighted mean**, whereas integration produces
an **area-weighted total** over a stated domain.

We use the same simulated lobster catches as the [main article](influ2.html).
The model below includes a monthly random intercept and a negative-binomial
observation distribution. No simulation of new catches is needed for its index.

```{r index-model, eval=has_glmmTMB}
data(lobsters_per_pot)
lobster_model <- glmmTMB::glmmTMB(
  lobsters ~ year + poly(depth, 3) + poly(soak, 3) + (1 | month),
  family = glmmTMB::nbinom2(), data = lobsters_per_pot
)
```

We explicitly choose median observed depth and a 24-hour soak. The year
column is omitted: `cpue_index()` repeats this same profile in every observed
year. The monthly random effect is set to zero. That represents a zero-effect
month on the link scale, **not** the average response across the population
of months. These choices belong to the scientific definition of the index.

```{r index-calculate, eval=has_glmmTMB}
reference <- data.frame(depth = median(lobsters_per_pot$depth), soak = 24)
lobster_index <- cpue_index(
  lobster_model, year = "year", method = "standardised",
  reference_data = reference, units = "lobsters per pot"
)
knitr::kable(index_table(lobster_index), digits = 3)
```

`method = "standardized"` works identically. There is no need to refit a model
when printing, exporting, or plotting this result.

```{r index-plot, eval=has_glmmTMB, fig.cap="Standardised expected lobsters per pot at the stated depth and soak-time reference profile, with the monthly random effect set to zero. The ribbon is a pointwise 95% delta-method confidence interval for the index, not the spread of individual catches."}
plot_index(lobster_index)
```

## What do the uncertainty columns mean?

For GLMs, GAMs, and glmmTMB, `Mean` is the fitted expected-response estimate;
`SD` is its propagated standard error. `CV` is `SD / Mean` when the estimate
is positive. `Median` is deliberately missing: an MLE is not a posterior
median. Positive estimates use log-scale delta-method intervals.

For a complete brms fit, `Mean`, `Median`, `SD`, and the interval columns
summarise posterior draws of the **expected-response index**. They do not
describe the mean and median of individual catches. The native
[`posterior_epred()`](https://paulbuerkner.com/brms/reference/posterior_epred.brmsfit.html)
calculation combines the distributional parameters, including zero and
positive components where appropriate, within each draw.

```{r index-brms, eval=FALSE}
# Use an existing complete fit, not the compact influence-vignette fixture.
bayesian_index <- cpue_index(
  complete_brms_fit, year = "year", reference_data = reference,
  ndraws = 1000, batch_size = 250, draw_batch_size = 100,
  retain = "summary"
)
as.data.frame(bayesian_index)
```

Prediction batches preserve the same draw identities across years and
reference rows. Only the annual summaries and their two small covariance
matrices are retained by default. Setting
`retain = "draws"` retains a draw-by-year matrix, not the much larger
reference-row-by-draw prediction matrix. The fitted model and reference data
are not embedded in the output object. Native prediction routines can still
allocate their own working memory.

`uncertainty = "none"` omits uncertainty summaries. For frequentist models
this also avoids covariance propagation. For brms it still averages existing
posterior expected-response draws, so it is not a shortcut that substitutes
posterior-mean coefficients and changes the quantity being estimated.

## Lognormal assessment parameters

Some assessments accept an index median and a log-scale SD, whereas others
accept an arithmetic mean and CV. `index_table(index, format = "lognormal")`
provides an explicit marginal lognormal approximation from the index's `Mean`
and `SD`:

\[
  s^2 = \log(1 + (\mathrm{SD}/\mathrm{Mean})^2),\qquad
  m = \log(\mathrm{Mean}) - s^2/2.
\]

The additional columns are `Meanlog = m`, `SDlog = s`, and
`LognormalMedian = exp(m)`. Substituting them into the usual lognormal moment
formulae recovers the input mean and variance. They describe **uncertainty in
the annual index**, not the distribution or residual dispersion of the fitted
response. A negative-binomial CPUE model can therefore supply a lognormal
assessment approximation without becoming a lognormal response model.

```{r index-lognormal, eval=has_glmmTMB}
assessment_marginals <- index_table(lobster_index, format = "lognormal")
knitr::kable(assessment_marginals[c("Year", "Mean", "SD", "CV",
  "Meanlog", "SDlog", "LognormalMedian")], digits = 3)
stopifnot(isTRUE(all.equal(
  exp(assessment_marginals$Meanlog + assessment_marginals$SDlog^2 / 2),
  assessment_marginals$Mean
)))
```

An actual posterior `Median`, where available, is preserved separately; it is
not overwritten by `LognormalMedian`. Frequentist reporting omits the missing
`Median` by default, while `include_median = "always"` retains that column.
This resolves the empty-column presentation without inventing posterior
summaries. Non-positive means, invalid SDs, preview results, and year-effect
contrasts cannot produce this lognormal table.

The moment-matched `Meanlog` is not `log(Mean)`. Similarly, `SDlog` is not
generally the log-index SE from the joint calculation below. Choose the
assessment likelihood's convention explicitly: do not silently substitute
one centre or uncertainty measure for the other.

## Joint annual uncertainty

`index_vcov(index)` (or `vcov(index)`) returns covariance among **log annual
indices**; `scale = "response"` gives covariance in the reported units.
The calculation uses the same reference population, random-effect target,
and normalisation as the table. It never refits the model or simulates new
observations. The diagonal of the response matrix matches `SD^2`.

| Backend | Joint covariance calculation |
| --- | --- |
| GLM, GAM, glmmTMB | Delta method through the weighted annual response means; log-index gradients for log covariance. |
| brms | Sample covariance of annual posterior expected-response draws; take logs of those same draws for log covariance. |
| sdmTMB, tinyVAST | Sample covariance of joint Gaussian parameter/field simulations of the annual index; these are not MCMC draws. |

Shared parameters and fields can correlate estimates in different years.
This is not a residual autocorrelation estimate. A matrix cannot be recovered
from marginal SDs alone; old saved summaries without covariance must be
recalculated from their fitted models. `uncertainty = "none"` deliberately
does not provide a matrix. Log covariance requires positive estimates and,
for simulation methods, positive draws; values are not clipped to pass this
check. A response-scale matrix can still exist when log covariance cannot.

```{r index-joint-uncertainty, eval=has_glmmTMB}
Sigma <- index_vcov(lobster_index, require_pd = TRUE)
stopifnot(identical(rownames(Sigma), as.character(lobster_index$table$Year)))
stopifnot(isTRUE(all.equal(
  unname(diag(index_vcov(lobster_index, scale = "response"))),
  lobster_index$table$SD^2
)))

# Select and order both dimensions together, then match the table to them.
selected_years <- tail(lobster_index$table$Year, 5)
selected_Sigma <- index_vcov(lobster_index, years = selected_years)
selected_table <- index_table(lobster_index)[
  match(rownames(selected_Sigma), lobster_index$table$Year), ]
stopifnot(identical(selected_table$Year, rownames(selected_Sigma)))
```

```{r index-correlation, eval=has_glmmTMB, fig.cap="Correlation among the unscaled annual log CPUE indices. The off-diagonal cells preserve dependence from the shared fitted model; these are not correlations among observed responses or residuals. The full covariance, rather than this unitless display, is the assessment input."}
plot(lobster_index, type = "correlation")
```

For a covariance display, use `plot(lobster_index, type = "covariance")`.
The [main article](influ2.html#passing-annual-covariance-to-an-assessment)
shows both and an example assessment-input bundle. Keeping estimated
dependence is preferable to discarding it *when the joint uncertainty model
is appropriate*, but it is not a guarantee of improved stock-status estimates.
@Hoyle2024 discuss covariance propagation in index construction (Section 5.8)
and the additional sources of uncertainty an assessment may need (Section 5.5).

Retain `rescale = "raw"` for an assessment with an estimated catchability.
Draw-wise geometric-mean normalisation removes a common level and makes
the log covariance singular. The default extractor permits positive
semidefinite matrices for inspection; `require_pd = TRUE` rejects a matrix
that cannot safely be Cholesky-factorised. It never repairs eigenvalues or
adds diagonal jitter. Too few draws or other model constraints can also
produce rank deficiency.

Area integration retains the same matrices. Multiplying by a known positive
area or unit-conversion constant multiplies response covariance by its square
and leaves log covariance unchanged. Reference-weight and catchability
uncertainty, additional assessment process/error variance, and cross-series
covariance between independently calculated objects are not supplied.
For brms, the matrix summarises a posterior, not a prior-free likelihood;
check prior reuse before treating it as new independent assessment evidence.

## A common reference population and relative scaling

Supply multiple reference rows and explicit weights to standardise over a
population. Predictions are averaged on the response scale, after applying
the inverse link, using the same weights in each year. The function never
chooses missing factor values or an exposure variable on the user's behalf.

```{r index-weighted-reference, eval=has_glmmTMB}
population <- data.frame(depth = c(25, 45, 65), soak = 24)
relative_index <- cpue_index(
  lobster_model, year = "year", reference_data = population,
  reference_weights = c(0.2, 0.5, 0.3), rescale = 1
)
geo_mean(relative_index$table$Mean)
```

The weights here are illustrative population proportions, **not cell areas**.
For models of catch with an exposure offset, put the offset in the formula
and explicitly supply the desired exposure in the reference data. For a
per-pot index from a catch model, this would usually be one pot. This example
already models catches from individual pots.

`rescale = 1` normalises across all returned years. Frequentist uncertainty
includes covariance with that denominator. For brms, each complete annual
draw is normalised before summarising; the geometric mean of the resulting
posterior-mean column need not be exactly one. Do not separately normalise
the mean, median, and interval columns: that changes their relative scales.

`geo_mean()` is also available independently. It calculates on the log scale
to avoid numerical overflow, includes zeros, and supports `na.rm = TRUE`.
Negative values and infinities are rejected.

## Comparing models

`plot_compare()` accepts calculated `influ_index` objects. Use the same
response definition, reference population, weights, exposure, and scale.
Automatic metadata checks cannot establish scientific comparability.

```{r index-model-comparison, eval=has_glmmTMB, fig.cap="Expected-response indices from two negative-binomial glmmTMB candidates at the same reference profile. Both use the same monthly random-effect convention and response units. Intervals describe each index separately, not uncertainty in the difference between the models."}
linear_model <- glmmTMB::glmmTMB(
  lobsters ~ year + depth + soak + (1 | month),
  family = glmmTMB::nbinom2(), data = lobsters_per_pot
)
linear_index <- cpue_index(linear_model, reference_data = reference,
  units = "lobsters per pot")
plot_compare(list(Curved = lobster_index, Linear = linear_index))
```

When supplied fitted models or `influ_diag` objects instead, `plot_compare()`
keeps its existing year-effect-contrast behaviour. It does not silently switch
to response predictions. `plot_step()` also remains a year-effect comparison.

## Current boundaries

Response standardisation and area integration support all six backends above,
subject to model-structure checks. glmmTMB indices require a converged ML fit and do not
yet support random effects in the zero component. brms autocorrelation and
Gaussian-process predictions need a separate joint-prediction adapter. GAM
smooths are evaluated as fitted, including random-effect smooths; their target
can differ from the zero-group-effect convention used for glmmTMB and brms.

For binomial GLMs and glmmTMB, native response predictions are probabilities.
For brms binomial models they are expected counts at the reference number of
trials: supply one trial for a per-trial comparison. Quasi families and
transformed responses are not supported here. Use an explicit lognormal
family instead of asking the package to guess a back-transformation.

For sdmTMB and tinyVAST, response predictions include fitted spatial and
spatiotemporal fields by default. `spatial_fields` can select `"all"`,
`"spatial"`, `"spatiotemporal"`, or `"none"` without refitting. Joint Gaussian
parameter/field draws preserve dependence across cells and years. These are
not MCMC draws. The reported frequentist `Mean` is a plug-in expectation,
with simulation-based uncertainty, **not a Laplace bias-corrected total**.
Native `sdmTMB::get_index()` and `tinyVAST::integrate_output()` offer bias
correction; agreement checks use their uncorrected estimates to compare the
same calculation. Frequentist `Median` remains undefined in the full schema
and is omitted from the default reporting table.

All calculations use a fixed common reference domain in the observed years.
They do not forecast or automatically supply year-varying environmental
covariates. Multivariate tinyVAST response integration and sdmTMB nonlocal
covariate operators are rejected explicitly. Integration over a new population
of latent random effects is a different target, not implied by these methods.

## Area integration for ordinary and spatial models

No specialist spatiotemporal model is required. Given a prediction grid,
`integrate_index()` sums `cell area * expected response` in each year.
The same routine works for a GLM with no spatial effects, a GAM with a spatial
smooth, or any of the other supported model classes. A model without a spatial
term cannot reveal spatial variation that is not explained by its other
predictors, but its predictions can still be integrated.

This separate simulated example has a density response in kg/km², so multiplying
by cell areas in km² yields kg. It is not a conversion of the lobster pot catches
above into biomass. The changing survey footprint is deliberately imbalanced.

```{r area-simulation}
set.seed(710)
survey <- data.frame(year = factor(rep(2011:2016, each = 100)))
survey$x <- runif(nrow(survey), 0, 10)
survey$y <- runif(nrow(survey), 0, 10)
survey$depth <- 20 + 3 * survey$x + 2 * survey$y
eta <- 2 + 0.08 * as.integer(survey$year) - 0.012 * survey$depth +
  0.7 * sin(survey$x / 2) * cos(survey$y / 3)
survey$density <- rgamma(nrow(survey), shape = 5, scale = exp(eta) / 5)
sampling_centre <- 1 + 8 * (as.integer(survey$year) - 1) / 5
survey <- survey[runif(nrow(survey)) <
  0.35 + 0.65 * exp(-(survey$x - sampling_centre)^2 / 8), ]

# One row per 0.5 x 0.5 km cell: 400 cells over a 100 km² domain.
density_grid <- expand.grid(x = seq(0.25, 9.75, by = 0.5),
                           y = seq(0.25, 9.75, by = 0.5))
density_grid$depth <- 20 + 3 * density_grid$x + 2 * density_grid$y
density_grid$area_km2 <- 0.25
```

```{r area-gam, eval=has_mgcv}
gam_density <- mgcv::gam(density ~ year + s(x, y, k = 20),
  family = Gamma(link = "log"), data = survey, method = "REML")
glm_density <- glm(density ~ year + depth,
  family = Gamma(link = "log"), data = survey)

gam_total <- integrate_index(gam_density, density_grid, area = "area_km2",
  area_units = "km^2", response_units = "kg/km^2", units = "kg")
glm_total <- integrate_index(glm_density, density_grid, area = "area_km2",
  area_units = "km^2", response_units = "kg/km^2", units = "kg")
knitr::kable(index_table(gam_total), digits = 3)
```

```{r area-comparison, eval=has_mgcv, fig.cap="Area-integrated expected density over the same 100 km² domain, comparing a GAM with a spatial smooth against a GLM with a depth effect but no explicit spatial term. Both return totals in kg under the simulated density units. Pointwise 95% confidence intervals propagate joint coefficient uncertainty; they do not describe new catches or uncertainty in the difference between models."}
plot_compare(list(`Spatial GAM` = gam_total, `Depth GLM` = glm_total))
```

Replace the fitted model with a glmmTMB or complete brms fit to use exactly
the same integration call. The [spatial article](spatial-spatiotemporal.html#response-indices-and-area-totals)
executes both index calculations for sdmTMB and tinyVAST, including their fields.

Area integration and area-weighted standardisation are related, but not identical:
with no catchability conversion, the total is the area-weighted mean times total
area. Their units differ. `rescale = 1` gives a relative series and includes the
shared normalising denominator in uncertainty calculations.

### Exposure, catchability, and repeated seasons

Integrating kg/tow over km² gives an area-weighted CPUE index, **not kg of fish**.
Either model a density already expressed per km² or supply a defensible known
`catchability` conversion with compatible units. The package divides predictions
by that conversion; it does not estimate catchability or its uncertainty. It
also does not infer cell areas from longitude and latitude: use suitable
geodesic or equal-area calculations outside this function.

If each cell appears in several seasonal reference rows, provide
`averaging_weights` that sum to one within each cell. For example, four equally
weighted quarters need `averaging_weights = 1 / 4`. Without this adjustment,
repeating the full cell area four times computes a sum over four seasons rather
than an annual mean surface. This is also relevant to month-by-area GAM grids:
the integration weights must represent the intended seasonal and spatial target.

```{r area-seasonal, eval=FALSE}
seasonal_total <- integrate_index(seasonal_model, quarter_grid,
  area = "area_km2", averaging_weights = 1 / 4,
  area_units = "km^2", response_units = "kg/km^2", units = "kg")
```

Only the compact annual table and covariance matrices are retained by default.
Spatial and brms indices
can retain annual draws with `retain = "draws"`; grid-by-draw arrays are not
stored. Independent cell standard errors are never added together. All area,
covariate, and seasonal weights are treated as known, and valid units, domain
coverage, and extrapolation remain scientific decisions for the analyst.

## References
