---
title: "Hurdle and zero-inflated models"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Hurdle and zero-inflated models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
has_brms <- requireNamespace("brms", quietly = TRUE) &&
  requireNamespace("posterior", quietly = TRUE)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

## Why retain components?

A hurdle or delta model has an occurrence component and a positive component.
Its unconditional expected response is

\[
E(Y) = \Pr(Y > 0) E(Y \mid Y > 0).
\]

A zero-inflated count model instead combines an additional-zero probability
with a count distribution:

\[
E(Y) = \{1 - \Pr(\text{extra zero})\}\mu_{\text{count}}.
\]

An influence diagnostic should show both component-specific effects and the
combined change in expected response. Treating one component as though it were
the complete CPUE index can obscure why an index changed.

## BRMS hurdle-lognormal example

A hurdle-lognormal model can be fitted with separate formulae for positive
catch and the probability of a zero:

```{r brms-hurdle-fit, eval=FALSE}
fit <- brms::brm(
  brms::bf(y ~ year + group, hu ~ 1),
  data = simulated_cpue,
  family = brms::hurdle_lognormal(),
  chains = 2,
  iter = 2000,
  seed = 42,
  file = "m1",
  file_refit = "on_change"
)
```

The vignette uses the prefitted object supplied with its source. Joint
posterior draws are reduced directly to compact component and
unconditional-mean contrasts.

```{r brms-hurdle, eval=has_brms}
library(influ2)
fit <- readRDS("m1.rds")
hurdle_diagnostic <- influ(fit, focus = "year", ndraws = 250)
unique(hurdle_diagnostic$influence$component)
summary(hurdle_diagnostic)
```

```{r brms-hurdle-components, eval=has_brms, fig.cap="Positive, occurrence, and unconditional-mean influence from a BRMS hurdle-lognormal model."}
plot(hurdle_diagnostic, type = "components")
```

Because this example has `hu ~ 1`, the probability of occurrence has no
year-varying model term. Consequently, its unconditional year influence is the
same as its positive-component year influence. A varying hurdle formula would
allow both components to contribute.

## glmmTMB hurdle example

For `glmmTMB`, the truncated conditional family and zero formula form a hurdle
model. The package simulates the conditional and zero-formula coefficients
jointly from the fitted covariance before combining them.

```{r glmmtmb-hurdle, eval=has_glmmTMB}
set.seed(31)
n <- 480
cpue <- data.frame(
  year = factor(rep(1:4, each = n / 4)),
  depth = rnorm(n)
)
positive_mean <- exp(0.5 + 0.1 * as.numeric(cpue$year) + 0.2 * cpue$depth)
zero_probability <- plogis(-0.8 + 0.15 * as.numeric(cpue$year))
cpue$catch <- ifelse(
  rbinom(n, 1, zero_probability),
  0,
  1 + rnbinom(n, mu = positive_mean, size = 2)
)

hurdle_fit <- glmmTMB::glmmTMB(
  catch ~ year + depth,
  ziformula = ~year,
  family = glmmTMB::truncated_nbinom2(),
  data = cpue
)
glmmTMB_hurdle <- influ(
  hurdle_fit,
  focus = "year",
  ndraws = 500,
  seed = 1
)
unique(glmmTMB_hurdle$influence$component)
```

```{r glmmtmb-hurdle-components, eval=has_glmmTMB, fig.cap="Positive, occurrence, and unconditional-mean influence from a glmmTMB hurdle model."}
plot(glmmTMB_hurdle, type = "components")
```

Zero-inflated Poisson and negative-binomial models follow the same pathway,
with `zero_probability` and `count` labels replacing `occurrence` and
`positive`.

## Dependence and memory

Component combination must be performed draw by draw, or with a joint
covariance or precision calculation. Multiplying independently calculated
marginal intervals is invalid because it loses dependence between the two
linear predictors. `influ2` therefore combines:

- BRMS population-level components draw by draw from the joint posterior; and
- `glmmTMB` fixed components draw by draw from their joint maximum-likelihood
  covariance; and
- `sdmTMB` and `tinyVAST` latent fields draw by draw from their sparse joint
  precision matrices.

Only derived focus-by-term draws are retained when requested. The much larger
observation-by-draw arrays are never stored.

For lognormal models, a constant residual-scale adjustment cancels from an
influence ratio. Distributional models in which the scale itself varies need a
future scale-component adapter.

## Spatial delta models

`sdmTMB` and `tinyVAST` delta models expose occurrence and positive fixed effects
and latent fields as named components. Their fixed effects are combined into an
unconditional mean with joint covariance simulation. For example:

```{r sdmtmb-delta, eval=FALSE}
fit <- sdmTMB::sdmTMB(
  cpue ~ 0 + as.factor(year) + depth,
  data = cpue_data,
  mesh = mesh,
  time = "year",
  family = sdmTMB::delta_gamma()
)

diagnostic <- influ(fit, focus = "year")
plot(diagnostic, type = "components")
```

The combined delta mean also includes latent spatial and spatiotemporal fields.
Both components are evaluated from the same sparse joint-precision draw before
the unconditional mean is calculated, preserving posterior dependence rather
than assembling independent marginal intervals. The motivation for this
spatial extension follows @Hsu2022.

## References
