---
title: "Hurdle, delta, and zero-inflated influence"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Hurdle, delta, and zero-inflated influence}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why retain components?

A hurdle or delta model has an occurrence component and a positive component.
Its unconditional expected response is

\[
E(Y) = \Pr(Y > 0) E(Y \mid Y > 0).
\]

A zero-inflated count model instead combines an additional-zero probability
with a count distribution:

\[
E(Y) = \{1 - \Pr(\text{extra zero})\}\mu_{\text{count}}.
\]

An influence diagnostic should show both the component-specific effects and
the combined change in expected response. Treating one component as though it
were the complete CPUE index can obscure the reason that the index changed.

## Fitting and calculating

The `influ()` method reads the component formulas from the fitted model. For
example, a `brms` hurdle-lognormal model may be fitted and analysed as follows:

```{r brms-hurdle, eval=FALSE}
fit <- brms::brm(
  brms::bf(cpue ~ year + vessel + depth, hu ~ year + vessel),
  data = cpue_data,
  family = brms::hurdle_lognormal()
)

diagnostic <- influ(fit, focus = "year")
plot(diagnostic, type = "components")
```

`sdmTMB` delta models expose occurrence and positive spatial fields separately:

```{r sdmtmb-delta, eval=FALSE}
fit <- sdmTMB::sdmTMB(
  cpue ~ 0 + as.factor(year) + depth,
  data = cpue_data,
  mesh = mesh,
  time = "year",
  family = sdmTMB::delta_gamma()
)

diagnostic <- influ(fit, focus = "year")
plot(diagnostic, type = "components")
```

The component labels are stable across backends: `occurrence`, `positive`,
`count`, and `zero_probability`, with spatial or spatiotemporal qualifiers when
appropriate.

## Uncertainty

Component combination must be performed draw by draw, or with a joint
covariance/precision calculation. Multiplying independently calculated
marginal intervals is not valid because it loses dependence between the two
linear predictors. Consequently, `influ2` reports component influence now and
will add the combined unconditional-mean result only through a joint
calculation.

This is particularly important for lognormal models, where conversion from the
log-location parameter to the response mean may additionally depend on the
scale parameter. The family adapter, rather than the plotting code, owns that
parameterisation.

The spatial extension follows the motivation of @Hsu2022, while retaining
prediction-grid weighting and full-resolution field calculations as distinct
future estimands.

## References
