---
title: "Residual-implied effects"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Residual-implied effects}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, fig.width = 11, fig.height = 6, dev.args = list(bg = "white"))
library(influ2)
library(ggplot2)
theme_set(theme_bw())
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

## Restore the original question

Residual-implied plots ask whether individual groups suggest a different
annual trajectory from the fitted model. The original comparison is useful:
show a common fitted annual pattern, then show how residual departures in
each group would adjust that pattern. These are **exploratory implied effects**,
not separately fitted interactions or regional abundance indices.

`plot_implied_residuals()` now defaults to a local **likelihood adjustment**.
The original model's parameters, fitted random effects, smooths, dispersion,
and exposure offsets stay fixed. Only one additional effect-scale shift is
estimated in each supported group-year stratum. For a log-link count model,
that shift multiplies the expected response, while respecting the fitted
count distribution. No full model is refitted and no MCMC is run.

The zero-centred PIT summaries previously displayed under this name are now
available as `plot_grouped_residuals()`. That calculation has not changed.
It answers a different question: whether a group's observations tend to lie
high or low within their predictive distributions. **Normal-score PIT
residuals are not added to model coefficients.**

## Three conventions, explicitly distinguished

| Calculation | Adjustment added to the selected fitted baseline | Interpretation |
|:--|:--|:--|
| `method = "likelihood"` | Local shift maximising the fitted response likelihood | Default effect-scale diagnostic |
| `method = "traditional"` | Mean ordinary log-response residual | Classical lognormal construction; constant log variance required |
| Traditional with `traditional_scale = "standardised"` | Mean globally centred `rstandard()` residual | Actual historical analyser GLM convention; retained for comparison, not a universal effect-scale calculation |

The [historical analyser source](https://github.com/trophia/analyser/blob/a82c5b41bfd0ebeece8e1c6526465da006435497/R/diagnoser.r)
uses `rstandard()` for GLMs, centres those residuals globally, and adds selected
fitted term contributions. Its `implieds()` helper can include the group main
effect as well as year. Thus "traditional" was not one universal residual
definition. Fisheries examples include @StarrKendrick2019 and @Middleton2025.

The new calculation agrees with **ordinary log-residual arithmetic** when
the variance is constant; it does not generally equal the historical
standardised-residual arithmetic. The examples below demonstrate both claims.

The baseline defaults to `"year_group"`: centred fixed year and group main
effects, where the latter is present. Choose `baseline = "year"` for only the
shared year term. All selected term contributions are centred over the same
fitted observations. They are not independently re-centred in each panel.
Other covariate effects, random effects, and smooths stay in the observation
predictions used to calculate the adjustment, but are not added to this baseline.

## A simulated lobster design with differing group trends

We retain the uneven year/month coverage, depth, and soak time from
`lobsters_per_pot`, but simulate new responses for this comparison. These are
not modifications to the package dataset. Three seasonal groups are defined
from month, and their trends intentionally differ. Vessel effects add
repeated-observation structure. Both responses represent CPUE per standardised
sampling unit: one continuous positive response and one count response.

```{r implied-data}
data(lobsters_per_pot)
example <- lobsters_per_pot
example$season <- factor(ceiling(as.integer(example$month) / 4),
  labels = c("Early season", "Mid-season", "Late season"))
set.seed(12092026)
example$vessel <- factor(sample(1:24, nrow(example), replace = TRUE))
time <- as.integer(example$year)
trend <- time - mean(time)
vessel_effect <- rnorm(24, sd = .35)
eta <- 1.5 + .15 * sin(time / 2) - .015 * (example$depth - 40) +
  .008 * (example$soak - 30) + .15 * (as.integer(example$season) - 2) +
  .07 * trend * (as.integer(example$season) - 2) +
  vessel_effect[as.integer(example$vessel)]
example$cpue_positive <- exp(eta + rnorm(nrow(example), sd = .6))
example$count_cpue <- rnbinom(nrow(example), mu = exp(eta), size = 4)
```

### Constant log variance: matching the classical construction

Fit a Gaussian model to log CPUE, omitting the generating season-by-year
interaction. This is a lognormal response model with constant log-scale
variance, not a Gaussian model fitted directly to untransformed CPUE.

```{r implied-log-model, eval=has_glmmTMB}
log_fit <- glmmTMB::glmmTMB(
  log(cpue_positive) ~ year + season + depth + soak + (1 | vessel),
  family = gaussian(), data = example
)
new_log <- implied_effects(log_fit, groups = "season", interval = "descriptive")
traditional_log <- implied_effects(log_fit, groups = "season", method = "traditional")
max(abs(new_log$table$estimate - traditional_log$table$estimate))
stopifnot(isTRUE(all.equal(new_log$table, traditional_log$table, tolerance = 1e-10)))
```

The estimates agree because the conditional Gaussian likelihood shift is the
mean log-response residual when the variance is constant. Here we deliberately
use **the same descriptive one-SE bars** for both displays, so the whole
plotted table agrees. Those bars are not full fitted-model uncertainty intervals.

```{r implied-log-agreement, eval=has_glmmTMB, fig.height=9, fig.cap="Ordinary log-residual implied effects (left) and the new likelihood-based effects (right) for the same Gaussian log-CPUE glmmTMB fit. Both use the same year-plus-season baseline, fitted vessel effects, and descriptive one-SE bars. Points, trajectories, and bars agree numerically to 1e-10. Grey lines are fixed baseline contributions, not independently fitted seasonal indices. The deliberately omitted seasonal trend is visible as departures from those baselines.", fig.alt="Side-by-side identical traditional log-residual and likelihood implied-effect plots, each with three seasonal panels."}
patchwork::wrap_plots(
  plot(traditional_log, ncol = 1) + labs(title = "Ordinary log-residual construction",
    subtitle = "Year + season | one descriptive SE", caption = NULL),
  plot(new_log, ncol = 1) + labs(title = "Likelihood adjustment",
    subtitle = "Year + season | one descriptive SE", caption = NULL), ncol = 2
)
```

The optional `glmmTMB` examples are skipped if that package is unavailable.
This equality is not a promise that every lognormal parameterisation behaves
identically. For example, glmmTMB's directly specified `lognormal()` family
models mean and SD on the response scale: holding that SD fixed is different
from holding log-scale SD fixed. This first implied-effect implementation
therefore supports the explicit Gaussian-log-response route, not an automatic
conversion of directly parameterised lognormal models.

### Reproduce the actual historical standardised-residual recipe

For a direct source comparison, fit a plain Gaussian GLM to log CPUE: that is
one of the classes handled by the historical `Diagnoser`. Both methods below
use this **same** GLM. This comparison does not contrast a fixed-effect fit with
a mixed model.

```{r implied-historical}
historical_fit <- glm(log(cpue_positive) ~ year + season + depth + soak,
  family = gaussian(), data = example)
historical <- implied_effects(historical_fit, groups = "season",
  method = "traditional", traditional_scale = "standardised")
modern <- implied_effects(historical_fit, groups = "season")

# Independent reconstruction of the historical GLM arithmetic.
term_effects <- predict(historical_fit, type = "terms")
reference <- rowSums(term_effects[, c("year", "season")])
standardised <- rstandard(historical_fit)
standardised <- standardised - mean(standardised)
manual <- do.call(rbind, lapply(seq_len(nrow(historical$table)), function(j) {
  cell <- historical$table[j, ]
  keep <- example$year == cell$level & example$season == cell$group
  data.frame(estimate = mean(reference[keep] + standardised[keep]),
    std_error = sd(standardised[keep]) / sqrt(sum(keep)))
}))
max(abs(manual$estimate - historical$table$estimate))
stopifnot(isTRUE(all.equal(manual$estimate, historical$table$estimate, tolerance = 1e-10)))
stopifnot(isTRUE(all.equal(manual$std_error, historical$table$std_error, tolerance = 1e-10)))
```

```{r implied-historical-comparison, fig.height=9, fig.cap="Actual historical analyser GLM convention (left) versus the new likelihood-based implied effects (right), using the same Gaussian log-CPUE GLM and baseline. Left: globally centred rstandard residuals added to term contributions, with descriptive one-SE bars. Right: local log-scale likelihood shifts, with 95% conditional profile-likelihood intervals. Unlike the preceding ordinary-log-residual comparison, these points and bars need not agree: both the residual scale and interval definition differ. The standardised convention is retained as a labelled historical comparison, not a physically interpretable CPUE multiplier.", fig.alt="Historical standardised-residual and new likelihood implied-effect plots show different adjustment scales and interval definitions."}
patchwork::wrap_plots(
  plot(historical, ncol = 1) + labs(title = "Historical standardised convention",
    subtitle = "Year + season | one descriptive SE", caption = NULL),
  plot(modern, ncol = 1) + labs(title = "Likelihood adjustment",
    subtitle = "Year + season | 95% conditional intervals", caption = NULL), ncol = 2
)
```

This reproduces the source's GLM arithmetic, not every version of earlier
influ2 or every fisheries report. In particular, it does not revive the old
brms posterior-array helper or infer what every report meant by "standardised".

## Negative-binomial implied effects

For an NB2 log-link model, the local shift maximises the negative-binomial
likelihood using the fitted size/dispersion. It is not the average of PIT
scores or a renamed Pearson residual. Within a stratum, adding delta to the
linear predictor multiplies all its fitted means by `exp(delta)`.

```{r implied-nb-model, eval=has_glmmTMB}
nb_fit <- glmmTMB::glmmTMB(
  count_cpue ~ year + season + depth + soak + (1 | vessel),
  family = glmmTMB::nbinom2(), data = example
)
nb_implied <- implied_effects(nb_fit, groups = "season")
head(as.data.frame(nb_implied))
```

```{r implied-nb, eval=has_glmmTMB, fig.height=5, fig.cap="New likelihood-based residual-implied effects for the NB2 glmmTMB count-CPUE model, holding fitted vessel effects, dispersion, and all original coefficients fixed. Grey lines show centred year-plus-season contributions; purple trajectories add one local log-mean adjustment per season-year cell. Bars are 95% conditional profile-likelihood intervals for those adjustments, not uncertainty intervals for fully refitted interactions. Point area represents sample size. No response simulations or MCMC are used.", fig.alt="Three seasonal trajectories compare NB2 residual-implied effects with their fitted annual baselines."}
plot_implied_residuals(nb_implied)
```

For routine use, the shortcut computes the same default result:

```{r implied-shortcut, eval=FALSE}
plot_implied_residuals(nb_fit, groups = "season")
plot_implied_residuals(log_fit, groups = "season", method = "traditional")
```

Keeping `nb_implied` separates calculation from styling and supports save/reload
without the model. Its table retains sparse, empty, and boundary cells with
explicit statuses. Lines never bridge omitted cells. For an all-zero count
cell, the likelihood optimum is a zero mean (log shift `-Inf`); the table flags
that boundary and the plot omits the non-finite point, rather than inventing
a finite value by adding a pseudocount.

## Interpretation, uncertainty, and scope

The current adapters support `lm`, GLM, GAM, and ML glmmTMB for Gaussian
identity-link, Poisson log-link, and NB2 log-link models. Fitted GAM smooths and
mixed-model effects stay fixed. Traditional comparison is restricted to
constant-variance Gaussian models of `log(response)`; the historical
standardised option additionally requires a plain GLM.

Conditional profile intervals ignore uncertainty in the original model,
its baseline, and its estimated latent effects. Descriptive one-SE bars also
ignore dependence. Neither should be read as full uncertainty in a regional
index or an interaction. A suitable refit/bootstrap or explicitly defined
posterior propagation would be a further development, not something these
bars already provide.

Other families, direct lognormal parameterisations, brms, sdmTMB, tinyVAST,
non-unit weights, and year interactions currently fail explicitly for this
**new implied-effect calculation**. They remain supported where documented by
the existing PIT diagnostics and other influ2 functions. Joint hurdle/delta
models require separate decisions about encounter, positive-response, and
combined-response shifts; a positive-component calculation is never silently
substituted for a combined-response diagnostic.

The tests independently reconstruct the historical recipe, check log-response
agreement, compare NB2 shifts and profile endpoints with native density
calculations, and cover offsets, fitted random effects, GAMs, row alignment,
unsupported cases, and compact save/reload. This validates the implemented
arithmetic, not universal scientific calibration of implied-effect intervals.

## References
