---
title: "Residual-implied effects"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Residual-implied effects}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
  warning = FALSE, fig.width = 11, fig.height = 6, dev.args = list(bg = "white"))
library(influ2)
library(ggplot2)
theme_set(theme_bw())
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
```

## Restore the original question

Residual-implied plots ask whether individual groups suggest a different
annual trajectory from the fitted model. The original comparison is useful:
show a common fitted annual pattern, then show how residual departures in
each group would adjust that pattern. These are **exploratory implied effects**,
not separately fitted interactions or regional abundance indices.

`plot_implied_residuals()` now defaults to a local **likelihood adjustment**.
The original model's parameters, fitted random effects, smooths, dispersion,
and exposure offsets stay fixed. Only one additional effect-scale shift is
estimated in each supported group-year stratum. For a supported log-link model,
that shift multiplies the expected response, while respecting the fitted
response distribution. No full model is refitted and no MCMC is run.

The zero-centred PIT summaries previously displayed under this name are now
available as `plot_grouped_residuals()`. That calculation has not changed.
It answers a different question: whether a group's observations tend to lie
high or low within their predictive distributions. **Normal-score PIT
residuals are not added to model coefficients.**

## Three conventions, explicitly distinguished

| Calculation | Adjustment added to the selected fitted baseline | Interpretation |
|:--|:--|:--|
| `method = "likelihood"` | Local shift maximising the fitted response likelihood | Default effect-scale diagnostic |
| `method = "traditional"` | Mean ordinary log-response residual | Classical lognormal construction; constant log variance required |
| Traditional with `traditional_scale = "standardised"` | Mean globally centred `rstandard()` residual | Actual historical analyser GLM convention; retained for comparison, not a universal effect-scale calculation |

The [historical analyser source](https://github.com/trophia/analyser/blob/a82c5b41bfd0ebeece8e1c6526465da006435497/R/diagnoser.r)
uses `rstandard()` for GLMs, centres those residuals globally, and adds selected
fitted term contributions. Its `implieds()` helper can include the group main
effect as well as year. Thus "traditional" was not one universal residual
definition. Fisheries examples include @StarrKendrick2019 and @Middleton2025.

The new calculation agrees with **ordinary log-residual arithmetic** when
the variance is constant; it does not generally equal the historical
standardised-residual arithmetic. The examples below demonstrate both claims.

The baseline defaults to `"year_group"`: centred fixed year and group main
effects, where the latter is present. Choose `baseline = "year"` for only the
shared year term. All selected term contributions are centred over the same
fitted observations. They are not independently re-centred in each panel.
Other covariate effects, random effects, and smooths stay in the observation
predictions used to calculate the adjustment, but are not added to this baseline.

## A simulated lobster design with differing group trends

We retain the uneven year/month coverage, depth, and soak time from
`lobsters_per_pot`, but simulate new responses for this comparison. These are
not modifications to the package dataset. Three seasonal groups are defined
from month, and their trends intentionally differ. Vessel effects add
repeated-observation structure. Both responses represent CPUE per standardised
sampling unit: one continuous positive response and one count response.

```{r implied-data}
data(lobsters_per_pot)
example <- lobsters_per_pot
example$season <- factor(ceiling(as.integer(example$month) / 4),
  labels = c("Early season", "Mid-season", "Late season"))
set.seed(12092026)
example$vessel <- factor(sample(1:24, nrow(example), replace = TRUE))
time <- as.integer(example$year)
trend <- time - mean(time)
vessel_effect <- rnorm(24, sd = .35)
eta <- 1.5 + .15 * sin(time / 2) - .015 * (example$depth - 40) +
  .008 * (example$soak - 30) + .15 * (as.integer(example$season) - 2) +
  .07 * trend * (as.integer(example$season) - 2) +
  vessel_effect[as.integer(example$vessel)]
example$cpue_positive <- exp(eta + rnorm(nrow(example), sd = .6))
example$count_cpue <- rnbinom(nrow(example), mu = exp(eta), size = 4)
```

### Constant log variance: matching the classical construction

Fit a Gaussian model to log CPUE, omitting the generating season-by-year
interaction. This is a lognormal response model with constant log-scale
variance, not a Gaussian model fitted directly to untransformed CPUE.

```{r implied-log-model, eval=has_glmmTMB}
log_fit <- glmmTMB::glmmTMB(
  log(cpue_positive) ~ year + season + depth + soak + (1 | vessel),
  family = gaussian(), data = example
)
new_log <- implied_effects(log_fit, groups = "season", interval = "descriptive")
traditional_log <- implied_effects(log_fit, groups = "season", method = "traditional")
max(abs(new_log$table$estimate - traditional_log$table$estimate))
stopifnot(isTRUE(all.equal(new_log$table, traditional_log$table, tolerance = 1e-10)))
```

The estimates agree because the conditional Gaussian likelihood shift is the
mean log-response residual when the variance is constant. Here we deliberately
use **the same descriptive one-SE bars** for both displays, so the whole
plotted table agrees. Those bars are not full fitted-model uncertainty intervals.

```{r implied-log-agreement, eval=has_glmmTMB, fig.height=9, fig.cap="Ordinary log-residual implied effects (left) and the new likelihood-based effects (right) for the same Gaussian log-CPUE glmmTMB fit. Both use the same year-plus-season baseline, fitted vessel effects, and descriptive one-SE bars. Points, trajectories, and bars agree numerically to 1e-10. Grey lines are fixed baseline contributions, not independently fitted seasonal indices. The deliberately omitted seasonal trend is visible as departures from those baselines.", fig.alt="Side-by-side identical traditional log-residual and likelihood implied-effect plots, each with three seasonal panels."}
patchwork::wrap_plots(
  plot(traditional_log, ncol = 1) + labs(title = "Ordinary log-residual construction",
    subtitle = "Year + season | one descriptive SE", caption = NULL),
  plot(new_log, ncol = 1) + labs(title = "Likelihood adjustment",
    subtitle = "Year + season | one descriptive SE", caption = NULL), ncol = 2
)
```

The optional `glmmTMB` examples are skipped if that package is unavailable.
This equality is not a promise that every lognormal parameterisation behaves
identically. For example, glmmTMB's directly specified `lognormal()` family
models mean and SD on the response scale: holding that SD fixed is different
from holding log-scale SD fixed. This first implied-effect implementation
therefore supports the explicit Gaussian-log-response route, not an automatic
conversion of directly parameterised lognormal models.

### Reproduce the actual historical standardised-residual recipe

For a direct source comparison, fit a plain Gaussian GLM to log CPUE: that is
one of the classes handled by the historical `Diagnoser`. Both methods below
use this **same** GLM. This comparison does not contrast a fixed-effect fit with
a mixed model.

```{r implied-historical}
historical_fit <- glm(log(cpue_positive) ~ year + season + depth + soak,
  family = gaussian(), data = example)
historical <- implied_effects(historical_fit, groups = "season",
  method = "traditional", traditional_scale = "standardised")
modern <- implied_effects(historical_fit, groups = "season")

# Independent reconstruction of the historical GLM arithmetic.
term_effects <- predict(historical_fit, type = "terms")
reference <- rowSums(term_effects[, c("year", "season")])
standardised <- rstandard(historical_fit)
standardised <- standardised - mean(standardised)
manual <- do.call(rbind, lapply(seq_len(nrow(historical$table)), function(j) {
  cell <- historical$table[j, ]
  keep <- example$year == cell$level & example$season == cell$group
  data.frame(estimate = mean(reference[keep] + standardised[keep]),
    std_error = sd(standardised[keep]) / sqrt(sum(keep)))
}))
max(abs(manual$estimate - historical$table$estimate))
stopifnot(isTRUE(all.equal(manual$estimate, historical$table$estimate, tolerance = 1e-10)))
stopifnot(isTRUE(all.equal(manual$std_error, historical$table$std_error, tolerance = 1e-10)))
```

```{r implied-historical-comparison, fig.height=9, fig.cap="Actual historical analyser GLM convention (left) versus the new likelihood-based implied effects (right), using the same Gaussian log-CPUE GLM and baseline. Left: globally centred rstandard residuals added to term contributions, with descriptive one-SE bars. Right: local log-scale likelihood shifts, with 95% conditional profile-likelihood intervals. Unlike the preceding ordinary-log-residual comparison, these points and bars need not agree: both the residual scale and interval definition differ. The standardised convention is retained as a labelled historical comparison, not a physically interpretable CPUE multiplier.", fig.alt="Historical standardised-residual and new likelihood implied-effect plots show different adjustment scales and interval definitions."}
patchwork::wrap_plots(
  plot(historical, ncol = 1) + labs(title = "Historical standardised convention",
    subtitle = "Year + season | one descriptive SE", caption = NULL),
  plot(modern, ncol = 1) + labs(title = "Likelihood adjustment",
    subtitle = "Year + season | 95% conditional intervals", caption = NULL), ncol = 2
)
```

This reproduces the source's GLM arithmetic, not every version of earlier
influ2 or every fisheries report. In particular, it does not revive the old
brms posterior-array helper or infer what every report meant by "standardised".

## Negative-binomial implied effects

For an NB2 log-link model, the local shift maximises the negative-binomial
likelihood using the fitted size/dispersion. It is not the average of PIT
scores or a renamed Pearson residual. Within a stratum, adding delta to the
linear predictor multiplies all its fitted means by `exp(delta)`.

```{r implied-nb-model, eval=has_glmmTMB}
nb_fit <- glmmTMB::glmmTMB(
  count_cpue ~ year + season + depth + soak + (1 | vessel),
  family = glmmTMB::nbinom2(), data = example
)
nb_implied <- implied_effects(nb_fit, groups = "season")
head(as.data.frame(nb_implied))
```

```{r implied-nb, eval=has_glmmTMB, fig.height=5, fig.cap="New likelihood-based residual-implied effects for the NB2 glmmTMB count-CPUE model, holding fitted vessel effects, dispersion, and all original coefficients fixed. Grey lines show centred year-plus-season contributions; purple trajectories add one local log-mean adjustment per season-year cell. Bars are 95% conditional profile-likelihood intervals for those adjustments, not uncertainty intervals for fully refitted interactions. Point area represents sample size. No response simulations or MCMC are used.", fig.alt="Three seasonal trajectories compare NB2 residual-implied effects with their fitted annual baselines."}
plot_implied_residuals(nb_implied)
```

For routine use, the shortcut computes the same default result:

```{r implied-shortcut, eval=FALSE}
plot_implied_residuals(nb_fit, groups = "season")
plot_implied_residuals(log_fit, groups = "season", method = "traditional")
```

Keeping `nb_implied` separates calculation from styling and supports save/reload
without the model. Its table retains sparse, empty, and boundary cells with
explicit statuses. Lines never bridge omitted cells. For an all-zero count
cell, the likelihood optimum is a zero mean (log shift `-Inf`); the table flags
that boundary and the plot omits the non-finite point, rather than inventing
a finite value by adding a pseudocount.

## Simulation check: what do these implied effects recover?

The **IV01** follow-up checks this question in 400 simulated NB2 glmmTMB fits:
100 independent datasets for each combination of balanced/uneven sampling
and no interaction/an omitted year-by-season trend. Each dataset has six
years, three seasons, 12 vessel random intercepts, and 648 observations.
Balanced cells each contain 36 observations; the uneven design contains one
empty cell, five cells below `min_n = 10`, and 12 supported cells. Thus the
comparison changes coverage, not total sample size. The generating size is 4,
vessel SD is 0.4, and the largest injected log-mean shift is 0.6.

Every fit retains year, season, depth, and vessel effects, but not their
year-by-season interaction. No failed fit is replaced. The
[fixed protocol and reproduction scripts](https://github.com/quantifish/influ2/tree/master/tools/implied-validation)
record the design, seed schedule, source hashes, and independent numerical
checks. This article reads a frozen compact result: **it does not refit the
400 models**, and neither does the package test suite.

```{r implied-validation-results}
validation <- readRDS(system.file("extdata", "implied-validation.rds", package = "influ2"))
knitr::kable(validation$fit_summary[, c("sampling", "signal", "attempted",
  "eligible", "errors", "warnings")],
  col.names = c("Sampling", "Interaction", "Attempted", "Eligible", "Errors", "Warnings"))
```

All 400 fits were eligible, and all 800 calculations (two routes per fit)
succeeded without warnings. Empty and sparse cells remained explicit; no
all-zero boundary cells occurred in this run. Separate deterministic tests
exercise that boundary behaviour.

### Separate the injected pattern from the fitted-conditional target

There are two checks on each dataset:

1. **Known-parameter control:** fix the true additive predictor, realised
   vessel effects, and size. The injected shift is then exactly the quantity
   the local likelihood is estimating.
2. **Fitted-model calculation:** use the public `implied_effects()` result
   from the fitted model. Its main effects and dispersion can absorb some
   omitted structure. The appropriate local target is the shift maximising
   *expected* likelihood under the generating response distribution, with
   that fitted predictor and size held fixed.

For the second route, an independent one-dimensional optimisation obtains
the target from the known generating means. Equivalently, its expected score
is zero:

\[
\sum_{i\ \mathrm{in\ cell}}
\frac{\widehat\phi_i\{\mu_i^{\mathrm{true}}-
  \exp(\widehat\eta_i+\delta)\}}
 {\widehat\phi_i+\exp(\widehat\eta_i+\delta)}=0.
\]

That target depends on the fitted data. Its interval **containment** is not
ordinary confidence coverage for a fixed population interaction, and adding
the displayed baseline does not propagate uncertainty in that baseline.

The following example is production replicate 1: the first dataset with all
four fits and calculations eligible, selected before inspecting its curves.
The teal marks are conditional targets added to the **same fitted baseline**,
not independently centred true annual indices.

```{r implied-validation-example, fig.height=8, fig.cap="Frozen IV01 uneven-sampling example: no omitted interaction (top) and an omitted seasonal trend (bottom). Purple points and lines are the public residual-implied effects, with 95% conditional profile intervals; grey lines are their fixed fitted baselines. Teal dashed lines and crosses add the independently calculated expected-likelihood target to the same baseline. Each dataset has 648 observations, but one empty and five sparse cells are omitted, with no lines spanning gaps. Replicate 1 was selected before visual inspection. These are local implied effects, not refitted interactions or regional indices.", fig.alt="Two rows of three seasonal panels compare fitted implied effects, fixed baselines, and known conditional targets, retaining gaps in uneven sampling."}
example_plots <- lapply(c("null", "trend"), function(signal) {
  case <- validation$examples$cases[[paste0("uneven:", signal)]]
  p <- plot(case$result)
  shown <- p$data[p$data$status == "ok", ]
  key <- function(d) paste(d$level, d$group, sep = ":")
  shown$target <- case$truth$target[match(key(shown), key(case$truth))]
  p + geom_line(data = shown, aes(y = baseline + target),
    colour = "#008080", linetype = 2) +
    geom_point(data = shown, aes(y = baseline + target),
      colour = "#008080", shape = 4, size = 2) +
    labs(title = if (signal == "null") "No omitted interaction" else "Omitted seasonal trend",
      subtitle = "Purple: implied effect | Teal: conditional target | Grey: fitted baseline",
      caption = NULL, x = "Year")
})
patchwork::wrap_plots(example_plots, ncol = 1)
```

### Interval behaviour, without turning the plot into a significance test

The table gives percentages, with **one Monte Carlo standard error** in
parentheses. Summaries first average over supported cells within each
dataset, then over 100 independent datasets. Monte Carlo errors use variation
between datasets; they do not treat the cells within a fitted model as
independent replicates.

```{r implied-validation-intervals}
rates <- validation$summary
percentage <- function(value, se) sprintf("%.1f (%.1f)", 100 * value, 100 * se)
knitr::kable(data.frame(
  Sampling = rates$sampling, Interaction = rates$signal,
  Route = ifelse(rates$route == "known_parameters", "Known parameters", "Fitted model"),
  `Target in interval (%)` = percentage(rates$containment, rates$containment_mcse),
  `Cells excluding zero (%)` = percentage(rates$zero_exclusion, rates$zero_exclusion_mcse),
  `Datasets with any exclusion (%)` = percentage(rates$any_zero_exclusion, rates$any_zero_exclusion_mcse),
  check.names = FALSE))
```

Under known parameters, pointwise coverage was 94.4--95.0%, close to the
nominal 95% in these cases. Fitted-conditional target containment was
94.2--96.4%, but that is a different, data-dependent target. Under the null,
only about 0.8--0.9% of fitted-model cell intervals excluded zero, and 9--12%
of datasets had at least one exclusion. By contrast, the known-parameter
control had a zero exclusion somewhere in 49--65% of null datasets: many
pointwise checks are not a simultaneous test. **Neither route provides a
calibrated whole-model test of a missing interaction.**

### Uneven sampling changes what remains for the diagnostic to detect

The next figure compares the average local adjustment and its appropriate
conditional target with the raw injected pattern. It shows all seasons,
including the middle season with no injected trend. Missing/sparse cells
have no diagnostic estimate; an input truth curve does not fill those gaps.

```{r implied-validation-recovery, fig.height=7, fig.cap="IV01 omitted-trend cases, averaged over 100 independent datasets per sampling design. Purple points and solid lines show average fitted-model local log-mean adjustments; teal crosses and dashed lines show average fitted-conditional expected-likelihood targets; grey dotted lines show the raw injected seasonal pattern. Top: balanced sampling; bottom: uneven sampling with the same total observations. Unsupported cells have no purple or teal estimate, and those lines do not bridge gaps. Fitted main effects and nuisance parameters absorb more of the injected pattern under uneven sampling, so raw interaction recovery is not the same estimand as the local conditional adjustment. These are mean curves, not confidence bands.", fig.alt="Balanced and uneven seasonal facets show close agreement between mean adjustments and conditional targets, with stronger separation from raw injected trends under uneven sampling."}
recovery <- subset(validation$cell_summary, signal == "trend" & route == "fitted_model")
recovery$year <- as.integer(recovery$level)
recovery$group <- factor(recovery$group, levels = validation$metadata$settings$seasons)
recovery <- recovery[order(recovery$sampling, recovery$group, recovery$year), ]
recovery$segment <- cumsum(c(TRUE, diff(recovery$year) != 1L |
  head(recovery$group, -1) != tail(recovery$group, -1) |
  head(recovery$sampling, -1) != tail(recovery$sampling, -1) |
  head(recovery$usable, -1) == 0 | tail(recovery$usable, -1) == 0))
supported <- subset(recovery, usable > 0)
ggplot(recovery, aes(year)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_line(aes(y = injected, colour = "Injected pattern", linetype = "Injected pattern")) +
  geom_line(data = supported, aes(y = adjustment, group = segment,
    colour = "Mean adjustment", linetype = "Mean adjustment")) +
  geom_point(data = supported, aes(y = adjustment), colour = "purple4", size = 2) +
  geom_line(data = supported, aes(y = target, group = segment,
    colour = "Conditional target", linetype = "Conditional target")) +
  geom_point(data = supported, aes(y = target), colour = "#008080", shape = 4, size = 2) +
  facet_grid(sampling ~ group) +
  scale_colour_manual(values = c("Injected pattern" = "grey50",
    "Mean adjustment" = "purple4", "Conditional target" = "#008080")) +
  scale_linetype_manual(values = c("Injected pattern" = 3,
    "Mean adjustment" = 1, "Conditional target" = 2)) +
  scale_x_continuous(breaks = 2011:2016) +
  labs(x = "Year", y = "Local log-mean shift", colour = NULL, linetype = NULL) +
  theme(legend.position = "bottom")
```

Across the two signal scenarios, the mean within-dataset RMSE around the
fitted-conditional target was 0.118 and 0.139 log units for balanced and
uneven sampling. The corresponding target-versus-injected-pattern RMSE was
0.107 and 0.317. Thus the attenuated uneven-sampling display is not evidence
that the local optimiser lost the signal: much of it has already entered
the fitted baseline or other parameters. Mean fitted NB2 size fell from
4.18/4.20 in the null cases to 2.79/3.47 in the trend cases; omitted structure
was also accommodated by greater estimated overdispersion.

For supported cells with pre-specified absolute injected shift at least 0.3,
the fitted adjustments recovered the injected direction 100.0% and 98.3%
of the time, but their intervals excluded zero only 92.5% and 32.8% of the
time, respectively. These are descriptive results at **one effect size**,
not a general power calculation. The supported strong cells differ between
designs, and the six unsupported uneven cells cannot contribute evidence.

```{r implied-validation-support}
widths <- subset(validation$cell_summary, signal == "trend" & route == "fitted_model")
width_summary <- aggregate(width ~ sampling + n, widths, mean)
knitr::kable(width_summary, digits = 3,
  col.names = c("Sampling", "Observations per supported cell", "Mean conditional interval width (log units)"))
```

Small supported cells generally have wider intervals. Empty and sparse cells
are intentionally absent from the width table, not assigned zero width.
Larger record counts do not by themselves validate ignoring uncertainty in
estimated vessel effects or other fitted parameters.

## Positive CPUE: Gamma with a log link

Gamma(log) fits use the same effect-scale question, without treating continuous
CPUE as a count or adding dimensionless PIT scores to coefficients. Write the
original fitted mean as `mu_i`, and the native fitted scale as `phi_i`, so that
the variance is `phi_i * mu_i^2` and the shape is `k_i = 1 / phi_i`. The local
adjustment is

\[
\widehat\delta = \log\left\{\frac{\sum_i k_i y_i/\mu_i}{\sum_i k_i}\right\}.
\]

For a constant fitted scale, this simplifies to `log(mean(response / fitted_mean))`.
Thus `exp(adjustment) = 1.5` suggests a 50% upward adjustment relative to the
original predictions in that cell. It is not a separately estimated regional
CPUE index. The plotted trajectory is the centred fixed baseline **plus** this
adjustment, not the adjustment alone.

The GAM adapter retains [`sig2`](https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/gamObject.html).
The GLM adapter uses `summary(fit)$dispersion`. glmmTMB's native Gamma
[dispersion prediction](https://glmmtmb.github.io/glmmTMB/reference/sigma.glmmTMB.html)
is `1 / sqrt(shape)`, so influ2 squares it to obtain `phi`. No alternative
shape estimate is fitted. The conditional intervals solve the Gamma
likelihood-ratio equation with these original scales held fixed; they are
asymmetric on the log-effect scale.

This small GAM example deliberately leaves differing seasonal trends out
of the fitted model. It includes a smooth, a vessel effect, an effort offset,
and an additional estimated effort slope, all retained in the predictions.

```{r implied-gamma-model, eval=requireNamespace("mgcv", quietly=TRUE)}
set.seed(821)
example$log_effort <- log(example$soak / mean(example$soak))
example$gamma_cpue <- rgamma(nrow(example), shape = 1.4,
  scale = exp(eta + 1.3 * example$log_effort) / 1.4)
gamma_fit <- mgcv::gam(gamma_cpue ~ year + season + s(depth, k = 5) +
  s(vessel, bs = "re") + log_effort + offset(log_effort),
  family = Gamma(link = "log"), method = "REML", data = example)
gamma_implied <- implied_effects(gamma_fit, groups = "season")
head(as.data.frame(gamma_implied))
```

```{r implied-gamma-plot, eval=requireNamespace("mgcv", quietly=TRUE), fig.height=5, fig.cap="Gamma(log) residual-implied effects for simulated positive CPUE. Grey lines show the centred fixed year-plus-season baseline; purple trajectories add the local Gamma likelihood adjustment. Bars are 95% conditional profile-likelihood intervals, holding the original coefficients, smooths, vessel effects, offsets, and native fitted scale fixed. They exclude uncertainty in the original fit and residual dependence. Point area represents sample size; no response simulation or model refitting is needed to calculate these diagnostics.", fig.alt="Three seasonal panels show Gamma implied-effect trajectories and conditional intervals against their fitted baselines."}
plot(gamma_implied)
```

The Gamma arithmetic is tested against independent `dgamma()` optimisation and
profile endpoints, including unequal fixed scales. Tests also check response-unit
invariance, missing-year gaps, sparse cells, row alignment, and saved-result
plotting. This is numerical validation, **not** a simulation calibration of Gamma
interval coverage. The NB2 study above does not supply that calibration.
Strictly positive responses and a log link are required. An entire joint delta
model is not silently replaced by its positive Gamma component.

## Interpretation, uncertainty, and scope

The current adapters support `lm`, GLM, GAM, and ML glmmTMB for Gaussian
identity-link, Poisson log-link, NB2 log-link, and Gamma log-link models. Fitted GAM smooths and
mixed-model effects stay fixed. Traditional comparison is restricted to
constant-variance Gaussian models of `log(response)`; the historical
standardised option additionally requires a plain GLM.

Conditional profile intervals ignore uncertainty in the original model,
its baseline, and its estimated latent effects. Descriptive one-SE bars also
ignore dependence. Neither should be read as full uncertainty in a regional
index or an interaction. A suitable refit/bootstrap or explicitly defined
posterior propagation would be a further development, not something these
bars already provide.

Other families, direct lognormal parameterisations, brms, sdmTMB, tinyVAST,
non-unit weights, and year interactions currently fail explicitly for this
**new implied-effect calculation**. They remain supported where documented by
the existing PIT diagnostics and other influ2 functions. Joint hurdle/delta
models require separate decisions about encounter, positive-response, and
combined-response shifts; a positive-component calculation is never silently
substituted for a combined-response diagnostic.

The tests independently reconstruct the historical recipe, check log-response
agreement, compare NB2 and Gamma shifts and profile endpoints with native density
calculations, and cover offsets, fitted random effects, GAMs, row alignment,
unsupported cases, and compact save/reload. This validates the implemented
arithmetic, not universal scientific calibration of implied-effect intervals.
IV01 adds bounded empirical evidence for NB2 glmmTMB only: one effect size,
two fixed sampling designs, and 100 datasets per scenario. It does not
calibrate Gamma intervals, validate binomial, spatial, Bayesian, or combined
two-part implied effects, nor calibrate these plots as formal significance tests.

## References
