---
title: "influ2"
author: "Darcy Webber"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{influ2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `influ2` package is based on the original `influ` package which was developed to generate influence statistics and associated plots. The `influ` package has functions that use outputs from the `glm` function. The `influ2` package was developed to use outputs from `brms` and relies heavily on the R packages `ggplot2` and `dplyr`. This vignette showcases the `influ2` package.


# Functions for exploring data

In this vignette I use the `iris` data set to generate a fake catch per unit effort (CPUE) data set. This data set has a response variable labelled `CPUE` and has the dependent variables `Year` (from 1990 to 2014), `Species` (3 species), `Area` (3 areas) and `Duration`. No step-wise variable selection or model selection is done in this vignette.

The main function for exploring data in `influ2` is `plot_bubble` which presents the number of records by two user defined groups within a data set. The bubble size in `plot_bubble` represents the number of records across the entire data set, and bubbles can either be a single colour or coloured by a third variable in the data set.

```{r echo=TRUE, message=FALSE}
library(knitr)
library(tidyverse)
library(reshape2)
library(brms)
library(influ2)
library(bayesplot)

# options(mc.cores = parallel::detectCores())

# Get Nokome's influ package code
# source("tests/testthat/influ.R")
source("../tests/testthat/influ.R")

# Get some data to use
data(iris)
glimpse(iris)

iris2 <- iris %>% 
  mutate(CPUE = Petal.Length, 
         Year = factor(Sepal.Width * 10 + 1970), 
         Area = factor(round(Petal.Width)), 
         Duration = Sepal.Length,
         Duration2 = cut(Sepal.Length, 20)) %>%
  select(CPUE, Year, Species, Area, Duration, Duration2)
glimpse(iris2)
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Distribution of data by year and duration."}
plot_bubble(df = iris2, group = c("Year", "Duration2"), fill = "green")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Distribution of data by year, species, and area."}
plot_bubble(df = iris2, group = c("Year", "Species"), fill = "Area")
```

# The origninal influ versus influ2

First I fit three different models of increasing complexity in `brms`. The third model includes the explanatory variables `Year`, `Species` and `Area`. I then fit another four models which include `Year` and different forms of the variable `Duration`. The different forms include a linear predictor, a third-order polynomial, a spline, and a random-effect. A few different distributions are used to mix things up as well including lognormal, Gaussian, Gamma, and Weibull.

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
# Fit a series of models using brms
# if (getwd() != "/home/darcy/Projects/influ2/vignettes") {
if (FALSE) {
  # A set of lognormal models
  fit0 <- brm(CPUE ~ Year, data = iris2, family = lognormal(), refresh = 0, iter = 1000)
  save(fit0, file = "mcmc0.rda")
  fit1 <- brm(CPUE ~ Year + Duration, data = iris2, family = lognormal(), refresh = 0, iter = 1000)
  save(fit1, file = "mcmc1.rda")
  fit2 <- brm(CPUE ~ Year + Duration + Species, data = iris2, family = lognormal(), refresh = 0, iter = 1000)
  save(fit2, file = "mcmc2.rda")
  
  # Gaussian, Gamma, and Weibull
  fit3 <- brm(CPUE ~ Year + poly(Duration, 3), data = iris2, refresh = 0, iter = 1000)
  save(fit3, file = "mcmc3.rda")
  fit4 <- brm(CPUE ~ Year + s(Duration), data = iris2, family = Gamma(link = "log"), refresh = 0, iter = 1000)
  save(fit4, file = "mcmc4.rda")
  fit5 <- brm(CPUE ~ Year + (1 | Duration2), data = iris2, family = weibull, refresh = 0, iter = 1000)
  save(fit5, file = "mcmc5.rda")
  fit6 <- brm(CPUE ~ Year + Duration2, data = iris2, family = weibull, refresh = 0, iter = 1000)
  save(fit6, file = "mcmc6.rda")
} else{
  load("mcmc0.rda")
  load("mcmc1.rda")
  load("mcmc2.rda")
  load("mcmc3.rda")
  load("mcmc4.rda")
  load("mcmc5.rda")
  load("mcmc6.rda")
}

# Fit a model using glm and generate influence statistics using the original influ package
fit_glm <- glm(log(CPUE) ~ Year + Duration + Species, data = iris2)
myInfl <- Influence$new(fit_glm)
myInfl$calc()
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "The original CDI plot from the influ package."}
myInfl$cdiPlot(term = "Species")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "The new Bayesian CDI plot from the influ2 package."}
# Generate a CDI plot for a factor
plot_bayesian_cdi(fit = fit2, xfocus = "Species", yfocus = "Year", xlab = "Species")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "The original step-plot from the influ package."}
myInfl$stepPlot()
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "The new Bayesian step-plot from the influ2 package. Note that the new step-plot requires that all models/steps be run in brms before the function can be used."}
fits <- list(fit0, fit1, fit2)
plot_step(fits = fits, year = "Year")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
plot_index(fit2, year = "Year")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
myInfl$stanPlot()
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
plot_influ(fit2, year = "Year")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
myInfl$influPlot()
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Comparison of influences produced by the original influ package (points) and the new influ2 calculation (lines)."}
i1 <- myInfl$influences %>%
  pivot_longer(cols = -level, names_to = "variable")
i_species <- get_influ(fit = fit2, group = c("Year", "Species")) %>%
  group_by(Year) %>%
  summarise(delta = mean(delta)) %>%
  mutate(variable = "Species")
i_duration <- get_influ(fit = fit2, group = c("Year", "Duration")) %>%
  group_by(Year) %>%
  summarise(delta = mean(delta)) %>%
  mutate(variable = "Duration")

ggplot(data = rbind(i_species, i_duration)) +
  geom_point(data = i1, aes(x = level, y = exp(value))) +
  geom_line(aes(x = Year, y = exp(delta), colour = variable, group = variable)) +
  facet_wrap(variable ~ ., ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Influence") +
  theme_bw() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```


# Functions for exploring models

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
# Here I evaluate model fit using loo and waic. 
# Other options include kfold, loo_subsample, bayes_R2, loo_R2 and marglik
criterion <- c("loo", "waic", "bayes_R2")
fit0 <- add_criterion(fit0, criterion = criterion)
fit1 <- add_criterion(fit1, criterion = criterion)
fit2 <- add_criterion(fit2, criterion = criterion)
fit3 <- add_criterion(fit3, criterion = criterion)
# fit4 <- add_criterion(fit4, criterion = criterion)
# fit5 <- add_criterion(fit5, criterion = criterion)
# fit6 <- add_criterion(fit6, criterion = criterion)

fit0$criteria$loo
fit0$criteria$waic
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
loo_compare(fit0, fit1, fit2, criterion = "loo") %>%
  kable(digits = 1)
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
loo_compare(fit0, fit1, fit2, criterion = "waic") %>%
  kable(digits = 1)
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
# bayes_R2 for the three model runs
get_bayes_R2(fits)
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "CDI plot for a linear variable."}
plot_bayesian_cdi(fit = fit1, xfocus = "Duration", yfocus = "Year", xlab = "Duration")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "CDI plot for a polynomial."}
plot_bayesian_cdi(fit = fit3, xfocus = "Duration", yfocus = "Year", xlab = "Duration")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "CDI plot for a spline."}
plot_bayesian_cdi(fit = fit4, xfocus = "Duration", yfocus = "Year", xlab = "Duration")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "CDI plot for a random-effect."}
plot_bayesian_cdi(fit = fit5, xfocus = "Duration2", yfocus = "Year", xlab = "Duration")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "CDI plot for a fixed-effect."}
plot_bayesian_cdi(fit = fit6, xfocus = "Duration2", yfocus = "Year", xlab = "Duration")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Comparison plot of all model runs."}
fits <- list(fit1, fit3, fit4, fit5, fit6)

plot_compare(fits = fits, 
             labels = c("linear", "poly", "spline", "random-effect", "fixed-effect"), 
             year = "Year")
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Unstandardised versus standardised series."}
plot_index(fit = fit4, year = "Year")
```

# Diagnostics

The `influ2` package is based on the original `influ` package which was developed to generate influence plots. The `influ` package has functions that use outputs from the `glm` function. The `influ2` package was developed to use outputs from `brms` and relies heavily on the R packages `ggplot2` and `dplyr`. This vignette showcases the `influ2` package with a focus on model diagnostics including posterior predictive distributions, residuals, and quantile-quantile plots.

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Comparison of the empirical distribution of the data (y) to the distributions of simulated/replicated data (yrep) from the posterior predictive distribution."}
yrep <- posterior_predict(fit4, nsamples = 100)

ppc_dens_overlay(y = iris2$CPUE, yrep = yrep) + 
  labs(x = "CPUE", y = "Density") +
  theme_bw()
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Posterior predictive check."}
ppc_ecdf_overlay(y = iris2$CPUE, yrep = yrep) +
  coord_cartesian(xlim = c(0, 10)) +
  theme_bw()
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Quantile-quantile plot from LOO."}
loo <- loo(fit4, save_psis = TRUE)
psis <- loo$psis_object
lw <- weights(psis)
yrep <- posterior_predict(fit4)
ppc_loo_pit_qq(y = iris2$CPUE, yrep = yrep, lw = lw)
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Quantile-quantile plot."}
plot_qq(fit = fit4)
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Residuals and a loess smoother."}
plot_predicted_residuals(fit = fit4, trend = "loess")
```

```{r echo=TRUE, fig.height=6, fig.width=10, message=FALSE, fig.cap = "Residuals and a linear fit by year."}
plot_predicted_residuals(fit = fit4, trend = "lm") +
  facet_wrap(Year ~ .)
```

```{r echo=TRUE, fig.height=6, fig.width=10, message=FALSE, fig.cap = "Residuals and a loess smoother by year."}
# A new style of residual plot
fit <- fit4

# Extract predicted values
pred <- fitted(fit) %>% data.frame()
names(pred) <- paste0("pred.", names(pred))

# Extract residuals
resid <- residuals(fit) %>% data.frame()
names(resid) <- paste0("resid.", names(resid))

df <- cbind(resid, pred, iris2)

p <- ggplot(data = df, aes(x = Year, y = .data$resid.Estimate)) +
  geom_pointrange(aes(ymin = .data$resid.Q2.5, ymax = .data$resid.Q97.5), alpha = 0.75) +
  facet_wrap(Area ~ Species) +
    geom_hline(yintercept = 0, linetype = "dashed") +
  # geom_errorbarh(aes(xmax = .data$pred.Q2.5, xmin = .data$pred.Q97.5, height = 0), alpha = 0.75) +
  # labs(x = "Predicted values", y = "Residuals") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p
```

# Implied residuals

Implied coefficients (points) are calculated as the normalised fishing year coefficient (grey line) plus the mean of the standardised residuals in each year for each category of a variable. These values approximate the coefficients obtained when an area x year interaction term is fitted, particularly for those area x year combinations which have a substantial proportion of the records. The error bars indicate one standard error of the standardised residuals. The information at the top of each panel identifies the plotted category, provides the correlation coefficient (rho) between the category year index and the overall model index, and the number of records supporting the category.

```{r echo=TRUE, fig.height=6, fig.width=10, message=FALSE, fig.cap = "Implied residuals by year for the categorical variable Area which is not included in the model."}
plot_implied_residuals(fit = fit2, data = iris2, year = "Year", groups = "Area") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r echo=TRUE, fig.height=6, fig.width=10, message=FALSE, fig.cap = "Implied residuals by year for the categorical variable Sepcies which is included in the model."}
plot_implied_residuals(fit = fit2, data = iris2, year = "Year", groups = "Species") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# Functions for generating outputs for stock assessment

Finally, in order to fit to the data in a stock assessment model then indices can be produced using the `get_index` function. The `Estimate` and `Est.Error` columns are what should be used in stock assessment.

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE}
unstd <- get_unstandarsied(fit = fit2, year = "Year")

get_index(fit = fit2, year = "Year") %>%
  mutate(Unstandardised = unstd$Estimate) %>%
  select(Year, Unstandardised, Estimate, Est.Error, Qlower, Q50, Qupper) %>%
  kable(digits = 3)
```

```{r echo=TRUE, fig.height=6, fig.width=6, message=FALSE, fig.cap = "Simulating from a lognormal distribution and comparing with table values. Simulating with the wrong SD underestimates the uncertainty at low values."}
cpue1 <- get_index(fit = fit2, year = "Year") %>% mutate(Type = "Estimate")
cpue2 <- cpue1 %>% mutate(Type = "Simulated wrong")
cpue3 <- cpue1 %>% mutate(Type = "Simulated")

for (ii in 1:nrow(cpue1)) {
  sdd <- cpue1$Est.Error[ii] # this is wrong
  r1 <- rlnorm(n = 5000, log(cpue1$Estimate[ii]), sdd)
  cpue2$Q50[ii] <- median(r1)
  cpue2$Qlower[ii] <- quantile(r1, probs = 0.025)
  cpue2$Qupper[ii] <- quantile(r1, probs = 0.975)
  
  sdd <- log(1 + cpue1$Est.Error[ii] / cpue1$Estimate[ii])
  r1 <- rlnorm(n = 5000, log(cpue1$Q50[ii]), sdd)
  cpue3$Q50[ii] <- median(r1)
  cpue3$Qlower[ii] <- quantile(r1, probs = 0.025)
  cpue3$Qupper[ii] <- quantile(r1, probs = 0.975)
}

ggplot(data = rbind(cpue1, cpue2, cpue3), aes(x = Year, y = Q50, colour = Type)) +
  geom_pointrange(aes(ymin = Qlower, ymax = Qupper), position = position_dodge(width = 0.5)) +
  labs(x = NULL, y = "CPUE") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
