---
title: "Influence diagnostics across fisheries CPUE models"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Influence diagnostics across fisheries CPUE models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
has_mgcv <- requireNamespace("mgcv", quietly = TRUE)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE) &&
  requireNamespace("posterior", quietly = TRUE)
has_sdmTMB <- requireNamespace("sdmTMB", quietly = TRUE)
has_tinyVAST <- requireNamespace("tinyVAST", quietly = TRUE) &&
  requireNamespace("fmesher", quietly = TRUE)
```

## Why influence diagnostics?

CPUE standardisation attempts to distinguish changes in abundance from changes
in fishing practice, location, season, vessel composition, and other
explanatory variables. A fitted model may estimate these relationships well
without making it obvious why the standardised index differs from the nominal
series.

The coefficient-distribution-influence (CDI) framework combines the fitted
effect of a term with changes in the sampled distribution of that term through
the focus variable, which is usually year [@Bentley2012]. Spatial and
spatiotemporal versions can similarly show whether changes in sampled location
move observations among areas of different predicted abundance [@Hsu2022].

`influ2` represents these results with one model-neutral S3 class. Each backend
extracts model components and joint uncertainty, while the calculation,
storage, summary, and plotting layers remain shared.

## A realistic lobster CPUE example

The main examples use the simulated `lobsters_per_pot` data supplied with the
package. These are 15,642 pot records from 2000 to 2017. The response is the
number of lobsters caught per pot, and the explanatory variables are month,
depth in metres, and soak time in hours. Annual sample sizes, seasonal
coverage, depths, and soak times all vary, giving the diagnostic something
meaningful to detect.

```{r lobster-data}
library(influ2)
data(lobsters_per_pot)

dim(lobsters_per_pot)
head(lobsters_per_pot)
```

The changing distribution of records by year and month is visible before a
model is fitted. Larger bubbles represent more pot records.

```{r lobster-sampling, fig.height=6, fig.cap="Sampling effort by year and month in the simulated lobster CPUE data."}
plot_bubble(
  df = lobsters_per_pot,
  group = c("year", "month"),
  fill = "month"
) +
  ggplot2::labs(x = "Month", y = "Year") +
  ggplot2::theme(legend.position = "none")
```

Depth and soak-time coverage also change through time.

```{r lobster-covariates, fig.cap="Observed depth and soak-time distributions by year."}
covariates <- tidyr::pivot_longer(
  lobsters_per_pot,
  cols = c("depth", "soak"),
  names_to = "covariate",
  values_to = "value"
)

ggplot2::ggplot(
  covariates,
  ggplot2::aes(x = year, y = value, group = year)
) +
  ggplot2::geom_boxplot(outlier.alpha = 0.08, linewidth = 0.25) +
  ggplot2::facet_wrap(~covariate, scales = "free_y", ncol = 1) +
  ggplot2::labs(x = "Year", y = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## GLM

A Poisson GLM provides a transparent starting point. This is an example model,
not a recommendation that Poisson variation is adequate for these data.

```{r lobster-glm}
lobster_glm <- glm(
  lobsters ~ year + month + poly(depth, 3) + poly(soak, 3),
  family = poisson(link = "log"),
  data = lobsters_per_pot
)

glm_diagnostic <- influ(lobster_glm, focus = "year")
glm_diagnostic
summary(glm_diagnostic)
```

The influence plot shows how changes in sampled month, depth, and soak time
move the annual standardised series. The year term is retained in the object,
but omitted from this plot because its within-year composition is the focus,
not a sampling-distribution effect.

```{r lobster-glm-influence, fig.cap="GLM influence ratios for the changing lobster covariate distributions."}
plot(
  glm_diagnostic,
  type = "influence",
  term = c("month", "poly(depth, 3)", "poly(soak, 3)")
)
```

The same object contains nominal and standardised indices.

```{r lobster-glm-index, fig.cap="Nominal and Poisson-GLM-standardised lobster indices."}
plot(glm_diagnostic, type = "index")
```

## GAM

The `mgcv` backend works with parametric terms and smooths [@Wood2017]. Here,
negative-binomial variation is combined with smooth depth and soak effects.

```{r lobster-gam, eval=has_mgcv}
lobster_gam <- mgcv::gam(
  lobsters ~ year + month + s(depth, k = 5) + s(soak, k = 5),
  family = mgcv::nb(),
  method = "REML",
  data = lobsters_per_pot
)

gam_diagnostic <- influ(lobster_gam, focus = "year")
summary(gam_diagnostic)
```

```{r lobster-gam-influence, eval=has_mgcv, fig.cap="GAM influence ratios for month, depth, and soak time."}
plot(
  gam_diagnostic,
  type = "influence",
  term = c("month", "s(depth)", "s(soak)")
)
```

This is the same `influ_diag` interface as the GLM. No plotting code needs to
know that two of the terms are smooths.

## glmmTMB

`glmmTMB` adds mixed effects, negative-binomial models, hurdle models, and
zero-inflation [@Brooks2017]. The example treats month as a random intercept.

```{r lobster-glmmtmb, eval=has_glmmTMB}
lobster_glmmTMB <- glmmTMB::glmmTMB(
  lobsters ~ year + poly(depth, 2) + poly(soak, 2) + (1 | month),
  family = glmmTMB::nbinom2(),
  data = lobsters_per_pot
)

glmmTMB_diagnostic <- influ(lobster_glmmTMB, focus = "year")
summary(glmmTMB_diagnostic)
```

```{r lobster-glmmtmb-influence, eval=has_glmmTMB, fig.cap="glmmTMB fixed- and random-effect influence ratios."}
plot(glmmTMB_diagnostic, type = "components")
```

Fixed effects use the joint maximum-likelihood covariance. The month random
effect uses its fitted conditional modes and is labelled separately, so its
current uncertainty treatment cannot be mistaken for fixed-effect uncertainty.
For hurdle and zero-inflated fits, `influ()` also calculates population-level
unconditional-mean influence using the joint covariance of both components.

## brms

The Bayesian backend uses genuine joint posterior draws, but projects each draw
directly to the much smaller focus-by-term diagnostic. It never constructs an
observation-by-draw-by-term array [@Buerkner2017]. A model for the same lobster
data can be fitted as follows:

```{r lobster-brms-fit, eval=FALSE}
lobster_brms <- brms::brm(
  lobsters ~ year + (1 | month) + s(depth, k = 3),
  family = brms::negbinomial(),
  data = lobsters_per_pot,
  chains = 2,
  iter = 2000,
  seed = 1,
  file = "fit2",
  file_refit = "on_change"
)
```

The vignette uses the fitted model shipped with its source, rather than running
MCMC whenever the documentation is built.

```{r lobster-brms, eval=has_brms}
lobster_brms <- readRDS("fit2.rds")
brms_diagnostic <- influ(
  lobster_brms,
  focus = "year",
  ndraws = 250,
  retain = "summary"
)
summary(brms_diagnostic)
```

```{r lobster-brms-influence, eval=has_brms, fig.cap="BRMS population-level, month group-level, and depth-smooth influence ratios."}
plot(brms_diagnostic, type = "components")
```

Population-level, group-level, and smooth contributions all preserve posterior
dependence during calculation. The default object retains interval summaries;
`retain = "derived_draws"` retains only the compact diagnostic draws.

## Spatial and spatiotemporal models

The lobster data deliberately have no coordinates. Inventing coordinates from
depth or soak time would create a misleading spatial example, so the spatial
backends use small data sets that genuinely contain locations.

### sdmTMB

This example uses the Pacific cod data and mesh supplied by `sdmTMB`
[@Anderson2025]. Fixed, spatial, and spatiotemporal contributions are returned
in the same schema as the lobster examples.

```{r sdmtmb-example, eval=has_sdmTMB}
data("pcod_2011", package = "sdmTMB")
data("pcod_mesh_2011", package = "sdmTMB")

spatial_model <- sdmTMB::sdmTMB(
  present ~ as.factor(year) + depth_scaled,
  data = pcod_2011,
  mesh = pcod_mesh_2011,
  family = binomial(),
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  silent = TRUE
)

sdmTMB_diagnostic <- influ(spatial_model, focus = "year")
summary(sdmTMB_diagnostic)
```

```{r sdmtmb-influence, eval=has_sdmTMB, fig.cap="sdmTMB fixed, spatial, and spatiotemporal influence components."}
plot(sdmTMB_diagnostic, type = "components")
```

### TinyVAST

The TinyVAST example uses a small reproducible spatial count data set. The
purpose is to exercise the spatial and spatiotemporal component interface, not
to represent a complete survey analysis [@Thorson2025].

```{r tinyvast-example, eval=has_tinyVAST}
set.seed(2)
n <- 120
spatial_counts <- data.frame(
  x = runif(n),
  ycoord = runif(n),
  time = rep(1:4, each = n / 4)
)
spatial_counts$year <- factor(spatial_counts$time)
spatial_counts$var <- "catch"
spatial_counts$dist <- "poisson"
eta <- 0.3 + 0.08 * spatial_counts$time +
  0.3 * sin(2 * pi * spatial_counts$x)
spatial_counts$catch <- rpois(n, exp(eta))

spatial_mesh <- fmesher::fm_mesh_2d(
  spatial_counts[c("x", "ycoord")],
  n = 25
)
tiny_model <- tinyVAST::tinyVAST(
  catch ~ year,
  data = spatial_counts,
  family = list(poisson = poisson()),
  spatial_domain = spatial_mesh,
  spacetime_term = "",
  space_columns = c("x", "ycoord")
)

tinyVAST_diagnostic <- influ(tiny_model, focus = "year")
summary(tinyVAST_diagnostic)
```

```{r tinyvast-influence, eval=has_tinyVAST, fig.cap="TinyVAST fixed and spatiotemporal influence components."}
plot(tinyVAST_diagnostic, type = "components")
```

The current spatial adapters propagate fixed-effect covariance. Fitted latent
fields are reduced immediately to focus-level influence from their conditional
modes. Sparse joint-precision uncertainty for those fields is kept explicit as
a development boundary rather than represented by misleading marginal bands.

## One interface and compact uncertainty

All fitted-model methods return an `influ_diag`:

```{r methods}
methods("influ")
```

The result contains compact tables for influence, coefficient or field
summaries, data composition, and index comparisons:

```{r extraction}
head(influ_effects(glm_diagnostic))
head(influ_composition(glm_diagnostic))
influ_indices(glm_diagnostic)
```

Calculation and retention are separate choices:

```r
# Calculate uncertainty, retaining summaries only.
d1 <- influ(fitted_model, focus = "year", retain = "summary")

# Retain exact draws only for the derived diagnostics.
d2 <- influ(fitted_model, focus = "year", retain = "derived_draws")

# Write compact diagnostic draws to disk.
d3 <- influ(
  fitted_model,
  focus = "year",
  retain = "disk",
  draws_path = "influence-draws.rds"
)

# Fast preview from coefficient estimates or posterior means.
d4 <- influ(fitted_model, focus = "year", uncertainty = "none")
```

For GLMs and GAMs, linear diagnostic contrasts use fitted joint covariance
analytically. Joint coefficient simulation is used when derived draws are
requested. The BRMS backend reduces joint posterior draws during calculation,
and `glmmTMB` jointly simulates fixed components for hurdle and zero-inflated
means. This separation draws on the posterior-processing approach used in
CPUETools [@CPUETools], while the S3 design and GAM implementation were also
informed by `gamInflu` [@Dunn2025].

## Families and response structures

The initial family set is:

```{r families}
influ_families()
```

It covers Gaussian, binomial, Poisson, negative binomial, lognormal, Gamma, and
Tweedie responses. Hurdle/delta combinations cover Gamma, lognormal, Poisson,
and negative-binomial positive components. Zero-inflated combinations cover
Poisson and negative-binomial counts. Quasi families and more specialised
positive distributions are deliberately outside the initial scope.

The separate [hurdle and zero-inflation vignette](hurdle.html) explains named
components and unconditional means. The
[Bentley validation vignette](bentley-validation.html) runs the frozen original
`proto` implementation, compares values, and displays old and new plots side by
side.

## References
