---
title: "Influence diagnostics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Influence diagnostics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
has_mgcv <- requireNamespace("mgcv", quietly = TRUE)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
has_brms <- requireNamespace("brms", quietly = TRUE) &&
  requireNamespace("posterior", quietly = TRUE)
has_sdmTMB <- requireNamespace("sdmTMB", quietly = TRUE)
has_tinyVAST <- requireNamespace("tinyVAST", quietly = TRUE) &&
  requireNamespace("fmesher", quietly = TRUE)
suppressPackageStartupMessages(library(ggplot2))
theme_set(theme_bw())
```

## Why influence diagnostics?

CPUE standardisation attempts to distinguish changes in abundance from changes
in fishing practice, location, season, vessel composition, and other
explanatory variables. A fitted model may estimate these relationships well
without making it obvious why the standardised index differs from the nominal
series.

The coefficient-distribution-influence (CDI) framework combines the fitted
effect of a term with changes in the sampled distribution of that term through
the focus variable, which is usually year [@Bentley2012]. Spatial and
spatiotemporal versions can similarly show whether changes in sampled location
move observations among areas of different predicted abundance [@Hsu2022].

`influ2` represents these results with one model-neutral S3 class. Each backend
extracts model components and joint uncertainty, while the calculation,
storage, summary, and plotting layers remain shared.

## A realistic lobster CPUE example

The main examples use the simulated `lobsters_per_pot` data supplied with the
package. They span 2000 to 2017, with deliberate gaps and several thousand pot
records. The response is the number of lobsters caught per pot, and the
explanatory variables are month, depth in metres, and soak time in hours.
Annual sample sizes, seasonal
coverage, depths, and soak times all vary, giving the diagnostic something
meaningful to detect. Every value is generated by the package's deterministic
simulator; no commercial fishing records are included.

```{r lobster-data}
library(influ2)
data(lobsters_per_pot)

dim(lobsters_per_pot)
head(lobsters_per_pot)
```

The changing distribution of records by year and month is visible before a
model is fitted. Larger bubbles represent more pot records. With no `fill`
argument, `plot_bubble()` uses its default purple palette.

```{r lobster-sampling-purple, fig.height=6, fig.cap="Sampling effort by year and month, using the default purple bubble style."}
plot_bubble(
  df = lobsters_per_pot,
  group = c("year", "month")
) +
  labs(x = "Month", y = "Year")
```

Mapping month to colour gives a second view of the same sampling pattern.

```{r lobster-sampling-colour, fig.height=6, fig.cap="Sampling effort by year and month, with a rainbow palette distinguishing months."}
plot_bubble(
  df = lobsters_per_pot,
  group = c("year", "month"),
  fill = "month"
) +
  labs(x = "Month", y = "Year") +
  theme(legend.position = "none")
```

Depth and soak-time coverage also change through time.

```{r lobster-covariates, fig.cap="Observed depth and soak-time distributions by year."}
covariates <- tidyr::pivot_longer(
  lobsters_per_pot,
  cols = c("depth", "soak"),
  names_to = "covariate",
  values_to = "value"
)

ggplot(
  covariates,
  aes(x = year, y = value, group = year)
) +
  geom_boxplot(outlier.alpha = 0.08, linewidth = 0.25) +
  facet_wrap(~covariate, scales = "free_y", ncol = 1) +
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(x = "Year", y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## GLM

A Poisson GLM provides a transparent starting point. This is an example model,
not a recommendation that Poisson variation is adequate for these data.

```{r lobster-glm}
lobster_glm <- glm(
  lobsters ~ year + month + poly(depth, 3) + poly(soak, 3),
  family = poisson(link = "log"),
  data = lobsters_per_pot
)

glm_diagnostic <- influ(lobster_glm, focus = "year")
glm_diagnostic
summary(glm_diagnostic)
```

The influence plot shows how changes in sampled month, depth, and soak time
move the annual standardised series. The year term is retained in the object,
but omitted from this plot because its within-year composition is the focus,
not a sampling-distribution effect.

```{r lobster-glm-influence, fig.cap="GLM influence ratios for the changing lobster covariate distributions."}
plot(
  glm_diagnostic,
  type = "influence",
  term = c("month", "poly(depth, 3)", "poly(soak, 3)")
)
```

The same object contains nominal and standardised indices. Here, **nominal**
means the observed annual arithmetic mean of lobsters per pot, including zero
catches, without adjusting for month, depth, or soak time. Each pot record has
equal weight in this example; supplying `weights` to `influ()` instead gives a
weighted annual mean. The standardised index is the exponentiated, centred
year effect from the fitted log-link model. It is a relative index, not an
estimate in lobsters per pot, so the two series appear in separate panels
labelled `response` and `ratio`, respectively.

```{r lobster-glm-index, fig.cap="Observed annual mean lobsters per pot (nominal, response scale) and the relative Poisson-GLM-standardised index (ratio scale)."}
plot(glm_diagnostic, type = "index")
```

The model-neutral overall and trend metrics reproduce Bentley's definitions.
The overall metric is the mean absolute link-scale influence transformed back
to a proportional scale. The trend metric is the fitted change per ordered
focus level, also transformed to a proportional scale.

```{r lobster-glm-metrics}
subset(
  influ_metrics(glm_diagnostic),
  term != "year",
  select = c(term, metric, estimate)
)
```

### Reading a CDI plot

A CDI plot aligns the fitted term effect above its observed distribution, with
the resulting annual influence beside it. By default, the top panel centres
the term on the same weighted reference distribution as the influence
calculation. For this log-link model, it shows relative month effects on a
logarithmic axis, with one as the reference. A value of 1.2 represents a 20%
higher monthly contribution to expected catch than the reference, holding the
other model components fixed. No month is singled out merely because it is
the model's first factor level.

The default reference uses the observations and any supplied `weights`.
Supplying `reference_data` and, optionally, `reference_weights` changes both
the influence reference and the CDI centring to that explicit distribution.
The bubbles continue to show the observed sampling distribution.

```{r lobster-cdi-options, eval=FALSE}
# Centred relative month effects, with 95% confidence intervals.
plot(glm_diagnostic, type = "cdi", term = "month")

# The same centred effects in additive log units.
plot(
  glm_diagnostic, type = "cdi", term = "month",
  coefficient_scale = "link"
)

# Original model coding: log effects relative to month 1 for this GLM.
plot(
  glm_diagnostic, type = "cdi", term = "month",
  coefficient_reference = "model"
)
```

The interval calculation includes uncertainty in the estimated centre and
its covariance with each term effect. The default interval covers 95%; set
`probs` when calling `influ()` to choose other bounds. The model-reference
option changes only the top panel, leaving the composition and influence
panels unchanged.

The display scale follows the component's link. Log-link and lognormal
components use relative effects; identity-link components use additive
effects. Logit, probit, and complementary-log-log components remain in their
labelled link units, centred on zero. Their top panels are not catch
multipliers or changes in encounter probability.

### A common prediction-grid reference

Observed-data standardisation is the default. For comparisons among fleets,
areas, or models, an explicit prediction grid prevents differences in their
observed samples from silently changing the reference distribution. This grid
crosses every year and month at common depth and soak values.

```{r lobster-reference-grid}
lobster_reference <- expand.grid(
  year = levels(lobsters_per_pot$year),
  month = levels(lobsters_per_pot$month),
  depth = median(lobsters_per_pot$depth),
  soak = median(lobsters_per_pot$soak)
)
lobster_reference$year <- factor(
  lobster_reference$year,
  levels = levels(lobsters_per_pot$year)
)
lobster_reference$month <- factor(
  lobster_reference$month,
  levels = levels(lobsters_per_pot$month)
)

glm_grid_diagnostic <- influ(
  lobster_glm,
  focus = "year",
  reference_data = lobster_reference
)
glm_grid_diagnostic$metadata[c("reference", "n_reference")]
```

Reference weights can be supplied with `reference_weights`. They may be a
numeric vector or the name of a column in `reference_data`. This makes the
standardisation estimand explicit, reproducible, and independent of accidental
sample imbalance.

## GAM

The `mgcv` backend works with parametric terms and smooths [@Wood2017]. Here,
negative-binomial variation is combined with smooth depth and soak effects.

```{r lobster-gam, eval=has_mgcv}
lobster_gam <- mgcv::gam(
  lobsters ~ year + month + s(depth, k = 5) + s(soak, k = 5),
  family = mgcv::nb(),
  method = "REML",
  data = lobsters_per_pot
)

gam_diagnostic <- influ(lobster_gam, focus = "year")
summary(gam_diagnostic)
```

```{r lobster-gam-influence, eval=has_mgcv, fig.cap="GAM influence ratios for month, depth, and soak time."}
plot(
  gam_diagnostic,
  type = "influence",
  term = c("month", "s(depth)", "s(soak)")
)
```

This is the same `influ_diag` interface as the GLM. No plotting code needs to
know that two of the terms are smooths.

## glmmTMB

`glmmTMB` adds mixed effects, negative-binomial models, hurdle models, and
zero-inflation [@Brooks2017]. The example treats month as a random intercept.

```{r lobster-glmmtmb, eval=has_glmmTMB}
lobster_glmmTMB <- glmmTMB::glmmTMB(
  lobsters ~ year + poly(depth, 2) + poly(soak, 2) + (1 | month),
  family = glmmTMB::nbinom2(),
  data = lobsters_per_pot
)

glmmTMB_diagnostic <- influ(lobster_glmmTMB, focus = "year")
summary(glmmTMB_diagnostic)
```

```{r lobster-glmmtmb-influence, eval=has_glmmTMB, fig.cap="glmmTMB fixed- and random-effect influence ratios."}
plot(glmmTMB_diagnostic, type = "components")
```

Fixed effects use the joint maximum-likelihood covariance. The month random
effect uses its fitted conditional modes, with uncertainty propagated from
their joint conditional latent covariance and labelled separately.
For hurdle and zero-inflated fits, `influ()` also calculates population-level
unconditional-mean influence using the joint covariance of both components.

## brms

The Bayesian backend uses genuine joint posterior draws, but projects each draw
directly to the much smaller focus-by-term diagnostic. It never constructs an
observation-by-draw-by-term array [@Buerkner2017]. A model for the same lobster
data can be fitted as follows:

```{r lobster-brms-fit, eval=FALSE}
lobster_brms <- brms::brm(
  lobsters ~ year + (1 | month) + s(depth, k = 3),
  family = brms::negbinomial(),
  data = lobsters_per_pot,
  chains = 2,
  iter = 2000,
  seed = 1,
  file = "fit2",
  file_refit = "on_change"
)
```

The vignette uses the fitted model shipped with its source, rather than running
MCMC whenever the documentation is built.

```{r lobster-brms, eval=has_brms}
lobster_brms <- readRDS(system.file(
  "extdata", "brms-fixtures", "fit2.rds", package = "influ2"
))
brms_diagnostic <- influ(
  lobster_brms,
  focus = "year",
  ndraws = 250,
  retain = "summary"
)
summary(brms_diagnostic)
```

```{r lobster-brms-influence, eval=has_brms, fig.cap="BRMS population-level, month group-level, and depth-smooth influence ratios."}
plot(brms_diagnostic, type = "components")
```

Population-level, group-level, and smooth contributions all preserve posterior
dependence during calculation. The default object retains interval summaries;
`retain = "derived_draws"` retains only the compact diagnostic draws.

The same posterior diagnostic can be displayed as a complete Bayesian CDI
plot. The top panel shows relative month effects centred on the observed
monthly distribution; the lower panels align that distribution with the
resulting yearly influence. Each joint posterior draw is centred before it
is exponentiated. Points are posterior means of those relative effects, and
bars are 95% credible intervals. This preserves dependence between each month
and the estimated reference, without retaining the full posterior array in
the diagnostic. The first month therefore has an estimated effect and an
interval, just like every other month.

```{r lobster-brms-cdi, eval=has_brms, fig.width=10, fig.height=6, fig.cap="Bayesian CDI for the monthly group-level effect: centred relative month effects with 95% credible intervals (top left), observed monthly composition (bottom left), and annual influence (bottom right)."}
plot(
  brms_diagnostic,
  type = "cdi",
  term = "month",
  component = "conditional:group_level"
)
```

## sdmTMB

The lobster data deliberately have no coordinates. Inventing coordinates from
depth or soak time would create a misleading spatial example, so this example
uses the Pacific cod data and mesh supplied by `sdmTMB` [@Anderson2025]. Fixed,
spatial, and spatiotemporal contributions are returned in the same schema as
the lobster examples.

```{r sdmtmb-example, eval=has_sdmTMB}
data("pcod_2011", package = "sdmTMB")
data("pcod_mesh_2011", package = "sdmTMB")

spatial_model <- sdmTMB::sdmTMB(
  present ~ as.factor(year) + depth_scaled,
  data = pcod_2011,
  mesh = pcod_mesh_2011,
  family = binomial(),
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  silent = TRUE
)

sdmTMB_diagnostic <- influ(
  spatial_model,
  focus = "year",
  ndraws = 100,
  seed = 1
)
summary(sdmTMB_diagnostic)
```

```{r sdmtmb-influence, eval=has_sdmTMB, fig.cap="sdmTMB fixed, spatial, and spatiotemporal influence components."}
plot(sdmTMB_diagnostic, type = "components")
```

## tinyVAST

The `tinyVAST` example uses a small reproducible spatial count data set. The
purpose is to exercise the spatial and spatiotemporal component interface, not
to represent a complete survey analysis [@Thorson2025].

```{r tinyvast-example, eval=has_tinyVAST}
set.seed(2)
n <- 120
spatial_counts <- data.frame(
  x = runif(n),
  ycoord = runif(n),
  time = rep(1:4, each = n / 4)
)
spatial_counts$year <- factor(spatial_counts$time)
spatial_counts$var <- "catch"
spatial_counts$dist <- "poisson"
eta <- 0.3 + 0.08 * spatial_counts$time +
  0.3 * sin(2 * pi * spatial_counts$x)
spatial_counts$catch <- rpois(n, exp(eta))

spatial_mesh <- fmesher::fm_mesh_2d(
  spatial_counts[c("x", "ycoord")],
  n = 25
)
tiny_model <- tinyVAST::tinyVAST(
  catch ~ year,
  data = spatial_counts,
  family = list(poisson = poisson()),
  spatial_domain = spatial_mesh,
  spacetime_term = "",
  space_columns = c("x", "ycoord")
)

tinyVAST_diagnostic <- influ(
  tiny_model,
  focus = "year",
  ndraws = 100,
  seed = 1
)
summary(tinyVAST_diagnostic)
```

```{r tinyvast-influence, eval=has_tinyVAST, fig.cap="tinyVAST fixed and spatial influence components."}
plot(tinyVAST_diagnostic, type = "components")
```

The spatial adapters propagate fixed-effect covariance and simulate spatial
and spatiotemporal fields from each model's sparse joint precision matrix. Each
draw is reduced immediately to focus-level influence, so the returned object
does not contain an observations-by-draws field array. Delta fields use the
same joint draw for occurrence and positive components before the
unconditional mean is calculated.

## Comparing standardised indices

`plot_compare()` accepts a list of fitted models from different backends, or
their already-calculated `influ_diag` objects. Reusing the diagnostics below
avoids repeating model fitting, posterior sampling, or influence calculations.

These four lobster models use the same response, pot records, and years.
They differ in both their model structure and their distribution: the GLM is
Poisson, the other models are negative binomial, and their covariate and
month-effect specifications differ. This is therefore a sensitivity comparison
between the fitted models, not a controlled comparison of fitting software.

```{r lobster-comparison-models}
lobster_diagnostics <- list(GLM = glm_diagnostic)
if (has_mgcv) lobster_diagnostics$GAM <- gam_diagnostic
if (has_glmmTMB) lobster_diagnostics$glmmTMB <- glmmTMB_diagnostic
if (has_brms) lobster_diagnostics$BRMS <- brms_diagnostic
```

`rescale = 1` gives each series a geometric mean of one over the common
2000--2017 period, making their relative trajectories directly comparable.
Uncertainty ribbons are omitted here to keep overlapping lines readable.

```{r lobster-model-comparison, fig.cap="Relative standardised lobster CPUE indices from the available GLM, GAM, glmmTMB, and BRMS examples. Each series has a geometric mean of one over 2000--2017; differences reflect the models' distributions and structures as well as their estimation methods.", fig.alt="Annual standardised lobster CPUE indices on a common relative scale, with separate labelled lines for each fitted model."}
plot_compare(
  lobster_diagnostics,
  labels = names(lobster_diagnostics),
  rescale = 1,
  show_probs = FALSE
) +
  labs(x = "Year", y = "Relative standardised CPUE", colour = "Model") +
  theme(legend.position = "bottom")
```

Set `show_probs = TRUE` to display the intervals already stored in the
diagnostics. For these objects, they are 95% confidence or credible intervals,
depending on the backend. Rescaling multiplies the estimates and interval
bounds by the same display constant; it does not recalculate uncertainty in
the estimated normalising constant or provide a test of differences between
models.

The same function can include `sdmTMB` and `tinyVAST` models when their indices
represent the same response, population, and time period on a compatible
scale. Their demonstrations above use different data, so they are deliberately
excluded here. For other comparisons, also check the reference distribution,
component (for example, unconditional mean versus positive catch), and shared
years before overlaying indices. A common plotting interface does not by
itself make those quantities comparable.

## Refitted step plots

A refitted step plot asks how the estimated year effect changes as terms are
added to a model. Each changed specification is fitted afresh, re-estimating
all its coefficients. The result therefore depends on the order in which terms are
added: it describes that sequence of models, rather than allocating an
order-independent share of the final result to each term.

For a simple GLM, `influ_steps()` can construct the sequence from the fitted
formula. It starts with the year term, preserves any offsets, and then adds
the remaining terms in formula order. This lobster example starts with year,
then adds month, the depth polynomial, and the soak-time polynomial. All
steps use the same pot records and observed reference distribution.

```{r lobster-refitted-steps}
lobster_steps <- influ_steps(
  lobster_glm,
  year = "year",
  refit = TRUE
)
lobster_steps
```

A step that exactly matches the original model or an earlier step reuses that
fit. The printed status records whether each model was refitted or reused;
this avoids repeating an identical fit while preserving the model comparison.

```{r lobster-refitted-step-plot, fig.height=8, fig.cap="Changes in the relative year effect as month, depth, and soak-time terms are added in a lobster GLM refitting sequence. Each panel highlights the current model with its 95% confidence interval and retains the preceding models for comparison.", fig.alt="Four sequential GLM panels showing centred year-effect ratios for year only, year plus month, then depth, then soak time."}
plot_step(lobster_steps)
```

These are centred year-effect contrasts. For the lobster log-link models,
they are exponentiated year effects relative to a common reference, not
area-weighted abundance indices or spatially integrated predictions.
The shaded interval belongs to the current fitted model. It is not an
interval for the difference between that model and the preceding one, because
the models were fitted to the same observations and their estimates are
dependent.

Once calculated, `plot(lobster_steps)` or `plot_step(lobster_steps)` reuses the
stored summaries without fitting again. The shortcut below creates the same
kind of plot directly from a fitted model, but it performs the refits each
time it is called. The automatic main-formula sequence also supports simple
GAMs and `glmmTMB` models.

```{r lobster-refitted-step-shortcuts, eval=FALSE}
plot_step(lobster_glm, year = "year", refit = TRUE)

gam_steps <- influ_steps(lobster_gam, year = "year", refit = TRUE)
plot(gam_steps)
```

Use an explicit named `steps` list when the order or model structure needs
closer control. Each formula is fitted against the original model's settings,
so the sequence is visible in the code. A common `reference_data` grid and
`reference_weights` may be passed to `influ_steps()` when the observed
reference is not the intended comparison.

```{r lobster-explicit-steps, eval=FALSE}
lobster_steps <- influ_steps(
  lobster_glm,
  year = "year",
  refit = TRUE,
  steps = list(
    "Year" = ~year,
    "Year + depth" = ~year + poly(depth, 3),
    "Year + depth + month" = ~year + poly(depth, 3) + month,
    "Year + depth + month + soak" =
      ~year + poly(depth, 3) + month + poly(soak, 3)
  ),
  reference_data = lobster_reference
)
```

For expensive models, particularly BRMS fits, supply an ordered list of
models that have already been fitted. This calculates and plots their year
contrasts without running MCMC again. The models should use the same response,
observations, focus levels, and reference distribution. Select the intended
`component` explicitly for models with several response components. Check the
convergence of supplied fits before including them; their `supplied` status
does not record a new convergence assessment.

```{r brms-existing-step-models, eval=FALSE}
brms_steps <- influ_steps(
  list(
    "Year" = brms_year_fit,
    "Year + month" = brms_month_fit,
    "Year + month + depth" = brms_depth_fit
  ),
  year = "year",
  ndraws = 250
)
plot(brms_steps)
```

Intervals default to 95% when step diagnostics are calculated; set `probs`
to change their coverage. The default `keep_fits = FALSE` keeps the result
compact. Set `keep_fits = TRUE` only when the fitted intermediate models are
needed afterwards, because retaining those models can substantially increase
the object's size. The [spatial vignette](spatial-spatiotemporal.html) shows
explicit sequences that add spatial and spatiotemporal structure while
continuing to compare the fitted year effects.

## One interface and compact uncertainty

All fitted-model methods return an `influ_diag`:

```{r methods}
methods("influ")
```

The result contains compact tables for influence, coefficient or field
summaries, data composition, and index comparisons:

```{r extraction}
head(influ_effects(glm_diagnostic))
head(influ_composition(glm_diagnostic))
influ_indices(glm_diagnostic)
influ_metrics(glm_diagnostic)
```

Calculation and retention are separate choices:

```r
# Calculate uncertainty, retaining summaries only.
d1 <- influ(fitted_model, focus = "year", retain = "summary")

# Retain exact draws only for the derived diagnostics.
d2 <- influ(fitted_model, focus = "year", retain = "derived_draws")

# Write compact diagnostic draws to disk.
d3 <- influ(
  fitted_model,
  focus = "year",
  retain = "disk",
  draws_path = "influence-draws.rds"
)

# Fast preview from coefficient estimates or posterior means.
d4 <- influ(fitted_model, focus = "year", uncertainty = "none")
```

For GLMs and GAMs, linear diagnostic contrasts use fitted joint covariance
analytically. Joint coefficient simulation is used when derived draws are
requested. The BRMS backend reduces joint posterior draws during calculation,
`glmmTMB` jointly simulates fixed components for hurdle and zero-inflated
means, and the spatial backends reduce sparse joint-precision draws to the same
compact schema. This separation draws on the posterior-processing approach used in
CPUETools [@CPUETools], while the S3 design and GAM implementation were also
informed by `gamInflu` [@Dunn2025].

## Families and response structures

The initial family set is:

```{r families}
influ_families()
```

It covers Gaussian, binomial, Poisson, negative binomial, lognormal, Gamma, and
Tweedie responses. Hurdle/delta combinations cover Gamma, lognormal, Poisson,
and negative-binomial positive components. Zero-inflated combinations cover
Poisson and negative-binomial counts. Quasi families and more specialised
positive distributions are deliberately outside the initial scope.

The separate [hurdle and zero-inflation vignette](hurdle.html) explains named
components and unconditional means. The
[Bentley validation vignette](bentley-validation.html) runs the frozen original
`proto` implementation, compares values, and displays old and new plots side by
side. The [spatial and spatiotemporal vignette](spatial-spatiotemporal.html)
maps persistent and time-varying fields alongside their influence components.

## References
