---
title: "Model-neutral influence diagnostics with influ2"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Model-neutral influence diagnostics with influ2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

## Why influence diagnostics?

CPUE standardisation attempts to distinguish changes in abundance from changes
in fishing practice, location, season, vessel composition, and other
explanatory variables. A fitted model may estimate these relationships well
without making it obvious why the standardised index differs from the nominal
series.

The coefficient-distribution-influence (CDI) framework combines the fitted
effect of a term with changes in the sampled distribution of that term through
the focus variable, which is usually year [@Bentley2012]. Spatial and
spatiotemporal versions can similarly show whether changes in sampled location
move observations among areas of different predicted abundance [@Hsu2022].

`influ2` represents these results with one model-neutral S3 class. The model
backend is responsible for extracting model components and joint uncertainty;
the calculation, storage, summary, and plotting layers are shared.

## A common interface

The public entry point is `influ()`:

```{r basic-example}
library(influ2)

fixture_file <- system.file(
  "extdata", "bentley-poisson-data.csv", package = "influ2"
)
dat <- read.csv(
  fixture_file,
  colClasses = c(year = "factor", area = "factor", vessel = "factor")
)

fit <- glm(
  catch ~ year + area + vessel,
  family = poisson(link = "log"),
  data = dat
)

diagnostic <- influ(fit, focus = "year")
diagnostic
summary(diagnostic)
```

The implemented methods share this contract:

```{r methods}
methods("influ")
```

The result contains compact tables for influence, coefficient or field
summaries, data composition, and index comparisons:

```{r extraction}
head(influ_effects(diagnostic))
head(influ_composition(diagnostic))
influ_indices(diagnostic)
```

## Scales and components

Every term has an additive `link`-scale influence. A second representation is
provided on a natural interpretation scale:

- a ratio centred on one for log-linked models and lognormal location models;
- a difference centred on zero for identity, logit, probit, and
  complementary log-log models.

Hurdle and zero-inflated models retain named components. Occurrence,
positive/count, spatial, and spatiotemporal effects therefore remain visible
instead of being silently collapsed. Combined unconditional-mean influence is
a separate estimand because it must preserve dependence between model
components.

The initial family set is returned by:

```{r families}
influ_families()
```

Hurdle/delta combinations initially cover Gamma, lognormal, Poisson, and
negative binomial positive components. Zero-inflated combinations initially
cover Poisson and negative binomial count components. Quasi families and more
specialised positive distributions are outside the first implementation.

## Low-memory uncertainty

Calculation and retention are separate choices. For a Bayesian fit, `influ2`
projects joint posterior coefficient draws directly into the much smaller
focus-by-term diagnostic contrasts. It does not first create an
observation-by-draw-by-term array.

```r
# Default: calculate with joint posterior draws, retain summaries only.
d1 <- influ(brms_fit, focus = "year", retain = "summary")

# Keep exact draws only for the derived diagnostics.
d2 <- influ(brms_fit, focus = "year", retain = "derived_draws")

# Fast preview: reduce posterior coefficients to their means first.
d3 <- influ(brms_fit, focus = "year", uncertainty = "none")
```

For GLMs and GAMs, linear diagnostic contrasts use the fitted joint covariance
analytically. Joint coefficient simulation is available when derived draws are
needed. `sdmTMB` and TinyVAST fixed effects use their TMB marginal covariance;
their fitted spatial and spatiotemporal modes are already reduced without
dense arrays. Sparse joint-precision simulation for full field uncertainty is
the next spatial uncertainty layer.

This division draws on the posterior-processing approach used in CPUETools
[@CPUETools] while retaining the full joint draw during each diagnostic
calculation. The S3 object and GAM implementation were also informed by
`gamInflu` [@Dunn2025].

## Standard plots

The same plot method works regardless of the original model backend:

```{r influence-plot}
plot(diagnostic, type = "influence")
```

```{r index-plot}
plot(diagnostic, type = "index")
```

```{r cdi-plot, fig.width=9, fig.height=5}
plot(diagnostic, type = "cdi", term = "area")
```

`autoplot()` delegates to the same implementation. Plot data can always be
obtained first through the extractor functions when a project needs a custom
graphic.

## Legacy Bentley agreement

The original `proto` implementation is frozen under `tools/legacy` and is no
longer loaded into the package namespace. Routine tests compare the new engine
against a compact fixture generated from that implementation.

```{r bentley-parity}
reference <- read.csv(system.file(
  "extdata", "bentley-poisson-reference.csv", package = "influ2"
))
current <- subset(
  influ_effects(diagnostic),
  term %in% c("area", "vessel") & scale == "link",
  select = c(focus, level, term, estimate)
)
names(current)[names(current) == "estimate"] <- "link_influence"

comparison <- merge(reference, current,
  by = c("focus", "level", "term"),
  suffixes = c("_legacy", "_new")
)
comparison$absolute_difference <- abs(
  comparison$link_influence_legacy - comparison$link_influence_new
)
comparison
stopifnot(max(comparison$absolute_difference) < 1e-8)
```

The fixture can be deliberately regenerated with
`tools/legacy/regenerate-bentley-fixture.R`. It is not regenerated during
normal tests, so a change in the new implementation cannot silently move both
sides of the comparison.

## Model backends

The package architecture targets GLMs, `mgcv` GAMs [@Wood2017], Bayesian
multilevel models fitted with `brms` [@Buerkner2017], mixed models fitted with
`glmmTMB` [@Brooks2017], spatial and spatiotemporal models fitted with `sdmTMB`
[@Anderson2025], and multivariate spatial models fitted with TinyVAST
[@Thorson2025]. The backends may expose different model components, but all of
them return an `influ_diag`.

Current staged limitations are recorded in each diagnostic object's
`metadata$notes`. In particular, BRMS group-level and smooth-term projections,
sparse field uncertainty, mixed-family multivariate TinyVAST models, and exact
combined unconditional-mean influence remain active development areas rather
than being returned with misleading precision.

## References
