---
title: "Regional CPUE indices"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Regional CPUE indices}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
  fig.height = 5, message = FALSE, warning = FALSE)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
library(influ2)
library(ggplot2)
theme_set(theme_bw())
```

## Two regions, one fitted model

Suppose four statistical areas are to supply two assessment series:
**Region A contains areas 1 and 2**, and **Region B contains areas 3 and 4**.
We fit one model to all observations, then average its response predictions
over a fixed reference population within each region. There is no regional
refitting, and no new regional extraction function is needed.

The workflow is `cpue_index()` for calculation, `index_table()` for reporting,
`index_vcov()` for annual covariance, and `plot_compare()` for the two curves.
See [CPUE indices](cpue-indices.html) for the general API and uncertainty
definitions. Here the target is **standardised expected CPUE**, not a year
coefficient, a sum of coefficients, or area-integrated biomass.

## Simulated observations and known truth

This separate simulation leaves the main lobster example unchanged. Each
row is one pot deployment, so the integer response is already lobsters per
pot. The modelled response could instead be another CPUE measure, with a
suitable observation distribution. If modelling counts with variable effort,
include that exposure in the model and predict at one unit of exposure.

There are eight years, four areas, vessel effects, and changing depth, soak
time, and sampling allocations. Every area/year combination is observed,
but some have substantially fewer observations than others. Areas have
different underlying trends, so the two regional indices have something
meaningful to reveal.

```{r regional-simulation}
set.seed(1809)
years <- as.character(2010:2017)
areas <- as.character(1:4)
area_region <- data.frame(area = areas,
  region = c("Region A", "Region A", "Region B", "Region B"))

# Expected CPUE at a zero vessel effect; also used for the reference truth.
expected_cpue <- function(d) {
  t <- (as.numeric(as.character(d$year)) - 2010) / 7
  a <- match(as.character(d$area), areas)
  exp(log(3) + c(-0.25, 0.3, -0.1, 0.35)[a] +
    c(-0.9, -0.3, 0.25, 1.0)[a] * t + 0.12 * sin(2 * pi * t) +
    0.018 * (d$depth - 50) + 0.025 * (d$soak - 24))
}

cells <- expand.grid(year = years, area = areas, stringsAsFactors = FALSE)
t <- (as.numeric(cells$year) - 2010) / 7
cells$n <- round(45 + 100 * ifelse(cells$area %in% c("1", "3"), 1 - t, t))
regional_data <- cells[rep(seq_len(nrow(cells)), cells$n), c("year", "area")]
rownames(regional_data) <- NULL
n <- nrow(regional_data)
t <- (as.numeric(regional_data$year) - 2010) / 7
regional_data$year <- factor(regional_data$year, levels = years)
regional_data$area <- factor(regional_data$area, levels = areas)
regional_data$region <- area_region$region[match(regional_data$area, areas)]
regional_data$depth <- runif(n, 30, 65) + 10 * t
regional_data$soak <- runif(n, 12, 30) + 8 * t
regional_data$vessel <- factor(sample(sprintf("V%02d", 1:18), n, replace = TRUE))
vessel_effect <- rnorm(18, sd = 0.25)
mu <- expected_cpue(regional_data) * exp(vessel_effect[regional_data$vessel])
regional_data$cpue <- rnbinom(n, mu = mu, size = 8)
```

```{r regional-sampling, fig.cap="Simulated sampling allocations by statistical area and year. Region A comprises areas 1 and 2, and Region B comprises areas 3 and 4. Sampling shifts towards areas 2 and 4 over time, although every area/year cell remains observed. These changing sample counts are not the reference weights used for standardisation."}
cells$region <- area_region$region[match(cells$area, areas)]
ggplot(cells, aes(as.numeric(year), n, colour = area)) +
  geom_line() + geom_point() + facet_wrap(~ region) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Year", y = "Pot deployments", colour = "Area")
```

## Allow different area trends

The `year * area` interaction supplies an area-specific effect in every year,
with depth and soak effects shared across all areas. The vessel random
intercept accounts for persistent vessel differences in fitting.

```{r regional-fit, eval=has_glmmTMB}
regional_fit <- glmmTMB::glmmTMB(
  cpue ~ year * area + depth + soak + (1 | vessel),
  family = glmmTMB::nbinom2(), data = regional_data
)
stopifnot(regional_fit$fit$convergence == 0, regional_fit$sdr$pdHess)
```

This example uses ML, and its index predictions set vessel effects to zero.
That means a zero-effect vessel on the log scale, **not** integration over
the vessel population. This follows the
[native glmmTMB prediction convention](https://glmmtmb.github.io/glmmTMB/reference/predict.glmmTMB.html)
for `re.form = NA`. The simulated truth below uses the same target.
Selective retention of random effects and REML index extraction are outside
this example and the current glmmTMB response-index adapter.

## Define a fixed regional reference

For each region, use its two areas equally, and four common depth/soak
profiles equally within each area. These profiles lie within the simulated
covariate support in every year. Their weights stay fixed across years:
annual changes in where or how fishing occurred cannot redefine the target.

For region \(r\) and year \(t\), the calculation is

\[
 I_{r,t} = \sum_{a\in r} w_{r,a}\sum_j v_j\,
 E(Y\mid t,a,\mathrm{profile}_j,b_{\mathrm{vessel}}=0),
 \qquad \sum_{a\in r} w_{r,a}=1,\quad \sum_j v_j=1.
\]

Average on the **response scale**, before any normalisation. Do not average
link predictions, already normalised area curves, SDs, or CVs.

```{r regional-reference}
profiles <- expand.grid(depth = c(45, 60), soak = c(21, 29))
reference <- merge(area_region, profiles, by = NULL)
reference$area <- factor(reference$area, levels = areas)
reference$weight <- 0.5 / nrow(profiles)
references <- split(reference, reference$region)
knitr::kable(reference, digits = 3, row.names = FALSE)
```

The reference table deliberately has no year column. Keep the full fitted
factor levels even when a regional table contains just two areas. The
`region` label is bookkeeping, not an additional predictor in this model.

```{r regional-indices, eval=has_glmmTMB}
calculate_regions <- function(fit, refs, rescale = "raw", uncertainty = "auto") {
  lapply(refs, function(ref) cpue_index(
    fit, year = "year", method = "standardised",
    reference_data = ref[c("area", "depth", "soak")],
    reference_weights = ref$weight, rescale = rescale,
    uncertainty = uncertainty, units = "lobsters per pot"
  ))
}
regional_indices <- calculate_regions(regional_fit, references)
regional_tables <- do.call(rbind, lapply(names(regional_indices), function(r) {
  data.frame(Region = r, index_table(regional_indices[[r]]), row.names = NULL)
}))
knitr::kable(regional_tables[c("Region", "Year", "Mean", "SD", "CV",
  "Qlower", "Qupper")], digits = 3, row.names = FALSE)
```

Both series come from the same `regional_fit`. Neither calculating another
reference nor drawing another plot refits it. The table's SD is uncertainty
in the estimated index, not the spread of individual responses.

```{r regional-index-plot, eval=has_glmmTMB, fig.cap="Two standardised regional CPUE series from one negative-binomial glmmTMB fit. Each region averages its two areas equally over the same four depth/soak profiles, with vessel effects set to zero. Ribbons are pointwise 95% delta-method confidence intervals, including dependence among the weighted reference predictions. These are expected-response indices, not area-integrated biomass."}
plot_compare(regional_indices) + labs(colour = "Region", fill = "Region")
```

## Compare with known truth and nominal CPUE

The known reference truth is obtained by applying exactly the same weights
and profiles to the simulation's expected response. Nominal CPUE is the
arithmetic observed mean within each region/year; its composition and vessel
mixture change with the observations. Consequently, nominal and standardised
values do not estimate the same reference population.

```{r regional-truth, eval=has_glmmTMB}
truth <- do.call(rbind, lapply(names(references), function(r) {
  ref <- references[[r]]
  data.frame(Region = r, Year = years, Truth = vapply(years, function(y) {
    d <- ref
    d$year <- y
    weighted.mean(expected_cpue(d), ref$weight)
  }, numeric(1)))
}))
nominal <- aggregate(cpue ~ year + region, data = regional_data, FUN = mean)
comparison <- merge(regional_tables, truth, by = c("Region", "Year"))
comparison <- merge(comparison, nominal, by.x = c("Region", "Year"),
  by.y = c("region", "year"))
```

```{r regional-truth-plot, eval=has_glmmTMB, fig.cap="Standardised regional CPUE (purple, with pointwise 95% intervals), known reference truth (dashed black), and nominal observed CPUE (orange). The standardised estimates and truth share fixed regional weights and a zero vessel effect. Nominal means retain changing sample composition and vessel effects. This single simulation illustrates the target and sampling corrections; it is not a confidence-interval coverage study."}
ggplot(comparison, aes(as.numeric(Year))) +
  geom_ribbon(aes(ymin = Qlower, ymax = Qupper), fill = "#792B8B", alpha = 0.15) +
  geom_line(aes(y = Mean, colour = "Standardised")) +
  geom_point(aes(y = Mean, colour = "Standardised")) +
  geom_line(aes(y = Truth, colour = "Reference truth"), linetype = 2) +
  geom_line(aes(y = cpue, colour = "Nominal")) +
  facet_wrap(~ Region) +
  scale_colour_manual(values = c(Standardised = "#792B8B",
    `Reference truth` = "black", Nominal = "#B65D13")) +
  labs(x = "Year", y = "Lobsters per pot", colour = NULL)
```

## Relative series and a weighting sensitivity

Use `rescale = 1` **during index calculation** to give each regional curve a
geometric mean of one over these same eight years. This propagates uncertainty
in the normalising denominator. It is different from simply dividing the
displayed limits by a point estimate. Normalisation occurs after aggregation.

```{r regional-relative, eval=has_glmmTMB, fig.cap="Relative regional indices, each normalised to geometric mean one over 2010–2017 after regional aggregation. Uncertainty includes the normalising denominator. The distinct trends are supported by year-by-area interactions in the common fitted model; the curves are not independent estimates."}
relative_indices <- calculate_regions(regional_fit, references, rescale = 1)
plot_compare(relative_indices) + labs(colour = "Region", fill = "Region")
```

Equal area weights are a choice, not a property of the model. For a
sensitivity, give the first area in each region weight 0.75 and the second
0.25, still fixed across years. These are declared standardisation weights,
not claims about habitat area. A change in the curve reflects a change in
the reference target, not another fitted model.

```{r regional-weight-sensitivity, eval=has_glmmTMB}
unequal_references <- lapply(references, function(ref) {
  ref$weight <- ifelse(ref$area %in% c("1", "3"), 0.75, 0.25) / nrow(profiles)
  ref
})
unequal_indices <- calculate_regions(regional_fit, unequal_references, rescale = 1)
```

```{r regional-weight-plot, eval=has_glmmTMB, fig.cap="Sensitivity of relative regional indices to fixed area weights: equal weights versus 75:25 weights favouring the first area in each region. The model, reference depth/soak profiles, and normalisation years are unchanged. Pointwise 95% intervals describe each stated target; overlap is not a test of the difference between weighting schemes."}
p_a <- plot_compare(list(`Equal weights` = relative_indices[["Region A"]],
  `75:25 weights` = unequal_indices[["Region A"]])) + labs(title = "Region A")
p_b <- plot_compare(list(`Equal weights` = relative_indices[["Region B"]],
  `75:25 weights` = unequal_indices[["Region B"]])) + labs(title = "Region B")
patchwork::wrap_plots(p_a, p_b, ncol = 2, guides = "collect")
```

## Check predictions and covariance independently

For a small grid we can request the native covariance of all response
predictions and average it explicitly. If \(W\) contains the reference
weights by year, the regional index covariance is
\(W\,\mathrm{Cov}(\widehat{\boldsymbol\mu})\,W^\mathsf{T}\).
Off-diagonal terms must be retained: area predictions share fitted parameters.
The native dense matrix here is just a small-grid verification, not a
recommended large-grid implementation. influ2 aggregates gradients in batches.

```{r regional-native-check, eval=has_glmmTMB}
native_checks <- lapply(names(references), function(r) {
  ref <- references[[r]]
  grid <- ref[rep(seq_len(nrow(ref)), times = length(years)), ]
  grid$year <- factor(rep(years, each = nrow(ref)), levels = years)
  grid$vessel <- factor(levels(regional_data$vessel)[1],
    levels = levels(regional_data$vessel))
  native <- predict(regional_fit, newdata = grid, re.form = NA,
    type = "response", se.fit = TRUE, cov.fit = TRUE)
  W <- kronecker(diag(length(years)), matrix(ref$weight / sum(ref$weight), nrow = 1))
  expected_mean <- drop(W %*% native$fit)
  expected_covariance <- W %*% native$cov.fit %*% t(W)
  index <- regional_indices[[r]]
  stopifnot(isTRUE(all.equal(index$table$Mean, expected_mean,
    tolerance = 1e-6, check.attributes = FALSE)))
  stopifnot(isTRUE(all.equal(index_vcov(index, scale = "response"),
    expected_covariance, tolerance = 1e-5, check.attributes = FALSE)))
  data.frame(Region = r,
    Max_mean_difference = max(abs(index$table$Mean - expected_mean)),
    Max_covariance_difference = max(abs(index_vcov(index, "response") - expected_covariance)))
})
knitr::kable(do.call(rbind, native_checks), digits = 10, row.names = FALSE)
```

For assessment export, keep each unscaled index table and its matching annual
log-covariance matrix together. For example:

```{r regional-assessment-export, eval=has_glmmTMB}
assessment_inputs <- lapply(regional_indices, function(index) {
  list(table = index_table(index), log_covariance = index_vcov(index))
})
stopifnot(all(vapply(assessment_inputs, function(x) {
  identical(x$table$Year, rownames(x$log_covariance)) &&
    identical(x$table$Year, colnames(x$log_covariance))
}, logical(1))))
```

These are **within-region annual covariance matrices**. Because both regions
use the same fitted model, their estimates are correlated across regions too.
Separate `cpue_index()` objects do not supply those cross-region blocks.
Do not combine them into a block-diagonal matrix, calculate a region-ratio
interval assuming independence, or treat the two series as independent
assessment likelihoods without an explicit justification. Joint regional
inference needs a separate calculation retaining cross-region dependence.
Relative normalisation also makes each full log-covariance matrix singular;
see [joint annual uncertainty](cpue-indices.html#joint-annual-uncertainty).

## Why a shared year effect is not enough

For comparison only, fit a second model **once** without the interaction.
With this additive log-link model and fixed references,
\(I_{r,t}=\exp(\alpha_t)C_r\): regions can differ in level, but their
proportional trends must be identical. No index extractor can recover two
different temporal trends from a model that has only one.

```{r regional-additive, eval=has_glmmTMB}
shared_year_fit <- update(regional_fit, . ~ . - year:area, data = regional_data)
shared_indices <- calculate_regions(shared_year_fit, references,
  rescale = 1, uncertainty = "none")
stopifnot(isTRUE(all.equal(shared_indices[[1]]$table$Mean,
  shared_indices[[2]]$table$Mean, tolerance = 1e-8)))
```

```{r regional-additive-plot, eval=has_glmmTMB, fig.cap="Counterexample: an additive shared-year log-link model produces coincident relative regional trends under fixed reference weights. The two point-estimate curves lie on top of one another. This is a consequence of the model, not a regional extraction error; uncertainty is deliberately omitted for this algebraic demonstration."}
plot_compare(shared_indices, show_probs = FALSE) + labs(colour = "Region")
```

## Scope of this example

The interaction model is a transparent illustration, not a recommendation to
fit an unrestricted area/year interaction to every dataset. Sparse cells,
unsupported covariate combinations, temporal smoothing, and partial pooling
need their own modelling choices. Inspect residuals and regional support;
pooled residual adequacy alone does not validate every regional curve.

The existing `cpue_index()` interface also supports GLMs, GAMs, brms, sdmTMB,
and tinyVAST within their documented prediction boundaries. This example
validates the stated glmmTMB interaction and zero-vessel-effect target, not
every possible regional or spatial specification. Area-weighted totals use
the separate `integrate_index()` workflow; they are not added to this
standardised regional-mean demonstration.
