---
title: "Residual validation"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Residual validation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 10,
  fig.height = 6, message = FALSE, warning = FALSE, dev.args = list(bg = "white"))
library(influ2)
library(ggplot2)
theme_set(theme_bw())
study <- readRDS(system.file("extdata", "n09-validation.rds", package = "influ2"))
has_bayesplot <- requireNamespace("bayesplot", quietly = TRUE) &&
  packageVersion("bayesplot") >= "1.16.0"
scenario_labels <- c(full = "Full model", omit_x = "Omit covariate", omit_st = "Omit yearly field")
scheme_labels <- c(fitted = "Fitted effects", conditional_draw = "One conditional draw",
  new_effects = "New effects", analytic_truth = "Known truth: analytic", finite_truth = "Known truth: 499 simulations")
labels_for <- function(data) {
  data$Model <- factor(scenario_labels[data$scenario], levels = unname(scenario_labels))
  data$Scheme <- factor(scheme_labels[data$conditioning], levels = unname(scheme_labels))
  data
}
percent <- function(x) sprintf("%.1f%%", 100 * x)
```

## What this study can establish

This is the first, deliberately bounded **N09 validation increment**. We
simulate data from known negative-binomial mixed and spatial models, fit both
correctly specified and deliberately incomplete candidates, and examine how
the existing residual diagnostics behave. It is a study of operating
characteristics, not just a check that plotting functions run.

There are three separate questions:

1. Do the PIT calculations behave sensibly when the generating distribution
   is known, before any parameters have been estimated?
2. How do parameter fitting and the choice of latent-effect conditioning
   change the diagnostics?
3. Which checks reveal deliberately omitted structure, and which can miss it?

**No calculation, conditioning default, or default panel has been changed by
this study.** The conclusions below apply to these designs, not to every CPUE
family or model. The [residual article](residual-diagnostics.html) remains the
guide to everyday usage. Randomised quantile residuals provide the underlying
distributional idea [@DunnSmyth1996]; fitted mixed-model diagnostics require
care over their conditioning and interpretation [@Waagepetersen2006].

## Fixed design and predictive targets

Each backend uses 100 independent datasets, with 480 observations over six
years. New latent effects and responses are generated for each dataset, while
the sampling design is held fixed. The response is a simulated count per
standardised sampling unit: it could represent CPUE, not necessarily total
catch. Both models use a log link and NB2 conditional variance
\(\mu + \mu^2/4\).

| Backend | Generating structure | Fitted candidates |
|:--|:--|:--|
| glmmTMB | Year, a continuous covariate, and 20 vessel random intercepts; four observations per vessel/year | Full model; omit the covariate |
| sdmTMB | Year, the same type of covariate, a persistent spatial field, and an independent yearly spatial field; 80 stations | Full model; omit the covariate; omit the yearly field |

The covariate distribution shifts gradually through time. The spatial case
uses a fixed 25-knot mesh in a 100 by 100 coordinate domain, range 30, spatial
SD 0.45, and yearly-field SD 0.6. The same mesh generates and fits the data;
this is **not** a study of mesh approximation error. Data generation uses the
native [`simulate_new()`](https://sdmtmb.github.io/sdmTMB/reference/simulate_new.html)
interface, with explicit known parameters, not a bootstrap from an estimated
model. The implementations are described by @Brooks2017 and @Anderson2025.

Every candidate is fitted once by maximum likelihood. There are no replacement
datasets or hidden convergence retries. The production study uses 499 response
simulations per diagnostic, batches of 50, and a 100-point response ECDF grid.
Three separate pilot datasets per backend checked execution and timing; they
do not appear in any results below.

| `conditioning` | Quantity held fixed or resimulated | Backends in this study |
|:--|:--|:--|
| `"fitted"` | Hold estimated latent effects at their fitted values; simulate new responses | Both |
| `"conditional_draw"` | Hold non-latent parameters fixed; draw one joint approximate conditional latent-effect vector and share it across all response simulations | sdmTMB |
| `"new_effects"` | Hold estimated non-latent parameters fixed; simulate new latent effects and responses in each replicate | Both |

The existing defaults are `"new_effects"` for glmmTMB and `"fitted"` for
sdmTMB. A single conditional draw is **not** a posterior predictive calculation
integrated over many latent draws. These schemes ask different questions; a
plot that looks closer to its reference is not, by itself, a reason to select
one. Native sdmTMB conditioning is documented in
[`simulate.sdmTMB()`](https://sdmtmb.github.io/sdmTMB/reference/simulate.sdmTMB.html).

### Fit eligibility and failures

A fit is eligible when its optimiser reports convergence, its reported Hessian
is positive definite, and its log likelihood is finite. Gradients and warnings
are recorded separately, not used as an undisclosed exclusion rule. The
following denominators include **every attempted fit**. Diagnostic summaries
use eligible fits only; their usable counts are shown below.

```{r n09-fit-table}
fit_table <- study$fit_summary
fit_table$scenario <- scenario_labels[fit_table$scenario]
fit_table$largest_gradient <- format(fit_table$largest_gradient, scientific = TRUE, digits = 3)
knitr::kable(fit_table[, c("backend", "scenario", "attempted", "eligible", "warnings", "errors", "largest_gradient")],
  col.names = c("Backend", "Candidate", "Attempted", "Eligible", "Warnings", "Errors", "Largest gradient"), digits = 4)
```

All 500 attempted fits were eligible, and all 1,960 diagnostic/control records
completed without captured warnings or errors. The largest recorded glmmTMB
gradient was approximately 0.0105; gradients were reported, not used to change
the pre-specified eligibility rule. An eligible fit is not proof of an adequate model or a perfectly estimated
latent covariance. Fit-status details, including excluded attempts, are
retained in the downloadable results object.

## First check the reference itself

Before interpreting any fitted-model results, an independent-uniform control
found a grid-alignment problem in **bayesplot 1.16.0**, the optional plotting
dependency used by influ2's PIT ECDF panels. Its last 100 reference limits are
calculated for `1:100 / 100`, but are displayed at
`seq(0, 1, length.out = 100)`. The empirical curve and the limits consequently
refer to different horizontal positions. This finding concerns that installed
version and method, not the underlying theory of simultaneous ECDF bands
[@Sailynoja2022].

The control uses 10,000 independent samples of 480 uniform values. Crossing
means that at least one empirical-CDF value lies outside the reference limits.
The aligned row is an **audit calculation**, not a patched plotting function.

```{r n09-band-table}
audit <- study$band_audit
audit$Reference <- c(displayed = "bayesplot 1.16.0, as displayed",
  aligned = "Same limits at their own grid", dkw = "Independent DKW bound")[audit$method]
audit$Crossing <- percent(audit$rate)
audit$Interval <- paste(percent(audit$lower), "to", percent(audit$upper))
knitr::kable(audit[, c("Reference", "n", "Crossing", "Interval")],
  col.names = c("Reference", "Samples", "Crossing rate", "95% Monte Carlo interval"))
```

The displayed limit crossing rate is `r percent(audit$rate[audit$method == "displayed"])`,
not approximately 5%. **Do not treat crossings of those displayed optional
bands as a calibrated 5% test.** This does not change the stored PIT values,
normal-score Q-Q calculation, or response ECDF panel. No dependency patch or
default change has been made here; the alignment correction is a separate
follow-up decision.

For the main study we therefore use an independent summary: the exact maximum
distance of the empirical PIT CDF from uniformity,
\(D=\sup_u|\widehat F(u)-u|\). We record whether it exceeds the 95%
Dvoretzky--Kiefer--Wolfowitz (DKW) bound
\(\sqrt{\log(2/0.05)/(2n)}\) [@Massart1990]. This controls crossing probability
under **independent uniform values**, but it is **not a calibrated hypothesis
test for fitted-model residuals**. There is no fitted-model p-value or automatic
pass/fail model selection here.

### Known-truth controls

For each generated dataset, two controls use the actual conditional mean,
realised latent effects, and NB2 size:

- The analytic randomised PIT is
  \(F(y-1)+U\{F(y)-F(y-1)\}\), with an independent uniform randomiser.
- The finite-simulation control generates 499 independent response replicates
  from that same known distribution and uses `as_influ_residuals()`.

Neither control estimates a model. Their comparison isolates finite-rank
behaviour from parameter fitting. Normal scores remain `qnorm(pit)`; Pearson
residuals are not used.

```{r n09-oracle-table}
oracle <- study$summary[study$summary$scenario == "oracle", ]
oracle$Control <- scheme_labels[oracle$conditioning]
oracle$Interval <- paste(percent(oracle$lower), "to", percent(oracle$upper))
knitr::kable(oracle[, c("backend", "Control", "usable", "crossings", "Interval", "residual_sd")],
  col.names = c("Backend", "Control", "Datasets", "Crossings", "95% Monte Carlo interval", "Mean residual SD"), digits = 3)
```

With only 100 datasets per backend, small differences between these controls
are Monte Carlo variation, not evidence of equivalence or a preferred count.
The larger independent-uniform audit checks the reference separately.

## Fitting and conditioning change the picture

The full glmmTMB model crossed the DKW reference in 0/100 datasets with fitted
effects and 5/100 with new effects. The full sdmTMB model crossed in 6/100 with
fitted effects, 6/100 with one conditional draw, and 28/100 with new fields.
The last result is not evidence that the correctly specified spatial model is
wrong: the iid reference does not account for the residual dependence under
that predictive target.

The following rates use 499 simulations and the primary seed. Error bars are
Wilson 95% Monte Carlo intervals **over independently generated datasets**,
not uncertainty intervals around the CPUE index. Repeated seed/count runs do
not increase these denominators.

```{r n09-crossing-rates, fig.cap="Crossing of the independent-uniform DKW reference across 100 generated datasets per backend, conditional on fit eligibility. Points show crossing rates and bars show Wilson 95% Monte Carlo intervals; the dotted line marks 5%. Colours distinguish latent-effect conditioning. These rates describe this simulation design and are not calibrated fitted-model rejection probabilities. The known-truth controls are tabulated separately.", fig.alt="DKW reference crossing rates by backend, omitted structure, and conditioning scheme."}
rates <- labels_for(study$summary[study$summary$scenario != "oracle", ])
ggplot(rates, aes(Model, rate, colour = Scheme)) +
  geom_hline(yintercept = .05, linetype = 3, colour = "grey45") +
  geom_pointrange(aes(ymin = lower, ymax = upper),
    position = position_dodge(width = .5)) +
  facet_wrap(~ backend, scales = "free_x") +
  scale_y_continuous(limits = c(0, 1), labels = function(x) paste0(100 * x, "%")) +
  labs(x = NULL, y = "Datasets crossing the DKW reference", colour = "Conditioning") +
  theme(legend.position = "bottom")
```

```{r n09-rate-table}
rates$Interval <- paste(percent(rates$lower), "to", percent(rates$upper))
knitr::kable(rates[, c("backend", "Model", "Scheme", "usable", "crossings", "Interval")],
  col.names = c("Backend", "Candidate", "Conditioning", "Usable", "Crossings", "95% Monte Carlo interval"))
```

### Structure can remain when the marginal distribution looks plausible

Distributional agreement alone is insufficient. We also record residual
association with the **known omitted covariate**, and a descriptive spatial
neighbour score. The latter uses a symmetric four-nearest-neighbour graph
within each year and residuals centred within year. It is not a new public
diagnostic function or a calibrated spatial test. It provides a known-design
check for spatial pattern that a pooled histogram or Q-Q plot can miss.

```{r n09-structure-table}
structure_table <- rates[, c("backend", "Model", "Scheme", "residual_sd", "covariate_spearman", "spatial_score")]
knitr::kable(structure_table,
  col.names = c("Backend", "Candidate", "Conditioning", "Mean residual SD", "Mean correlation with covariate", "Mean neighbour score"), digits = 3)
```

Residual SD is on the standard-normal scale. Correlations are Spearman
correlations within datasets, then averaged. A missing neighbour score for
glmmTMB means that no spatial graph was specified, not that spatial
independence was established. Full replicate-level results additionally
include fitted-mean association, year-mean variation, and the fractions of
Q-Q points and response-grid points outside their **pointwise** intervals.
Those fractions are descriptive: points within a plot are not independent
replicates.

Two failures of a distribution-only interpretation are particularly clear:

- Omitting the generating covariate leaves substantial residual association
  with it: mean Spearman correlation is about 0.61 for fitted-effect glmmTMB
  diagnostics and 0.61 for fitted-effect sdmTMB diagnostics. Yet their DKW
  crossing counts are only 0/100 and 1/100, respectively. Re-estimating the
  remaining parameters can produce a plausible pooled residual distribution
  without removing the omitted covariate pattern.
- Omitting the yearly spatial field gives 0/100 DKW crossings under fitted
  sdmTMB conditioning, but the mean within-year neighbour score is about 0.19.
  Spatial pattern remains despite the reassuring marginal distribution.

For the full sdmTMB model, fitted-effect normal scores have mean SD 0.906,
compared with 1.004 for the single conditional-draw scheme. The neighbour
scores also differ. This supports treating conditioning as an explicit
scientific choice, not declaring the method with the most normal-looking
scores universally preferable.

## Worked plots without refitting

The figures use the **first** dataset with all planned comparisons eligible:
glmmTMB replicate `r study$examples$glmmTMB$replicate` and sdmTMB replicate
`r study$examples$sdmTMB$replicate`. Selection was fixed in advance and did not
depend on how striking the plots looked. The multi-dataset results above carry
more evidence than these individual illustrations.

```{r n09-load-checks}
validation <- readRDS(system.file("extdata", "n09-validation.rds", package = "influ2"))
mixed <- validation$examples$glmmTMB$checks
spatial <- validation$examples$sdmTMB$checks
```

### glmmTMB: retain or omit the covariate

Both plots use fitted effects to make the conditional-response comparison
explicit; this is an example choice, **not a change to the glmmTMB default**.
They use the same observations and primary seed. Fitted-year effects can make
year-wise centres look reassuring even when a covariate is omitted.

```{r n09-mixed-full, fig.width=11, fig.height=8, fig.cap="Existing four-panel diagnostic for the correctly specified glmmTMB model in the pre-selected example dataset, holding fitted vessel effects fixed. Q-Q, fitted-value, and year panels use normal-score PIT residuals. The Q-Q ribbon is a pointwise independent-uniform reference; the response ECDF ribbon is a pointwise predictive band. Neither supplies a calibrated fitted-model pass/fail test.", fig.alt="Four-panel diagnostic for the full glmmTMB model with fitted vessel effects."}
plot(mixed[["full:fitted"]], response_scale = "log1p")
```

```{r n09-mixed-omitted, fig.width=11, fig.height=8, fig.cap="The same glmmTMB observations and fitted-effect conditioning after omitting the generating covariate. Other parameters, including negative-binomial dispersion, are re-estimated once in this incomplete candidate. A plausible marginal residual distribution does not establish that covariate-related structure has disappeared; the repeated-study correlation summary checks that separately.", fig.alt="Four-panel diagnostic for the glmmTMB candidate omitting the covariate."}
plot(mixed[["omit_x:fitted"]], response_scale = "log1p")
```

### sdmTMB: retain or omit the yearly spatial field

Here fitted-effect conditioning is the existing sdmTMB default. The full model
can use fitted fields to explain local variation, whereas the incomplete
candidate has no yearly field. Neither four-panel plot is a residual map;
the neighbour summary above checks an additional spatial question.

```{r n09-spatial-full, fig.width=11, fig.height=8, fig.cap="Existing four-panel diagnostic for the correctly specified sdmTMB model in the pre-selected spatial dataset, holding fitted persistent and yearly spatial effects fixed. The normal-score PIT and response-distribution panels are conditional model checks. Their nominal reference bands do not account automatically for estimating latent effects from these same observations.", fig.alt="Four-panel diagnostic for the full spatial and yearly-field sdmTMB model."}
plot(spatial[["full:fitted"]], response_scale = "log1p")
```

```{r n09-spatial-omitted, fig.width=11, fig.height=8, fig.cap="The same sdmTMB observations after omitting the yearly spatial field, retaining the fixed covariate and persistent spatial field. Fitted-effect conditioning is used again. The marginal panels and the separate within-year neighbour score address different aspects of this deliberate misspecification; absence of a strong Q-Q departure is not evidence that the omitted field is unnecessary.", fig.alt="Four-panel diagnostic for the sdmTMB candidate without the yearly spatial field."}
plot(spatial[["omit_st:fitted"]], response_scale = "log1p")
```

### The same PIT values on a uniform scale

The optional plots below reuse the saved full-model sdmTMB PIT values. They
make the conditioning contrast visible without running the models again.
The bands are supplied by the locally installed bayesplot; the known 1.16.0
alignment limitation above must be kept in mind. The numerical study is frozen
to its recorded version, even if a future plotting dependency changes.

```{r n09-pit-comparison, eval=has_bayesplot, fig.width=11, fig.height=8, fig.cap="PIT ECDF difference displays for the same full sdmTMB fit: fitted effects (upper left), one shared conditional latent draw (upper right), and new effects (lower left). Curves reuse stored PIT values; zero is the uniform reference. Pale limits are the native optional bayesplot display, which has the documented grid-alignment problem in version 1.16.0. Do not read crossings as calibrated rejection decisions. No new response simulations or model fits are performed.", fig.alt="Three PIT ECDF difference displays contrasting sdmTMB conditioning targets."}
patchwork::wrap_plots(
  plot(spatial[["full:fitted"]], type = "pit_ecdf_diff") + labs(title = "Fitted effects"),
  plot(spatial[["full:conditional_draw"]], type = "pit_ecdf_diff") + labs(title = "One conditional draw"),
  plot(spatial[["full:new_effects"]], type = "pit_ecdf_diff") + labs(title = "New effects"),
  ncol = 2
)
```

## Seed and simulation-count sensitivity

For dataset IDs 1--10 only, we repeat each eligible diagnostic with a second
seed at 499 simulations and with the primary seed at 1,999 simulations. Models
are **not refitted**. A conditional-draw seed also changes the one shared
latent vector, so its sensitivity is not solely finite response-simulation
noise. More response simulations do not average over that single latent draw.

```{r n09-sensitivity-table}
sensitivity <- labels_for(study$sensitivity)
sensitivity$Change <- ifelse(sensitivity$variant == "second_seed", "Second seed, 499", "Same seed, 1,999")
knitr::kable(sensitivity[, c("backend", "Model", "Scheme", "Change", "paired", "flag_changes", "mean_absolute_distance_change")],
  col.names = c("Backend", "Candidate", "Conditioning", "Rerun", "Pairs", "Crossing changes", "Mean absolute change in D"), digits = 4)
```

These ten paired datasets are a sensitivity screen, not an estimate precise
enough to recommend a universal simulation count. Near a reference boundary,
changing a binary crossing flag can exaggerate a small numerical change; read
the continuous distance alongside the count. The defaults remain unchanged.

Across the 40 paired glmmTMB comparisons per sensitivity variant, no DKW flags
changed. Across the 90 sdmTMB comparisons, three changed with the second seed
and two with 1,999 simulations. Those counts pool different scenarios and
schemes and are descriptive only; the disaggregated table is the primary
sensitivity record.

## Interpretation and the next review

This first increment separates three sources of behaviour: finite-simulation
ranks, estimating a model from the checked observations, and the predictive
target defined by conditioning. They should not be combined into a single
claim that a backend's residuals are either "valid" or "invalid".

Review the following with Nicholas before changing statistical defaults:

1. Address the independently reproduced optional reference-band alignment
   problem separately from residual calibration.
2. Decide which scientific question each conditioning scheme should answer,
   using the full-model and omitted-structure results together. Do not pick a
   default simply because its Q-Q curve looks closest to normal.
3. Decide whether a later increment needs calibrated fitted-model envelopes,
   repeated refitting, or held-out prediction checks. None is provided here.
4. Identify the next genuinely different design to validate, rather than
   extending these conclusions to all six backends or all response families.

Limitations include one family, one sample size, one fixed design and spatial
mesh, balanced yearly effort, modest replication, and diagnostics evaluated
on the fitting observations. There are no brms, tinyVAST, hurdle/delta, sparse
sampling, range-boundary, or out-of-sample validation claims. In particular,
the default four panels do not replace a spatial residual investigation.

## Reproduction and compact results

The [fixed protocol and developer scripts](https://github.com/quantifish/influ2/tree/master/tools/n09)
contain the data generator, seed schedule, metrics, timing pilot, and report
builder. From a checkout with the recorded optional dependencies installed:

```sh
Rscript --vanilla tools/n09/test-study.R
Rscript --vanilla tools/n09/run.R production /absolute/new/output-directory
Rscript --vanilla tools/n09/report.R /absolute/new/output-directory
```

Use one BLAS/OpenMP thread for the recorded timing setup. Each dataset is
checkpointed; resuming requires identical configuration and source hashes.
The report refuses incomplete production runs. It writes replicate-level CSV
files to the output directory and the compact RDS used here to `inst/extdata/`.
The source commit records the package implementation baseline; separate hashes
identify the study scripts and protocol, which were added by this increment.

```{r n09-versions}
knitr::kable(data.frame(Package = names(study$metadata$package_versions),
  Version = unname(study$metadata$package_versions)))
```

The installed artefact contains all attempted-fit statuses, successful and
failed diagnostic attempts, seeds, versions, source hashes, summaries, and the
two selected sets of compact diagnostic objects. It contains no fitted models
or observation-by-simulation matrices. Its compressed size is
`r round(file.info(system.file("extdata", "n09-validation.rds", package = "influ2"))$size / 1024)` KiB.
The article and ordinary package tests read these results; **CRAN and CI do not
rerun the Monte Carlo study**. Rebuilding this article is not evidence that the
study was rerun under newer dependency versions.

## References
