---
title: "Residual diagnostics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Residual diagnostics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 5,
  message = FALSE, warning = FALSE, dev.args = list(bg = "white")
)
has_MASS <- requireNamespace("MASS", quietly = TRUE)
has_DHARMa <- has_MASS && requireNamespace("DHARMa", quietly = TRUE)
library(influ2)
library(ggplot2)
theme_set(theme_bw())
```

## Three complementary questions

A CPUE model needs more than a plausible-looking index. We need to ask whether
it describes the observations adequately, what drives the standardised index,
and how robust that index is to modelling choices. Residual checks address the
first question; influence, CDI, and step plots primarily address the second.
Comparing credible candidate models helps address the third. None is a
substitute for the others.

This article uses the same **simulated lobster CPUE** example as
[Influence diagnostics](influ2.html). It includes deliberately incomplete and
poorly fitting candidates, not just the preferred example model. The aim is
to connect a visible diagnostic pattern to a possible modelling response,
without turning individual plots or p-values into automatic selection rules.

| influ2 helper | Question it helps answer | Important limitation |
|:--|:--|:--|
| `plot_predicted_residuals()` | Does residual behaviour change with the fitted mean? | Uses the fitted model's native residual definition; it is not a predictive interval plot. |
| `plot_implied_residuals()` | Do groups suggest departures from the common year effect? | An exploratory CPUE display, not a fitted interaction or a distributional goodness-of-fit test. |
| `plot_qq()` | How does the residual distribution compare with normal quantiles? | Count-model Pearson and deviance residuals need not be normal, even under a suitable model. |

These functions take a fitted model, not an `influ_diag` summary: the latter
deliberately does not retain all observation-level predictions and residuals.

## Fit competing lobster models

The data include uneven sampling across years and months, changing depth and
soak time, and gaps in coverage. Catch was simulated from a negative-binomial
distribution. We fit a reduced negative-binomial model that omits depth and
soak time, a fuller negative-binomial model, and a Poisson model with the same
full mean formula. All use exactly the same observations.

```{r residual-models, eval=has_MASS}
data(lobsters_per_pot)

lobster_reduced <- MASS::glm.nb(
  lobsters ~ year + month,
  data = lobsters_per_pot
)
lobster_nb <- MASS::glm.nb(
  lobsters ~ year + month + poly(depth, 3) + poly(soak, 3),
  data = lobsters_per_pot
)
lobster_poisson <- glm(
  formula(lobster_nb), family = poisson(link = "log"),
  data = lobsters_per_pot
)
```

`MASS` is optional. The executed model examples are skipped if it is not
installed. Nothing in this article fits an MCMC model.

## Residuals against fitted values and predictors

Pearson residuals divide response errors by the model's observation standard
deviation. Look for a changing centre, unusual spread, or influential tails,
but remember that low fitted counts produce asymmetry and discrete bands.
The residual axis must retain both negative and positive values.

```{r residual-fitted, eval=has_MASS, fig.cap="Pearson residuals against fitted lobster catch for reduced and full negative-binomial models. Blue smooths help reveal changes in the residual centre; the dotted line marks zero.", fig.alt="Two residual-versus-fitted panels comparing reduced and full negative-binomial models."}
patchwork::wrap_plots(
  plot_predicted_residuals(lobster_reduced) + labs(title = "Reduced NB"),
  plot_predicted_residuals(lobster_nb) + labs(title = "Full NB"),
  ncol = 2
)
```

A fitted-value plot can hide structure in an omitted covariate. Plotting the
same residuals against depth and soak time is more direct. Here a systematic
pattern in the reduced model motivates including those relationships, rather
than treating extra dispersion alone as a solution.

```{r residual-predictors, eval=has_MASS, fig.height=6, fig.cap="Pearson residuals against depth and soak time. The reduced model omits both predictors; the full model includes their polynomial effects. Smooth trends reveal structure that a fitted-value plot can conceal.", fig.alt="Four panels compare residual patterns against depth and soak time for reduced and full models."}
residual_data <- do.call(rbind, lapply(
  c("Reduced NB", "Full NB"), function(label) {
    fit <- if (label == "Reduced NB") lobster_reduced else lobster_nb
    r <- residuals(fit, type = "pearson")
    rbind(
      data.frame(model = label, predictor = "Depth (m)",
                 value = lobsters_per_pot$depth, residual = r),
      data.frame(model = label, predictor = "Soak time (hours)",
                 value = lobsters_per_pot$soak, residual = r)
    )
  }
))
residual_data$model <- factor(
  residual_data$model, levels = c("Reduced NB", "Full NB")
)
ggplot(residual_data, aes(value, residual)) +
  geom_hline(yintercept = 0, linetype = 3, colour = "grey45") +
  geom_point(alpha = 0.08, size = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  facet_grid(model ~ predictor, scales = "free_x") +
  labs(x = NULL, y = "Pearson residual")
```

For real fisheries, repeat this examination against vessel, gear, year,
season, and location. Sparse regions and changing fleet composition deserve
particular attention. A smooth is descriptive: these are in-sample residuals,
not independent validation observations.

## Implied residual coefficients

New Zealand inshore CPUE reports use implied coefficients to explore
departures from a shared year effect [@StarrKendrick2019; @Middleton2025]. For
each year-by-group stratum, `plot_implied_residuals()` adds its mean residual
to the centred link-scale year effect. The common year effect is grey; purple
points show the implied departures. Here the groups are months, but an area,
vessel group, or other categorical variable could be used.

```{r residual-implied, eval=has_MASS, fig.height=8, fig.cap="Implied year coefficients by month for the full negative-binomial lobster model. Grey lines show the common centred year effect; purple values add each stratum's mean Pearson residual. Bars are plus or minus one standard error of that residual mean, and point size represents record count. Strata with fewer than 10 records are omitted.", fig.alt="Twelve monthly panels compare implied coefficients with the same grey annual baseline."}
plot_implied_residuals(
  lobster_nb,
  data = lobsters_per_pot,
  year = "year", groups = "month", min_n = 10
)
```

The bars exclude uncertainty in the year coefficient and its covariance with
the residual mean; they are **not** confidence intervals for a fitted
year-by-month interaction. Adding a Pearson residual to a link-scale effect
is an interpretive convention, not an exact interaction estimator. Native
residual scaling also differs between model packages. Persistent departures
suggest checking sampling support and an explicit interaction or other
structure, then refitting and reassessing the model.

Keep the original observation row names when supplying `data`. influ2 aligns
omitted, subsetted, or reordered rows with the fitted model, and rejects data
whose identity cannot be verified. This prevents residuals from being assigned
silently to the wrong year or group.

## Normal Q-Q plots: useful, but not universal

```{r residual-normal-qq, eval=has_MASS, fig.cap="Normal Q-Q plot of Pearson residuals from the full negative-binomial model. Curvature is not, by itself, evidence against the count model: these residuals are not expected to be normally distributed.", fig.alt="A normal quantile comparison of the negative-binomial model's Pearson residuals, with a quartile reference line."}
plot_qq(lobster_nb, type = "pearson")
```

The reference line in `plot_qq()` passes through the chosen quantiles (the
first and third quartiles by default). Its `probs` argument does not define a
confidence envelope. Normal residual Q-Q plots are most natural for Gaussian
errors; for counts, zeros, skewed positive responses, and mixtures, use a
distribution-aware check as well.

## Simulation-based checks with DHARMa

DHARMa supplies simulation-based quantile residuals and associated diagnostics
[@DHARMa2026]. It is a suggested, optional dependency, not part of the compact
influence calculation. The following example contrasts Poisson and
negative-binomial variation while holding the mean formula fixed.

```{r residual-dharma, eval=has_DHARMa}
dharma_poisson <- DHARMa::simulateResiduals(
  fittedModel = lobster_poisson, n = 250, refit = FALSE, seed = 20260907
)
dharma_nb <- DHARMa::simulateResiduals(
  fittedModel = lobster_nb, n = 250, refit = FALSE, seed = 20260907
)
```

Each object is calculated once and reused. With `refit = FALSE`, simulated
responses are generated from the fitted parameters; the model is not refitted
250 times. This quick example uses 250 simulations. For substantive work,
increase that number and check stability. Simulation storage still scales
with observations times simulations, independently of influ2's compact
uncertainty-retention settings.

```{r residual-dharma-qq, eval=has_DHARMa, fig.cap="DHARMa uniform Q-Q checks for Poisson and negative-binomial lobster models with identical mean formulas. The Poisson model does not accommodate the simulated overdispersion. Formal tests are suppressed here so interpretation starts with the diagnostic pattern.", fig.alt="Two uniform quantile plots contrast Poisson misfit with the negative-binomial candidate."}
par(mfrow = c(1, 2))
DHARMa::plotQQunif(
  dharma_poisson, testUniformity = FALSE, testOutliers = FALSE,
  testDispersion = FALSE, main = "Poisson"
)
DHARMa::plotQQunif(
  dharma_nb, testUniformity = FALSE, testOutliers = FALSE,
  testDispersion = FALSE, main = "Negative binomial"
)
par(mfrow = c(1, 1))
```

DHARMa residuals are approximately **uniform on zero to one** under an
adequate simulation model, not normally distributed. Do not feed them into a
normal Q-Q plot without a deliberate transformation. Distributional agreement
does not rule out remaining covariate patterns:

```{r residual-dharma-depth, eval=has_DHARMa, fig.cap="Simulation-based quantile residuals against depth for the two full-mean lobster models. The thin black dashed line marks the uniform median of 0.5; the thicker dashed curve is a descriptive smooth. Depth is rank-scaled. These are not independent validation tests.", fig.alt="Two plots show uniform-scale simulation residuals against ranked depth."}
par(mfrow = c(1, 2))
DHARMa::plotResiduals(
  dharma_poisson, form = lobsters_per_pot$depth, quantreg = FALSE,
  main = "Poisson: depth"
)
DHARMa::plotResiduals(
  dharma_nb, form = lobsters_per_pot$depth, quantreg = FALSE,
  main = "Negative binomial: depth"
)
par(mfrow = c(1, 1))
```

The simulated responses can also check quantities of direct fisheries
interest. Compare the observed proportion of empty pots and the upper catch
quantile with the same quantities calculated from each simulated dataset.
These are predictive checks at fitted parameters, not posterior predictive
checks or confidence intervals for an index.

```{r residual-predictive-summaries, eval=has_DHARMa, fig.height=6, fig.cap="Observed empty-pot proportions and upper catch quantiles (purple lines) against their distributions across 250 simulated datasets. The histograms expose consequences of the candidate observation distributions that a standardised index alone does not show.", fig.alt="Four histograms compare simulated empty-pot proportions and 95th catch percentiles with their observed values."}
statistics <- list(
  "Empty-pot proportion" = function(x) mean(x == 0),
  "95th catch percentile" = function(x) unname(quantile(x, 0.95))
)
simulations <- list(Poisson = dharma_poisson, NB = dharma_nb)
predictive_checks <- do.call(rbind, lapply(names(simulations), function(model) {
  do.call(rbind, lapply(names(statistics), function(statistic) {
    data.frame(
      model = model, statistic = statistic,
      value = apply(simulations[[model]]$simulatedResponse, 2, statistics[[statistic]])
    )
  }))
}))
observed_checks <- data.frame(
  statistic = names(statistics),
  observed = vapply(statistics, function(f) f(lobsters_per_pot$lobsters), numeric(1))
)
ggplot(predictive_checks, aes(value)) +
  geom_histogram(bins = 20, fill = "grey75", colour = "white") +
  geom_vline(data = observed_checks, aes(xintercept = observed),
             colour = "purple4", linewidth = 0.8) +
  facet_grid(model ~ statistic, scales = "free_x") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Statistic across one simulated dataset", y = "Simulated datasets")
```

More zeros than simulated do not prove that a zero-inflated model is needed:
the mean structure or dispersion can also be wrong. For hurdle or delta
models, check encounter probability, the positive-response distribution, and
the combined response. The combined check alone can conceal compensating
errors in the components.

## Which residuals answer which question?

| Residual or approach | Useful for | What to watch |
|:--|:--|:--|
| Response, `y - fitted` | Errors in catch or CPUE units; patterns against predictors | Variance often changes with the mean; tails and zeros depend on the family. |
| Pearson | Mean/variance patterns after scaling by the model's observation SD | Not generally normal for non-Gaussian data; native weighting and Bayesian definitions differ. |
| Deviance | Signed contributions based on likelihood discrepancy | Family-specific, not automatically normal, and not supplied by every backend. |
| PIT / randomised quantile | Distribution-aware checks using a predictive CDF | Discrete observations require randomisation; conditioning and estimated parameters affect calibration. |
| DHARMa | Simulation-based quantile checks, predictive summaries, and residual-pattern tools | It implements a PIT-like approach, rather than a distinct residual principle; simulations must match the response and hierarchy being checked. |
| One-step-ahead (OSA) | Sequential predictive checks in dependent latent-variable models | Requires suitable implementation, an observation order, and accurate conditional prediction; often more expensive. |

At a probability mass, a randomised PIT value is drawn between the CDF just
below the observation, `F(y-)`, and `F(y)`. For integer counts this becomes
`F(y - 1)` to `F(y)`; hurdle and Tweedie responses require randomisation at
their zero atom. Applying `qnorm()` puts the result on the standard-normal
scale [@DunnSmyth1996]. At continuous responses, use `F(y)` directly. Exact
reference distributions assume the correct predictive distribution with known
parameters; fitted-model checks generally have approximate calibration.
Keep a reproducible seed and check sensitivity to randomisation, rather than
choosing the most attractive plot.

### Random effects, space, and time

Conditioning is a separate choice from residual scale. Conditional simulations
hold fitted random effects fixed and focus on observation-level behaviour;
unconditional simulations regenerate random effects and inspect more of the
hierarchy. Neither by itself validates the whole model. For latent spatial
fields, plugging in fitted modes can distort goodness-of-fit diagnostics.

DHARMa changed its default to conditional simulations in version 0.5.0.
Record the package version and the conditioning used, and consult the native
simulation method rather than assuming that identical arguments mean the same
thing in every backend. The DHARMa development NEWS also records a fix after
0.5.0 for mutation of glmmTMB simulation settings; check that fix before using
that combination for repeated production diagnostics. See the
[DHARMa documentation](https://CRAN.R-project.org/package=DHARMa) and
[NEWS](https://florianhartig.r-universe.dev/DHARMa/NEWS).

Map residuals and inspect them through time as well as against predictors.
Spatial or temporal dependence can invalidate routine independent-residual
test p-values. Repeated locations and timestamps need appropriate grouping,
and that grouping changes the question being tested. A well-calibrated
in-sample plot does not establish out-of-sample predictive skill.

### sdmTMB

sdmTMB recommends `type = "mle-mvn"` PIT residuals: fixed effects stay at
their maximum-likelihood estimates, while one approximate conditional
random-effect draw is used. These are **not OSA residuals**. Its
`"mle-eb"` plug-in mode is discouraged for goodness-of-fit checks; the
`"mle-mcmc"` alternative requires a suitable converged MCMC draw. See the
[native residual reference](https://sdmtmb.github.io/sdmTMB/reference/residuals.sdmTMB.html)
and [worked residual examples](https://sdmtmb.github.io/sdmTMB/articles/residual-checking.html).

For a single-response model such as `pcod_model` in
[Spatial and spatiotemporal diagnostics](spatial-spatiotemporal.html), the
native normal-scale PIT residuals can be displayed with influ2:

```{r residual-sdmtmb-recipe, eval=FALSE}
set.seed(41)
plot_qq(pcod_model, type = "mle-mvn") +
  geom_abline(intercept = 0, slope = 1, colour = "purple4") +
  labs(subtitle = "Purple: standard-normal identity; grey: fitted quartile line")
```

Unlike the fitted quartile line, the identity reference also reveals location
and scale departures from the expected standard-normal PIT distribution.

sdmTMB also has a dedicated DHARMa bridge. The following recipe reuses an
already fitted model; it is not evaluated again in this article.

```{r residual-sdmtmb-dharma-recipe, eval=FALSE}
set.seed(41)
pcod_simulations <- simulate(pcod_model, nsim = 250, type = "mle-mvn")
pcod_dharma <- sdmTMB::dharma_residuals(
  pcod_simulations, pcod_model, plot = FALSE, return_DHARMa = TRUE
)
DHARMa::plotQQunif(
  pcod_dharma, testUniformity = FALSE,
  testOutliers = FALSE, testDispersion = FALSE
)
```

The [sdmTMB bridge](https://sdmtmb.github.io/sdmTMB/reference/dharma_residuals.html)
can check the combined delta/hurdle response. Its analytical residuals instead
select a component. influ2's three generic residual helpers currently reject
sdmTMB delta fits: pairing occurrence residuals with unconditional fitted
catch would be misleading. Use the native component-specific workflow and
matching predictions instead. Native Pearson support is also family-specific;
influ2 does not silently replace it with another residual type.

### tinyVAST

The tinyVAST 1.6.2 interface provides deviance and response residual types,
not Pearson residuals or an OSA convenience interface. In checks against that
version, single-response deviance residuals worked with all three influ2
helpers. Its native response method returned an empty vector, so influ2 now
reports that failure clearly rather than drawing an empty diagnostic.

```{r residual-tinyvast-recipe, eval=FALSE}
plot_predicted_residuals(tiny_model, type = "deviance")
plot_qq(tiny_model, type = "deviance")
```

The [tinyVAST simulation interface](https://vast-lib.github.io/tinyVAST/reference/simulate.tinyVAST.html)
can support an external `DHARMa::createDHARMa()` workflow, but it needs careful
response selection and observation alignment for multivariate models. Its
`"mle-mvn"` simulation draws a new approximate random-effect vector for each
replicate, unlike sdmTMB's single-draw residual recipe. This article does not
claim those procedures have identical calibration, or provide an automatic
multivariate DHARMa adapter.

### GLM, GAM, glmmTMB, and BRMS

The ordinary GLM, GAM, and glmmTMB residual plots use their native methods.
Current glmmTMB also provides a `"dunn-smyth"` option, but family and version
limitations matter, including fixes for multi-trial binomial responses.
Do not assume it handles every zero-inflated mixture correctly. The
[glmmTMB residual source](https://github.com/glmmTMB/glmmTMB/blob/master/glmmTMB/R/methods.R)
and [release notes](https://glmmtmb.r-universe.dev/glmmTMB/NEWS) describe the
available implementation. Being built on TMB does not itself provide an OSA
interface.

For BRMS, use the **original complete `brmsfit`**. The compact example fixtures
in influ2 retain joint draws for influence calculations but omit the Stan
state needed by native prediction and residual methods. They cannot support
these residual helpers, `get_bayes_R2()`, or `table_criterion()`. Those two
comparison helpers remain BRMS-specific; they are not generic frequentist
model-selection tables.

BRMS's native residual summaries depend on its prediction method; Pearson
residuals are based on predictive dispersion, not simply the GLM formula.
Posterior predictive checks (`brms::pp_check()`) can retain features obscured
by a residual mean. Reusing a full fitted model does not rerun MCMC, but
prediction across posterior draws can still require substantial memory.
See the [BRMS residual reference](https://paulbuerkner.com/brms/reference/residuals.brmsfit.html).

### OSA is a specialised additional workflow

OSA residuals use the predictive distribution of each observation conditional
on the preceding observations, which can account for latent dependence
[@Thygesen2017]. The order and conditioning set must therefore be explicit.
TMB offers [`oneStepPredict()`](https://search.r-project.org/CRAN/refmans/TMB/html/oneStepPredict.html),
but its methods require suitable template support and approximation checks.
They are not universally available through `residuals()` in glmmTMB, sdmTMB,
or tinyVAST. influ2 does not yet implement an OSA adapter. Likewise, a
refitting simulation study is a separate, potentially costly validation
exercise, not a default action performed by a residual plot.

## Connect model checks back to the index

An observation model that fails to reproduce dispersion or zeros may still
produce a plausible year curve. Conversely, changing a well-supported mean
relationship can materially alter the index. After inspecting residuals,
use `plot_compare()` and refitted step plots to explain those consequences;
the [main article](influ2.html) demonstrates both.

Keep the same observations, years, response component, and index scale when
comparing models. influ2's current comparisons are centred year-effect
contrasts, **not area-weighted abundance indices**. Prediction-grid coverage,
area expansion, and extrapolation are additional downstream checks.

The practical sequence is to identify a residual pattern, consider plausible
data or model explanations, fit defensible alternatives, and then inspect
both model adequacy and index sensitivity. Residual tests alone should not
decide which CPUE series enters an assessment.

## References
