---
title: "Spatial and spatiotemporal diagnostics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Spatial and spatiotemporal diagnostics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
has_sdmTMB <- requireNamespace("sdmTMB", quietly = TRUE)
has_tinyVAST <- requireNamespace("tinyVAST", quietly = TRUE) &&
  requireNamespace("fmesher", quietly = TRUE)
suppressPackageStartupMessages(library(ggplot2))
```

## Why separate field components?

A spatiotemporal model can contain at least three distinct contributions:

- fixed effects, including measured covariates and temporal effects;
- a persistent spatial field, shared through time; and
- a spatiotemporal field, representing time-specific spatial departures.

Their sum describes the fitted linear predictor, but their influence
interpretations differ. A persistent spatial-field influence can reveal a
change in where observations were collected. A spatiotemporal-field influence
can also contain a genuine fitted annual spatial departure. `influ2` therefore
keeps the fields separate instead of silently combining them.

The examples below follow the packages' own worked examples: the
[`sdmTMB` introduction](https://sdmtmb.github.io/sdmTMB/articles/basic-intro.html)
and the [`tinyVAST` vector-autoregressive vignette](https://vast-lib.github.io/tinyVAST/articles/web_only/VAST.html).
They are deliberately small enough to rebuild as package documentation.

## sdmTMB

### Data and mesh

The `sdmTMB` example uses its Pacific cod survey data from Queen Charlotte
Sound [@Anderson2025]. The response is whether Pacific cod were encountered in
a tow. Depth is included as a quadratic effect, and year is the diagnostic
focus.

```{r sdmtmb-data, eval=has_sdmTMB, fig.cap="Observed Pacific cod survey tows, coloured by catch density."}
library(influ2)
data("pcod", package = "sdmTMB")
data("qcs_grid", package = "sdmTMB")

ggplot(pcod, aes(X, Y, colour = density)) +
  geom_point(size = 1.2) +
  coord_fixed() +
  scale_colour_viridis_c(trans = "sqrt") +
  labs(x = "Easting (km)", y = "Northing (km)", colour = "Density") +
  theme_bw()
```

The triangulated mesh approximates a continuous Matérn spatial field. The
coarse cutoff is appropriate for a documentation example, rather than a
recommendation for production analysis.

```{r sdmtmb-mesh, eval=has_sdmTMB, fig.cap="Triangulated mesh used for the Pacific cod spatiotemporal model."}
pcod_mesh <- sdmTMB::make_mesh(pcod, c("X", "Y"), cutoff = 10)
plot(pcod_mesh)
```

### Fit and diagnose

This is the package website's binomial spatiotemporal model, with a persistent
spatial field and independent year-specific spatiotemporal fields.

```{r sdmtmb-fit, eval=has_sdmTMB}
pcod_model <- sdmTMB::sdmTMB(
  present ~ depth_scaled + depth_scaled2,
  data = pcod,
  mesh = pcod_mesh,
  family = binomial(link = "logit"),
  spatial = "on",
  time = "year",
  spatiotemporal = "IID",
  silent = TRUE
)

pcod_diagnostic <- influ(pcod_model, focus = "year")
summary(pcod_diagnostic)
```

The common components plot shows the influence associated with changing depth
coverage, persistent spatial structure, and year-specific spatial departures.

```{r sdmtmb-components, eval=has_sdmTMB, fig.height=8, fig.cap="Fixed, spatial, and spatiotemporal influence components from the sdmTMB Pacific cod model."}
plot(
  pcod_diagnostic,
  type = "components",
  term = c(
    "depth_scaled", "depth_scaled2", "spatial_field",
    "spatiotemporal_field"
  )
)
```

### Fitted field surfaces

`sdmTMB::predict()` exposes the persistent field as `omega_s`, the
spatiotemporal field as `epsilon_st`, and their combined contribution as
`est_rf`. Mapping the same components that enter the influence diagnostic is an
important interpretation check.

```{r sdmtmb-fields, eval=has_sdmTMB, fig.width=10, fig.height=8, fig.cap="Persistent spatial field, year-specific spatiotemporal field, and their sum for two Pacific cod survey years."}
mapping_grid <- do.call(
  rbind,
  lapply(c(2003L, 2017L), function(selected_year) {
    transform(qcs_grid, year = selected_year)
  })
)
field_predictions <- predict(pcod_model, newdata = mapping_grid)
field_surfaces <- tidyr::pivot_longer(
  field_predictions,
  cols = c(omega_s, epsilon_st, est_rf),
  names_to = "component",
  values_to = "link_contribution"
)
field_surfaces$component <- factor(
  field_surfaces$component,
  levels = c("omega_s", "epsilon_st", "est_rf"),
  labels = c("Persistent spatial", "Spatiotemporal", "Combined fields")
)

ggplot(field_surfaces, aes(X, Y, fill = link_contribution)) +
  geom_raster() +
  coord_fixed() +
  facet_grid(component ~ year) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0
  ) +
  labs(
    x = "Easting (km)", y = "Northing (km)",
    fill = "Link-scale\ncontribution"
  ) +
  theme_bw()
```

## tinyVAST

### Simulated spatiotemporal process

The `tinyVAST` example adapts the package's official univariate
vector-autoregressive simulation [@Thorson2025]. It is smaller, uses a Poisson
response, and adds changing spatial coverage so that an influence diagnostic
has a sampling pattern to detect. The known process contains a persistent
spatial field and an AR(1) spatiotemporal field.

```{r tinyvast-simulation, eval=has_tinyVAST}
set.seed(101)
n_x <- n_y <- 7
n_time <- 6
rho <- 0.45
spatial_sd <- 0.70
spatiotemporal_sd <- 0.50

spatial_correlation <- exp(
  -0.5 * abs(outer(seq_len(n_x), seq_len(n_y), FUN = "-"))
)
spatial_covariance <- kronecker(spatial_correlation, spatial_correlation)
spatial_chol <- t(chol(
  spatial_sd^2 * spatial_covariance + diag(1e-8, n_x * n_y)
))
spatiotemporal_chol <- t(chol(
  spatiotemporal_sd^2 * spatial_covariance + diag(1e-8, n_x * n_y)
))

spatiotemporal_field <- t(replicate(
  n_time,
  as.numeric(spatiotemporal_chol %*% rnorm(n_x * n_y))
))
for (time_index in 2:n_time) {
  spatiotemporal_field[time_index, ] <-
    rho * spatiotemporal_field[time_index - 1, ] +
    sqrt(1 - rho^2) * spatiotemporal_field[time_index, ]
}
spatial_field <- as.numeric(spatial_chol %*% rnorm(n_x * n_y))

linear_predictor <-
  1 +
  outer(seq(-0.15, 0.15, length.out = n_time), rep(1, n_x * n_y)) +
  outer(rep(1, n_time), spatial_field) +
  spatiotemporal_field

tiny_grid <- data.frame(
  expand.grid(
    time = seq_len(n_time),
    x = seq_len(n_x),
    ycoord = seq_len(n_y)
  ),
  true_linear_predictor = as.vector(linear_predictor)
)
tiny_grid$mean <- exp(tiny_grid$true_linear_predictor)
tiny_grid$count <- rpois(nrow(tiny_grid), tiny_grid$mean)
tiny_grid$var <- "density"
tiny_grid$dist <- "poisson"

sampling_centre <- 1 +
  (tiny_grid$time - 1) * (n_x - 1) / (n_time - 1)
sampling_probability <- 0.45 + 0.50 * exp(
  -(tiny_grid$x - sampling_centre)^2 / (2 * 2^2)
)
tiny_data <- tiny_grid[runif(nrow(tiny_grid)) < sampling_probability, ]
rownames(tiny_data) <- NULL
```

The observation window shifts from left to right through time. That imbalance
allows the persistent field to affect temporal summaries even though the field
itself does not change through time.

```{r tinyvast-sampling, eval=has_tinyVAST, fig.width=9, fig.height=6, fig.cap="Changing spatial coverage in the simulated tinyVAST example."}
ggplot(tiny_data, aes(x, ycoord, colour = count, size = count + 1)) +
  geom_point(alpha = 0.8) +
  coord_equal() +
  facet_wrap(~time, nrow = 2) +
  scale_colour_viridis_c() +
  scale_size_area(max_size = 5) +
  labs(x = "X", y = "Y", colour = "Count", size = "Count + 1") +
  theme_bw()
```

### Fit and diagnose

The structural-equation strings match the official `tinyVAST` notation: one
persistent spatial variance, plus an AR(1) spatiotemporal process.

```{r tinyvast-fit, eval=has_tinyVAST}
tiny_mesh <- fmesher::fm_mesh_2d(
  tiny_data[c("x", "ycoord")],
  cutoff = 1
)
space_term <- "
  density <-> density, spatial_sd
"
spacetime_term <- "
  density -> density, 1, rho
  density <-> density, 0, spatiotemporal_sd
"

tiny_model <- tinyVAST::tinyVAST(
  count ~ factor(time),
  data = tiny_data,
  family = list(poisson = poisson()),
  spatial_domain = tiny_mesh,
  space_term = space_term,
  spacetime_term = spacetime_term,
  space_columns = c("x", "ycoord")
)

tiny_diagnostic <- influ(tiny_model, focus = "time")
summary(tiny_diagnostic)
```

```{r tinyvast-components, eval=has_tinyVAST, fig.height=7, fig.cap="Fixed, spatial, and spatiotemporal influence components from the tinyVAST model."}
plot(tiny_diagnostic, type = "components")
```

### Fitted field surfaces

`tinyVAST::project()` exposes the same latent components used by the adapter.
The combined map below is calculated on the link scale before applying the
inverse link.

```{r tinyvast-fields, eval=has_tinyVAST, fig.width=10, fig.height=8, fig.cap="Fitted persistent spatial field, spatiotemporal field, and their sum from the tinyVAST example."}
tiny_mapping_grid <- subset(tiny_grid, time %in% c(1, 3, 6))
tiny_mapping_grid$spatial <- tinyVAST::project(
  tiny_model,
  extra_times = numeric(0),
  newdata = tiny_mapping_grid,
  what = "pomega1_g",
  future_var = FALSE,
  past_var = FALSE,
  parm_var = FALSE
)
tiny_mapping_grid$spatiotemporal <- tinyVAST::project(
  tiny_model,
  extra_times = numeric(0),
  newdata = tiny_mapping_grid,
  what = "pepsilon1_g",
  future_var = FALSE,
  past_var = FALSE,
  parm_var = FALSE
)
tiny_mapping_grid$combined <-
  tiny_mapping_grid$spatial + tiny_mapping_grid$spatiotemporal

tiny_surfaces <- tidyr::pivot_longer(
  tiny_mapping_grid,
  cols = c(spatial, spatiotemporal, combined),
  names_to = "component",
  values_to = "link_contribution"
)
tiny_surfaces$component <- factor(
  tiny_surfaces$component,
  levels = c("spatial", "spatiotemporal", "combined"),
  labels = c("Persistent spatial", "Spatiotemporal", "Combined fields")
)

ggplot(tiny_surfaces, aes(x, ycoord, fill = link_contribution)) +
  geom_raster() +
  coord_equal() +
  facet_grid(component ~ time) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0
  ) +
  labs(x = "X", y = "Y", fill = "Link-scale\ncontribution") +
  theme_bw()
```

## Current interpretation boundary

The fixed-effect bands use each model's joint maximum-likelihood covariance.
The latent spatial and spatiotemporal results currently use fitted conditional
modes. They are therefore useful for component attribution and exploratory
diagnosis, but they do not yet carry uncertainty from the sparse joint
precision matrix. Uncertainty-aware field influence, and the combination of
field uncertainty across delta-model components, remain explicit development
targets rather than being approximated with independent marginal bands.

## References
